### 2021-07-02 analise/customerDropoutMembership.Rmd
upd

---
title: |
  |
  |
  | \vspace{1cm}Customer dropout membership*^[*Corresponding address: sobreiro@esdrm.ipsantarem.pt. The current template adapts part of the Rmd code by [Paul C. Bauer](https://github.com/paulcbauer/Writing_a_reproducable_paper_in_rmarkdown), Mannheim Centre for European Social Research.]\vspace{0.5cm}
  |
author: |
  | Pedro Sobreiro, Javier Berrocal, Domingos Martinho, José Garcia Alonso
  |
  | Extremadura University
  |
date: |
  |
  |
  | `r gsub("^0", "", format(Sys.time(), "%d %B, %Y"))`
  |
  |
linestretch: 1.2
colorlinks: true
abstract: \noindent\setstretch{1}Abstract of the article.\vspace{.8cm}
bibliography: references.bib
csl: american-sociological-association.csl
output:
  bookdown::pdf_document2:
    includes:
    toc: no
    keep_tex: true
    latex_engine: xelatex
mainfont: Times New Roman
sansfont: Times New Roman
fontsize: 12pt
link-citations: true
documentclass: article
geometry: margin=1in
always_allow_html: yes
header-includes:
   - \usepackage{dcolumn}
   - \usepackage{color}
   - \usepackage{pdfpages}
   - \usepackage{amsmath}
---

```{r Setup, include=FALSE}
   knitr::opts_chunk$set(cache = FALSE)
   # Use chache = TRUE if you want to speed up compilation
   # set path
   # get rmarkdown directory
   caminho<-getwd()
   # set working directory
   setwd(caminho)
   print(caminho)

   # A function to allow for showing some of the inline code
   rinline <- function(code){
     html <- '<code  class="r">``` `r CODE` ```</code>'
     sub("CODE", code, html)

     ##https://opensource.com/article/19/5/python-3-default-mac

     ##Sys.setenv(RETICULATE_PYTHON = "C:/Users//AppData/Local/Microsoft/WindowsApps/python3")

   ##install.packages("reticulate")
   library(reticulate)
   ##use_python("/Library/Frameworks/Python.framework/Versions/3.8/bin/python3")

   ##use_virtualenv("C:/Users/mangelo.EEG/Documents/python")

   ##knitr::opts_chunk$set(python.reticulate=FALSE)

   # library(devtools) # before this you may need to install devtools
   # install_github("hemken/Statamarkdown")

   }
```

```{r RETICULATE,echo=FALSE}
#Sys.setenv(RETICULATE_PYTHON = "C:/Program Files/Python39/python.exe")
#library(reticulate)
#use_virtualenv("C:/Users/mangelo.EEG/Documents/python")
```

\clearpage

\renewcommand{\baselinestretch}{0.5}\normalsize
[//]: # \tableofcontents
\renewcommand{\baselinestretch}{1.1}\normalsize

\clearpage

```{r Stats, echo=TRUE, message=FALSE}
library(dplyr)
library(dlookr)
library(ggplot2)

##eda_report(nlswork,output_dir = "C:/Users/mangelo.EEG/Documents/GitHub/prjs/reports/",output_file = "eda_report.pdf")

## The data

names(airquality)
##summary(nlswork)

## Missing values

library(visdat)
vis_dat(airquality)

library(naniar)
vis_miss(airquality)

gg_miss_upset(airquality)
## GRAPHS
dplyr::glimpse(cars$Ozone)
d <- density(airquality$Temp)
plot(d)

plot(airquality$Ozone, airquality$Solar.R)
ggplot(airquality, aes(x = Ozone, y = Solar.R)) +
geom_miss_point()

ggplot(airquality, aes(x = Wind, y = Temp)) +
geom_miss_point() +
facet_wrap(vars(Month))

stats <- summary(airquality$Temp)
stats

describe(airquality)

```

The average age in our data is `r round(mean(airquality$Temp,na.rm=TRUE),1)`.

# Tables {#sec:tables}
R Markdown PDF is now able to produce good tables with our output. For `stargazer` the label is contained in the function, while for `kable` it's contained in the chunk name.

## stargazer(): Summary and regression tables

Table \@ref(tab1) shows data's summary statistics.^[You can reference the table as \@ref(tab1cars).] `stargazer()` is and excellent solution to export outputs.

```{r Export Stats, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
library(stargazer)
stargazer(cars,
          title = "Summary table with stargazer",
          label="tab1cars",
          table.placement = "h",
          header=FALSE)
```

Table \@ref(tab2) reports regression outputs. Name the models as you can refer to their names in the text (M1, M2, M3).

```{r Export Regressions, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
library(stargazer)
model1 <- lm(speed ~ dist, data = cars)
model2 <- lm(speed ~ dist, data = cars)
model3 <- lm(dist ~ speed, data = cars)
stargazer(model1, model2, model3,
          title = "Regression table with stargazer",
          label="tab2",
          table.placement = "h",
          column.labels = c("M1", "M2", "M3"),
          model.numbers = FALSE,
          header=FALSE)
```

# Figures

## Graphs with R
You can insert figures like this. One would like to produce and insert them on the fly in the `.rmd` file. Figure \@ref(fig:fig-1) is such an example.

```{r Figures 1, fig-1, fig.align="center", fig.cap="Scatterplot of Speed and Distance", fig.pos="h", message=FALSE, warning=FALSE, paged.print=FALSE}
plot(cars$speed, cars$dist)
```

However, in some cases it does not work.

## Example: ggplot2 graphs
See the `ggplot2` output reported in Figure \@ref(fig:fig-2).

```{r Fig-2, fig.align="center", fig.cap="Miles per gallon according to the weight", fig.pos="h", fig.width=6, fig.height=3, message=FALSE, warning=FALSE, paged.print=FALSE}
mtcars$cyl <- as.factor(mtcars$cyl) # Convert cyl to factor
library(ggplot2)
ggplot(mtcars, aes(x=wt, y=mpg, shape=cyl)) + geom_point() +
  labs(x="Weight (lb/1000)", y = "Miles/(US) gallon",
       shape="Number of \n Cylinders") + theme_classic()
```

## Another example using Plotly
With `Plotly` we can produce interactive graphs which play well, for example, once can embeded in html webpages (drop by  [here](https://paulcbauer.shinyapps.io/visualizing-causal-scenarios/) for an example). One can insert this type of graphs in R Markdown PDF using `Orca` (it generates static images from Plotly graphs). Go [here](https://github.com/plotly/orca#installation) to check how to install it. See Figure \@ref(fig:fig-3) for an example.

```{r Fig-3, message=FALSE, warning=FALSE}
library(plotly)
p <- plot_ly(cars, type = "scatter", mode="markers",
        x=~speed,
        y=~dist)
#Sys.setenv('MAPBOX_TOKEN' = '12423423') # set arbitrary token
#orca(p, "logs/plotly-plot.pdf")
```

\begin{figure}[ht]
\centering
\caption{Example: export a Plotly figure using `orca`}\label{fig:fig-3}
		\includegraphics[width=0.9\linewidth]{../figures/teste.png}
\begin{flushleft}
\end{flushleft}
\end{figure}
\vspace{-1.2cm}

```{r exemploMedia}
# Lets create a value for example

media <- mean(cars$speed)

```
The criminal rate is `r media`\%o.

\vspace{0.3cm}

# Miguel's tests

## R

Example of an equation

$$\int_0^{2\pi} \sin x~dx$$

*Example of a matrix*

$$
\mathbf{X} = \left[\begin{array}
{rrr}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & 9
\end{array}\right]
$$

or

\begin{equation}
f\left(k\right)=\binom{n}{k}p^k\left(1-p\right)^{n-k} \label{eq:binom}
\end{equation}

See Equation \@ref(eq:binom).

\begin{align}
y_{ijt} = \beta x_{ijt} + \eta_i + \gamma_j + \lambda_t + \varepsilon_{ijt}
\end{align}

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
library(stargazer)
stargazer(cars,
          title = "Summary 24",
          label="tab24",
          table.placement = "ht",
          header=FALSE)
```

# Final remarks

Check the replication package for Bonhomme, Lamadon and Manresa (2019): https://github.com/tlamadon/blm-replicate

\newpage

# References {-}

<div id="refs"></div>

# Appendix: Chunk options {-}

## Software versioning

### R

```{r fig-versioning, echo=TRUE}
cat(paste("#", capture.output(sessionInfo()), "\n", collapse =""))
  # or use message() instead of cat()
```

### 2021-07-02 analise/exemplo.Rmd
upd

---
title: |
  |
  |
  | \vspace{1cm}Literate programming with Python, R, Julia and Stata*^[*Corresponding address: miguel.portela@eeg.uminho.pt. The current template adapts part of the Rmd code by [Paul C. Bauer](https://github.com/paulcbauer/Writing_a_reproducable_paper_in_rmarkdown), Mannheim Centre for European Social Research.]\vspace{0.5cm}
  |
author: |
  | Pedro Sobreiro, Javier Berrocal, Domingos Martinho, José Garcia Alonso
  |
  | Extremadura University
  |
date: |
  |
  |
  | `r gsub("^0", "", format(Sys.time(), "%d %B, %Y"))`
  |
  |
linestretch: 1.2
colorlinks: true
abstract: \noindent\setstretch{1}Abstract of the article.\vspace{.8cm}
bibliography: references.bib
csl: american-sociological-association.csl
output:
  bookdown::pdf_document2:
    includes:
    toc: no
    keep_tex: true
    latex_engine: xelatex
mainfont: Times New Roman
sansfont: Times New Roman
fontsize: 12pt
link-citations: true
documentclass: article
geometry: margin=1in
always_allow_html: yes
header-includes:
   - \usepackage{dcolumn}
   - \usepackage{color}
   - \usepackage{pdfpages}
   - \usepackage{amsmath}
---

```{r Setup, include=FALSE}
   knitr::opts_chunk$set(cache = FALSE)
   # Use chache = TRUE if you want to speed up compilation

   # A function to allow for showing some of the inline code
   rinline <- function(code){
     html <- '<code  class="r">``` `r CODE` ```</code>'
     sub("CODE", code, html)

     ##https://opensource.com/article/19/5/python-3-default-mac

     ##Sys.setenv(RETICULATE_PYTHON = "C:/Users//AppData/Local/Microsoft/WindowsApps/python3")

   ##install.packages("reticulate")
   library(reticulate)
   ##use_python("/Library/Frameworks/Python.framework/Versions/3.8/bin/python3")

   ##use_virtualenv("C:/Users/mangelo.EEG/Documents/python")

   ##knitr::opts_chunk$set(python.reticulate=FALSE)

   # library(devtools) # before this you may need to install devtools
   # install_github("hemken/Statamarkdown")

   }
```

```{r RETICULATE,echo=FALSE}
#Sys.setenv(RETICULATE_PYTHON = "C:/Program Files/Python39/python.exe")
#library(reticulate)
#use_virtualenv("C:/Users/mangelo.EEG/Documents/python")
```

\clearpage

\renewcommand{\baselinestretch}{0.5}\normalsize
[//]: # \tableofcontents
\renewcommand{\baselinestretch}{1.1}\normalsize

\clearpage

```{r Stats, echo=TRUE, message=FALSE}
library(dplyr)
library(dlookr)
library(ggplot2)

##eda_report(nlswork,output_dir = "C:/Users/mangelo.EEG/Documents/GitHub/prjs/reports/",output_file = "eda_report.pdf")

## The data

names(airquality)
##summary(nlswork)

## Missing values

library(visdat)
vis_dat(airquality)

library(naniar)
vis_miss(airquality)

gg_miss_upset(airquality)
## GRAPHS
dplyr::glimpse(cars$Ozone)
d <- density(airquality$Temp)
plot(d)

plot(airquality$Ozone, airquality$Solar.R)
ggplot(airquality, aes(x = Ozone, y = Solar.R)) +
geom_miss_point()

ggplot(airquality, aes(x = Wind, y = Temp)) +
geom_miss_point() +
facet_wrap(vars(Month))

stats <- summary(airquality$Temp)
stats

describe(airquality)

```

The average age in our data is `r round(mean(nlswork$age,na.rm=TRUE),1)`.

# Tables {#sec:tables}
R Markdown PDF is now able to produce good tables with our output. For `stargazer` the label is contained in the function, while for `kable` it's contained in the chunk name.

## stargazer(): Summary and regression tables

Table \@ref(tab1) shows data's summary statistics.^[You can reference the table as \@ref(tab1cars).] `stargazer()` is and excellent solution to export outputs.

```{r Export Stats, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
library(stargazer)
stargazer(cars,
          title = "Summary table with stargazer",
          label="tab1cars",
          table.placement = "H",
          header=FALSE)
```

Table \@ref(tab2) reports regression outputs. Name the models as you can refer to their names in the text (M1, M2, M3).

```{r Export Regressions, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
library(stargazer)
model1 <- lm(speed ~ dist, data = cars)
model2 <- lm(speed ~ dist, data = cars)
model3 <- lm(dist ~ speed, data = cars)
stargazer(model1, model2, model3,
          title = "Regression table with stargazer",
          label="tab2",
          table.placement = "H",
          column.labels = c("M1", "M2", "M3"),
          model.numbers = FALSE,
          header=FALSE)
```

# Figures

## Graphs with R
You can insert figures like this. One would like to produce and insert them on the fly in the `.rmd` file. Figure \@ref(fig:fig-1) is such an example.

```{r Figures 1, fig-1, fig.align="center", fig.cap="Scatterplot of Speed and Distance", fig.pos="H", message=FALSE, warning=FALSE, paged.print=FALSE}
plot(cars$speed, cars$dist)
```

However, in some cases it does not work.

## Example: ggplot2 graphs
See the `ggplot2` output reported in Figure \@ref(fig:fig-2).

```{r Fig-2, fig.align="center", fig.cap="Miles per gallon according to the weight", fig.pos="H", fig.width=6, fig.height=3, message=FALSE, warning=FALSE, paged.print=FALSE}
mtcars$cyl <- as.factor(mtcars$cyl) # Convert cyl to factor
library(ggplot2)
ggplot(mtcars, aes(x=wt, y=mpg, shape=cyl)) + geom_point() +
  labs(x="Weight (lb/1000)", y = "Miles/(US) gallon",
       shape="Number of \n Cylinders") + theme_classic()
```

## Another example using Plotly
With `Plotly` we can produce interactive graphs which play well, for example, once can embeded in html webpages (drop by  [here](https://paulcbauer.shinyapps.io/visualizing-causal-scenarios/) for an example). One can insert this type of graphs in R Markdown PDF using `Orca` (it generates static images from Plotly graphs). Go [here](https://github.com/plotly/orca#installation) to check how to install it. See Figure \@ref(fig:fig-3) for an example.

```{r Fig-3, message=FALSE, warning=FALSE}
library(plotly)
p <- plot_ly(cars, type = "scatter", mode="markers",
        x=~speed,
        y=~dist)
#Sys.setenv('MAPBOX_TOKEN' = '12423423') # set arbitrary token
#orca(p, "logs/plotly-plot.pdf")
```

\begin{figure}[ht]
\centering
\caption{Example: export a Plotly figure using `orca`}\label{fig:fig-3}
		\includegraphics[width=0.9\linewidth]{logs/plotly-plot.pdf}
\begin{flushleft}
\end{flushleft}
\end{figure}
\vspace{-1.2cm}

# Python

## API data download using Python

```{python Py,eval=TRUE,echo=TRUE}
import sys
print(sys.version)

import json
##from json.decoder import JSONDecodeError
import requests
import numpy as np
import pandas as pd

## INE: https://www.ine.pt/ine/json_indicador/pindica.jsp?
## op=2&varcd=0008074&Dim1=S7A2015&Dim2=200&Dim3=3&lang=PT

# api-endpoint

URL = "https://www.ine.pt/ine/json_indicador/pindica.jsp"

# define parameters

OP="2"
VARCD="0008074"
DIM1="S7A2015"
DIM2="200"
DIM3="3"
LANG="PT"

# defining a params dict for the parameters to be sent to the API
PARAMS = {'op':OP,'varcd':VARCD,'Dim1':DIM1,'Dim2':DIM2,'Dim3':DIM3,'lang':LANG}

# sending get request and saving the response as response object
r = requests.get(url = URL,params=PARAMS)

# extracting data in json format
data = r.json()

valor = data[0]['Dados']['2015'][0]['valor']

valor

```

The criminal rate is `r py$valor`\%o.

\vspace{0.3cm}

## Import data from PDF files

```{bash Bash,eval=FALSE}
  cd C:/Users/mangelo.EEG/Documents/GitHub/prjs/pdfs
	find . -name '*.pdf' -print0 | xargs -0 -n1 pdfsandwich -gray
	find . -name '*ocr.pdf' -print0 | xargs -0 -n1 pdftotext
```

```{python Python 2, eval=TRUE,echo=FALSE}
import os
import numpy as np
import pandas as pd
import re

## CHECK PyPDF2

## wget -A pdf -m -p -E -k -K -np https://joram.madeira.gov.pt/joram/4serie/
## find . -name '*.pdf' -print0 | xargs -0 -n1 pdfsandwich -gray
## find . -name '*ocr.pdf' -print0 | xargs -0 -n1 pdftotext

# Create list with .txt files for the specified folder
files_list = list()
for (dirpath, dirnames, filenames) in os.walk('C:/Users/mangelo.EEG/Documents/GitHub/prjs/pdfs/'):
    files_list += [os.path.join(dirpath, file)
                   for file in filenames if file.endswith('.txt')]

##print("START:FILES -- list")

##print(files_list)

##print("END:FILES -- list")

p1 = r'PORTARIA'
p2 = r'EXTENSAO'
p3 = r'Materiais'
p5 = r'PE das'

linha = []
output = []
other = []
palavra = []
source = []

for file in files_list:

    f = open(file, "r", encoding='latin8')
    data = f.read()
    f.close()

    line = []
    nh = 0

    tmp1 = str(data)
    #print(tmp1)
    tmp2 = tmp1.splitlines()
    #print(tmp2)
    for n,tmp3 in enumerate(tmp2):
        #print(tmp3)
        if (tmp3.find("PE das") == 0):
            tmp4 = tmp3 + tmp2[2]
            line.append(tmp4)
            #print(n)
            nh = 1
        elif (nh == 1):
            nh = 0
            continue
        elif (nh == 0):
            line.append(tmp3)

    print(line)

    print("   ")

    print("FILE: ", file[46:-4])

    for num, word in enumerate(line):
            if num == 0:
                continue
            else:
                match1 = re.search(p1, word)
                match2 = re.search(p2, word)
                match3 = re.search(p3, word)
                match4 = re.search(r'\d{9}', word)
                match5 = re.search(p5, word)
                ##print("   ")
                ##print("START: ",num)

                if match1:
                        ##print("   ")
                        print("match 1")
                        if match4:
                            ##print("   ")
                            print("match 4")
                            linha.append(num)
                            output.append(re.search(r'\d{9}', word).group())
                            other.append("vazio")
                            palavra.append(p1)
                            source.append(file[46:-4])
                elif match2:
                            ##print("   ")
                            print("match 2")
                            linha.append(num)
                            output.append(re.search(r'\d{9}', word).group())
                            other.append("vazio")
                            palavra.append(p2)
                            source.append(file[46:-4])
                elif match3:
                            ##print("   ")
                            print("match 3")
                            linha.append(num)
                            output.append(re.search(r'\d{9}', word).group())
                            other.append("vazio")
                            palavra.append(p3)
                            source.append(file[46:-4])
                elif match5:
                            ##print("   ")
                            print("-> match 5")
                            ##word.sub(" e o ", " e a ",1)
                            print(word)
                            linha.append(num)

                            if (word.find(" e o ") > 0):
                                print("11111")
                                output.append((word.split("re a", 1)[1]).split(" e o ", -1)[-2])
                                other.append((word.split("re a", 1)[1]).split(" e o ", 1)[1])
                            elif (word.find(" e a ") > 0):
                                print("99999")
                                output.append((word.split("re a", 1)[1]).split(" e a ", -1)[-2])
                                other.append((word.split("re a", 1)[1]).split(" e a ", 1)[1])

                            palavra.append(p5)
                            source.append(file[46:-4])
## o paragrafo tem de estar na mesma linha e temos de ter 'e a' em vez de 'e o'
df = pd.DataFrame({'linha': linha, 'output': output,
                   'outra': other, 'source': source})
print(df)

df.to_csv('data/PE.csv', index=False)
df.to_stata('data/PE.dta', write_index = False)

```

And now we use Stata to explore the data.

```{stata Stata Import CSV, echo=TRUE,message=FALSE}

quiet cd "C:/Users/mangelo.EEG/Documents/GitHub/prjs/logs"
quiet import delimited "C:/Users/mangelo.EEG/Documents/GitHub/prjs/data/PE.csv", encoding(ISO-8859-2) clear
tab source

```

## Run a Python script to read a PDF using the shell to run the commands wget, pdfsandwich and pdftotext

```{bash Bash 2}
python3 "C:/Users/mangelo.EEG/Documents/GitHub/prjs/chunks/python_chunk.py"
```

## Use Stata to list the fiscal numbers (read the data produced with Python)

```{stata Codebook, echo=TRUE,message=FALSE}

quietly{
cd C:/Users/mangelo.EEG/Documents/GitHub/prjs/chunks

use C:/Users/mangelo.EEG/Documents/GitHub/prjs/data/nipcs, clear
compress
contract nipc
drop _freq
drop if nipc == .
format %12.0f nipc
}

//codebook nipc

tab nipc
```

# Julia experiments

## Computations

```{julia Julia experiments, echo=FALSE}
## This is a julia language chunk.
## In julia, the command without ending semicolon will trigger the display
## so is JuliaCall package.
## The julia display will follow immediately after the corresponding command
## just as the R code in R Markdown.

using ReadStat
using StatFiles
using StatsBase
using DataFrames
using FixedEffectModels

@time results_hdfe1 = reg(DataFrame(load("C:/Users/mangelo.EEG/Documents/GitHub/prjs/data/data_short.dta")), @formula(lnrealwage ~ education + lnsales + fe(workerid) + fe(year)));

@time results_hdfe2 = reg(DataFrame(load("C:/Users/mangelo.EEG/Documents/GitHub/prjs/data/data_short.dta")), @formula(lnrealwage ~ education + lnsales + fe(workerid) + fe(firmid) + fe(year)));

using RegressionTables
regtable(results_hdfe1,results_hdfe2; renderSettings = latexOutput("logs/hdfe_output.tex"))

VERSION
```

## Grab results in R

```{r Grab results in R, echo=FALSE}
library(JuliaCall)

  julia_eval("results_hdfe2")

betas <- julia_eval("coef(results_hdfe2)")
r2 <- julia_eval("r2(results_hdfe2)")
```

## Insert Julia's results

\vspace{0.3cm}

The estimated return to education is `r round(betas[1:1]*100,1)`\%. The model has an $R^2$ of `r round(r2,4)`.

\vspace{0.3cm}

## Computation time: run the HDFE with Stata and R and compare to Julia

```{stata HDFE with Stata, echo=TRUE,message=FALSE}

use C:/Users/mangelo.EEG/Documents/GitHub/prjs/data/data_short, clear

timer on 1

	reghdfe lnrealwage education lnsales,absorb(workerid firmid year)

timer off 1
timer list 1
timer clear 1
```

```{r,eval=FALSE,echo=TRUE}
library(lfe)
data_short <- read_dta("C:/Users/mangelo.EEG/Documents/GitHub/prjs/data/data_short.dta")

system.time(est_hdfe <- felm(data_short$lnrealwage ~ data_short$education + data_short$lnsales | data_short$workerid + data_short$firmid + data_short$year))

summary(est_hdfe)
```

## Output Julia's table for HDFE

\begin{table}[ht]
\label{tab:hdfe}
  \input{logs/hdfe_output}
\end{table}

# Miguel's tests

## R

Table \@ref(tab3) ... See Section \@ref(sec:stata)

Example of an equation

$$\int_0^{2\pi} \sin x~dx$$

*Example of a matrix*

$$
\mathbf{X} = \left[\begin{array}
{rrr}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & 9
\end{array}\right]
$$

or

\begin{equation}
f\left(k\right)=\binom{n}{k}p^k\left(1-p\right)^{n-k} \label{eq:binom}
\end{equation}

See Equation \@ref(eq:binom).

\begin{align}
y_{ijt} = \beta x_{ijt} + \eta_i + \gamma_j + \lambda_t + \varepsilon_{ijt}
\end{align}

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
library(stargazer)
library(Statamarkdown)
stataexe <- "C:/Program Files/Stata16/StataMP-64.exe"
knitr::opts_chunk$set(engine.path=list(stata=stataexe))

setwd("C:/Users/mangelo.EEG/Documents/GitHub/prjs/logs")
rm(list = ls())
library(haven)
nlswork <- read_dta("../data/nlswork.dta")

auto <- read_dta("../data/auto.dta")

attach(nlswork)

regs1 <- lm(auto$price ~ auto$mpg + auto$weight)
regs2 <- lm(auto$price ~ auto$mpg + auto$weight + auto$rep78)
regs3 <- lm(auto$price ~ auto$mpg + auto$weight + auto$rep78 + auto$trunk)

regs4 <- lm(ln_wage ~ union)
regs5 <- lm(ln_wage ~ union + collgrad)
regs6 <- lm(ln_wage ~ union + collgrad + age)

##summary(auto)
##summary(regs1)

## https://www.jakeruss.com/cheatsheets/stargazer/

nls<-data.frame(nlswork)

stargazer(nls, summary.stat = c("n", "p75", "sd"), summary.logical = FALSE,
          title = "Summary table",
          label="tab23",
          table.placement = "ht",
          header=FALSE)

stargazer(regs1, regs2, regs3,
          title = "Regression table with stargazer",
          label="tab3",
          table.placement = "ht",
          column.labels = c("M1", "M2", "M3"),
          model.numbers = FALSE,
          header=FALSE,keep=c(0,1,2,3))

attach(auto)

library(naniar)
vis_miss(nlswork)

# plot(y=price,x=mpg)

```

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
library(stargazer)
stargazer(cars,
          title = "Summary 24",
          label="tab24",
          table.placement = "ht",
          header=FALSE)
```

## Grab Stata's output

\vspace{0.3cm}

The mean ln GDP is `r format(as.numeric(readLines('example.txt')), digits = 2)`.

\vspace{0.3cm}

```{r cleanup, include=FALSE} 
unlink("example.txt")
```

See https://www.ssc.wisc.edu/~hemken/Stataworkshops/stata.html#stata-and-r-markdown-the-statamarkdown-package

![Wage density](logs/density.pdf)

\begin{table}[ht]
\caption{Regression analysis}
\label{tab:stata}
  \input{logs/growth_analysis_frag}
\end{table}

## Use Stata to export statistics to Excel

We now export a set of statistics to an Excel file.

See Figure \@ref(fig:fig-tmp).

```{r fig-tmp, echo=FALSE, fig.align="center", fig.cap="Scatterplot test MP", fig.pos="ht", message=FALSE, warning=FALSE, paged.print=FALSE}
plot(x = mpg, y = price,
     pch = 16, frame = FALSE,
     xlab = "wt", ylab = "mpg", col = "#2E9FDF")
```

# Final remarks

Check the replication package for Bonhomme, Lamadon and Manresa (2019): https://github.com/tlamadon/blm-replicate

\newpage

# References {-}

<div id="refs"></div>

# Appendix: Chunk options {-}

## Software versioning

### R

```{r fig-versioning, echo=TRUE}
cat(paste("#", capture.output(sessionInfo()), "\n", collapse =""))
  # or use message() instead of cat()
```

### Python

```{python py2,eval=TRUE,echo=TRUE}
import sys
print(sys.version)
```

### Julia

```{julia echo=FALSE}
VERSION
```

## All the code in the paper

To simply attach all the code you used in the PDF file in the appendix see the R chunk in the underlying `.rmd` file:

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```

<!-- ## Exploratory data analysis report -->

<!-- [//]: # \includepdf[pages=-]{C:/Users/mangelo.EEG/Documents/GitHub/prjs/reports/eda_report.pdf} -->

### 2021-07-02 analysis/customerDropoutMembership.Rmd
upd

--- 2021-07-02 analise/customerDropoutMembership.Rmd
@@
   |
-  | \vspace{1cm}Customer dropout membership*^[*Corresponding address: sobreiro@esdrm.ipsantarem.pt.
+  | \vspace{1cm}Customer dropout membership^[Corresponding address: sobreiro@esdrm.ipsantarem.pt.
 The current template adapts part of the Rmd code by [Paul C.
@@
 ```
+
+# Introduction
+
+Research idea:
+
+-
+
+Context:
+An organization membership located in portugal. The organization offers an annual membership
+for the members, the service subscription has several payment options:
+
+- Men with a annual fee of 10€
+- Women annual fee of 6€
+- Correspondent fee 6€
+- Retired fee 5€
+- Student fee 2.5€
+- under-14 fee 1€
 

### 2021-07-06 article/customerDropoutMembership.Rmd
upd

--- 2021-07-02 analysis/customerDropoutMembership.Rmd
@@
 documentclass: article
-geometry: margin=1in
+geometry: margin = 1in
 always_allow_html: yes
@@
 
-```{r Setup, include=FALSE}
+```{r Setup, include = FALSE}
    knitr::opts_chunk$set(cache = FALSE)
-   # Use chache = TRUE if you want to speed up compilation
+   # Use cache = TRUE if you want to speed up compilation
    # set path
    # get rmarkdown directory
-   caminho<-getwd()
+   caminho <- getwd()
    # set working directory
@@
    # A function to allow for showing some of the inline code
-   rinline <- function(code){
-     html <- '<code  class="r">``` `r CODE` ```</code>'
+   rinline <- function(code) {
+     html <- '<code  class = "r">``` `r CODE` ```</code>'
      sub("CODE", code, html)
@@
 
-   ##knitr::opts_chunk$set(python.reticulate=FALSE)
+   ##knitr::opts_chunk$set(python.reticulate = FALSE)
 
@@
 
-```{r RETICULATE,echo=FALSE}
+```{r RETICULATE,echo = FALSE}
 #Sys.setenv(RETICULATE_PYTHON = "C:/Program Files/Python39/python.exe")
@@
 Context:
-An organization membership located in portugal. The organization offers an annual membership
+An organization membership located in Portugal. The organization offers an annual membership
 for the members, the service subscription has several payment options:
@@
 
-```{r Stats, echo=TRUE, message=FALSE}
+```{r Stats, echo = TRUE, message = FALSE}
 library(dplyr)
@@
 
-##eda_report(nlswork,output_dir = "C:/Users/mangelo.EEG/Documents/GitHub/prjs/reports/",output_file = "eda_report.pdf")
+#eda_report(nlswork,output_dir  = 
+#	"C:/Users/mangelo.EEG/Documents/GitHub/prjs/reports/",
+#	output_file = "eda_report.pdf")
 
@@
 names(airquality)
-##summary(nlswork)
+#summary(nlswork)
 
 ## Missing values
-
 library(visdat)
@@
 
-The average age in our data is `r round(mean(airquality$Temp,na.rm=TRUE),1)`.
+The average age in our data is `r round(mean(airquality$Temp,na.rm = TRUE),1)`.
 
@@
 
-```{r Export Stats, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
+```{r Export Stats, echo = TRUE, message = FALSE, warning = FALSE, paged.print = FALSE, results = "asis"}
 library(stargazer)
@@
           title = "Summary table with stargazer",
-          label="tab1cars",
+          label = "tab1cars",
           table.placement = "h",
-          header=FALSE)
+          header = FALSE)
 ```
@@
 
-```{r Export Regressions, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
+```{r Export Regressions, echo = TRUE, message = FALSE, warning = FALSE, paged.print = FALSE, results = "asis"}
 library(stargazer)
@@
           title = "Regression table with stargazer",
-          label="tab2",
+          label = "tab2",
           table.placement = "h",
@@
           model.numbers = FALSE,
-          header=FALSE)
+          header = FALSE)
 ```
@@
 
-```{r Figures 1, fig-1, fig.align="center", fig.cap="Scatterplot of Speed and Distance", fig.pos="h", message=FALSE, warning=FALSE, paged.print=FALSE}
+```{r Figures 1, fig-1, fig.align = "center", fig.cap = "Scatterplot of Speed and Distance", fig.pos = "h", message = FALSE, warning = FALSE, paged.print = FALSE}
 plot(cars$speed, cars$dist)
@@
 
-```{r Fig-2, fig.align="center", fig.cap="Miles per gallon according to the weight", fig.pos="h", fig.width=6, fig.height=3, message=FALSE, warning=FALSE, paged.print=FALSE}
+```{r Fig-2, fig.align = "center", fig.cap = "Miles per gallon according to the weight", fig.pos = "h", fig.width = 6, fig.height = 3, message = FALSE, warning = FALSE, paged.print = FALSE}
 mtcars$cyl <- as.factor(mtcars$cyl) # Convert cyl to factor
 library(ggplot2)
-ggplot(mtcars, aes(x=wt, y=mpg, shape=cyl)) + geom_point() +
-  labs(x="Weight (lb/1000)", y = "Miles/(US) gallon",
-       shape="Number of \n Cylinders") + theme_classic()
+ggplot(mtcars, aes(x = wt, y = mpg, shape = cyl)) + geom_point() +
+  labs(x = "Weight (lb/1000)", y = "Miles/(US) gallon",
+       shape = "Number of \n Cylinders") + theme_classic()
 ```
@@
 ## Another example using Plotly
-With `Plotly` we can produce interactive graphs which play well, for example, once can embeded in html webpages (drop by  [here](https://paulcbauer.shinyapps.io/visualizing-causal-scenarios/) for an example).
-One can insert this type of graphs in R Markdown PDF using `Orca` (it generates static images from Plotly graphs).
+With `Plotly` we can produce interactive graphs which play well, for example, once can embedded 
+in html webpages (drop by  [here](https://paulcbauer.shinyapps.io/visualizing-causal-scenarios/) 
+for an example). One can insert this type of graphs in R Markdown PDF using `Orca` (it generates 
+static images from Plotly graphs). 
 Go [here](https://github.com/plotly/orca#installation) to check how to install it. 
@@
 
-```{r Fig-3, message=FALSE, warning=FALSE}
+```{r Fig-3, message = FALSE, warning = FALSE}
 library(plotly)
-p <- plot_ly(cars, type = "scatter", mode="markers",
-        x=~speed,
-        y=~dist)
+p <- plot_ly(cars, type = "scatter", mode = "markers",
+        x = ~speed,
+        y = ~dist)
 #Sys.setenv('MAPBOX_TOKEN' = '12423423') # set arbitrary token
@@
 \caption{Example: export a Plotly figure using `orca`}\label{fig:fig-3}
-		\includegraphics[width=0.9\linewidth]{../figures/teste.png}
+	\includegraphics[width = 0.9\linewidth]{../figures/teste.png}
 \begin{flushleft}
@@
 \begin{equation}
-f\left(k\right)=\binom{n}{k}p^k\left(1-p\right)^{n-k} \label{eq:binom}
+f\left(k\right) = \binom{n}{k}p^k\left(1-p\right)^{n-k} \label{eq:binom}
 \end{equation}
@@
 
-```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results="asis"}
+```{r echo = TRUE, message = FALSE, warning = FALSE, paged.print = FALSE, results = "asis"}
 library(stargazer)
@@
           title = "Summary 24",
-          label="tab24",
+          label = "tab24",
           table.placement = "ht",
-          header=FALSE)
+          header = FALSE)
 ```
@@
 
-<div id="refs"></div>
+<div id = "refs"></div>
 
@@
 
-```{r fig-versioning, echo=TRUE}
-cat(paste("#", capture.output(sessionInfo()), "\n", collapse =""))
+```{r fig-versioning, echo = TRUE}
+cat(paste("#", capture.output(sessionInfo()), "\n", collapse  = ""))
   # or use message() instead of cat()

### 2021-07-25 article/articleCustomerDropoutMembership.Rmd
upd

--- 2021-07-06 article/customerDropoutMembership.Rmd
@@
 title: |
+  |
   |
@@
 Bauer](https://github.com/paulcbauer/Writing_a_reproducable_paper_in_rmarkdown), Mannheim Centre for European Social Research.]\vspace{0.5cm}
-  |
 author: |
@@
 colorlinks: true
-abstract: \noindent\setstretch{1}Abstract of the article.\vspace{.8cm}
+abstract: \noindent\setstretch{1}Abstract of the article. Here we can place more info.\vspace{.8cm}
 bibliography: references.bib
@@
    - \usepackage{amsmath}
+   - \usepackage{booktabs}
+   - \usepackage{makecell}
+   - \usepackage{hyperref}
+editor_options: 
+  markdown: 
+    wrap: 80
 ---
@@
 ##https://opensource.com/article/19/5/python-3-default-mac
+#Sys.setenv(RETICULATE_PYTHON = "C:/Users//AppData/Local/Microsoft/WindowsApps/python3")
 
-     ##Sys.setenv(RETICULATE_PYTHON = "C:/Users//AppData/Local/Microsoft/WindowsApps/python3")
-
-   ##install.packages("reticulate")
-   library(reticulate)
-   ##use_python("/Library/Frameworks/Python.framework/Versions/3.8/bin/python3")
-
-   ##use_virtualenv("C:/Users/mangelo.EEG/Documents/python")
-
-   ##knitr::opts_chunk$set(python.reticulate = FALSE)
-
-   # library(devtools) # before this you may need to install devtools
-   # install_github("hemken/Statamarkdown")
-
+#install.packages("reticulate")
+#library(reticulate)
+#virtualenv_create("artigosurvival",python_version="3.7.10")
+#use_virtualenv("artigosurvival", required = TRUE)
 }   
@@
 ```{r RETICULATE,echo = FALSE}
-#Sys.setenv(RETICULATE_PYTHON = "C:/Program Files/Python39/python.exe")
-#library(reticulate)
+#set environment first
+Sys.setenv(RETICULATE_PYTHON="C:\\Users\\sobre\\anaconda3\\envs\\survival\\python.exe")
+#call reticulate
+library(reticulate)
+#activate environment
+use_condaenv("survival", required = TRUE)
 #use_virtualenv("C:/Users/mangelo.EEG/Documents/python")
@@
 
-Context:
-An organization membership located in Portugal. The organization offers an annual membership
-for the members, the service subscription has several payment options:
+Context: An organization membership located in Portugal. The organization offers
+an annual membership for the members, the service subscription has several
+payment options:
 
@@
 \renewcommand{\baselinestretch}{0.5}\normalsize
-[//]: # \tableofcontents
+
 \renewcommand{\baselinestretch}{1.1}\normalsize
@@
 
-```{r Stats, echo = TRUE, message = FALSE}
+```{r Stats1, echo = TRUE, message = FALSE}
 library(dplyr)
@@
 
-The average age in our data is `r round(mean(airquality$Temp,na.rm = TRUE),1)`.
+# Experimental Results
+
+## Data description
+
+```{r get_data, echo = FALSE, message = FALSE, warning=FALSE}
+library(stargazer)
+library(readxl)
+library(dplyr)
+library(visdat)
+
+df_members <- read_excel("../data/membershipData.xlsx")
+
+names(df_members)
+# rename column labels
+names(df_members) <- c("num_socio", "dt_inscription", "year", "birth_date",
+                       "age", "sex", "marital_status", "category",
+                       "monthly_fee", "occupation", "zip_code",
+                       "dt_last_invoice", "dt_last_payment", "total_amount",
+                       "total_matches", "season_matches",
+                       "days_since_last_payment", "months_since_last_payment",
+                       "dropout", "years_membership", "stadium_access",
+                       "quart_stadium_entries", "inscription_month")
+
+names(df_members)
+
+# select relevant variables
+df_members <- df_members %>%
+select(year, age, sex, marital_status, monthly_fee, total_amount, total_matches,
+       season_matches, months_since_last_payment, dropout, years_membership,
+       stadium_access, quart_stadium_entries,inscription_month)
+
+str(df_members)
+vis_dat(df_members) #check
+```
+
+Table \@ref(tab:tab1cars) shows data's summary statistics.[^1] `stargazer()` is
+and excellent solution to export outputs.
+
+[^1]: You can reference the table as \@ref(tab1cars).
+
+```{r Export Stats0, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
+library(stargazer)
+df_summary <- summary(df_members, object)
+stargazer(df_members,
+          title = "Summary statistics",
+          label = "tab1cars",
+          table.placement = "h",
+          header = FALSE,
+          summary = TRUE,
+          summary.stat = c("min", "p25", "median", "p75", "max", "median", "sd")
+         )
+
+```
+
+Teste \@ref(tab:summarytable)
+
+```{r summarytable, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
+library(gtsummary)
+library(kableExtra)
+
+tbl <- df_members %>%
+    tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd})",
+                                 all_categorical() ~ "{p}%")
+               )
+as_kable_extra(tbl,booktabs=T,caption="Summary statistics")
+```
+
+```{python Py,eval=TRUE,echo=TRUE}
+from pysurvival.utils.display import correlation_matrix
+import pandas as pd
+import numpy as np
+
+col = ['sex','marital_status','quart_stadium_entries']
+
+df_members = r.df_members #copy r dataframe to python
+
+df_members = pd.get_dummies(df_members, columns=col,drop_first=True)
+
+# Creating the time and event columns
+time_column = 'years_membership'
+event_column = 'dropout'
+
+# Extracting the features
+features = np.setdiff1d(df_members.columns, [time_column, event_column] ).tolist()
+
+correlation_matrix(df_members, figure_size=(10,10), text_fontsize=6)
+```
+
+The average age in our data is `r round(mean(df_members$age,na.rm = TRUE),1)`.
 
 # Tables {#sec:tables}
-R Markdown PDF is now able to produce good tables with our output. For `stargazer` the label is contained in the function, while for `kable` it's contained in the chunk name.
+
+R Markdown PDF is now able to produce good tables with our output. For
+`stargazer` the label is contained in the function, while for `kable` it's
+contained in the chunk name.
 
@@
 
-Table \@ref(tab1) shows data's summary statistics.^[You can reference the table as \@ref(tab1cars).] `stargazer()` is and excellent solution to export outputs.
-
-```{r Export Stats, echo = TRUE, message = FALSE, warning = FALSE, paged.print = FALSE, results = "asis"}
-library(stargazer)
-stargazer(cars,
-          title = "Summary table with stargazer",
-          label = "tab1cars",
-          table.placement = "h",
-          header = FALSE)
-```
-
-Table \@ref(tab2) reports regression outputs. Name the models as you can refer to their names in the text (M1, M2, M3).
+Table \@ref(tab2) reports regression outputs. Name the models as you can refer
+to their names in the text (M1, M2, M3).
 
@@
 ## Graphs with R
-You can insert figures like this. One would like to produce and insert them on the fly in the `.rmd` file. Figure \@ref(fig:fig-1) is such an example.
+
+You can insert figures like this. One would like to produce and insert them on
+the fly in the `.rmd` file. Figure \@ref(fig:fig-1) is such an example.
 
@@
 ## Example: ggplot2 graphs
+
 See the `ggplot2` output reported in Figure \@ref(fig:fig-2).
@@
 ## Another example using Plotly
-With `Plotly` we can produce interactive graphs which play well, for example, once can embedded 
-in html webpages (drop by  [here](https://paulcbauer.shinyapps.io/visualizing-causal-scenarios/) 
-for an example). One can insert this type of graphs in R Markdown PDF using `Orca` (it generates 
-static images from Plotly graphs). 
-Go [here](https://github.com/plotly/orca#installation) to check how to install it. 
+
+With `Plotly` we can produce interactive graphs which play well, for example,
+once can embedded in html webpages (drop by
+[here](https://paulcbauer.shinyapps.io/visualizing-causal-scenarios/) for an
+example). One can insert this type of graphs in R Markdown PDF using `Orca` (it
+generates static images from Plotly graphs). Go
+[here](https://github.com/plotly/orca#installation) to check how to install it.
 See Figure \@ref(fig:fig-3) for an example.
@@
 
+```{=tex}
 \begin{figure}[ht]
@@
 \vspace{-1.2cm}
-
+```
 ```{r exemploMedia}
@@
 ```
-The criminal rate is `r media`\%o.
+
+The criminal rate is `r media`%o.
 
@@
 
+```{=tex}
 \begin{equation}
@@
 \end{equation}
-
+```
 See Equation \@ref(eq:binom).
 
+```{=tex}
 \begin{align}
@@
 \end{align}
-
+```
 ```{r echo = TRUE, message = FALSE, warning = FALSE, paged.print = FALSE, results = "asis"}
@@
 
-Check the replication package for Bonhomme, Lamadon and Manresa (2019): https://github.com/tlamadon/blm-replicate
+Check the replication package for Bonhomme, Lamadon and Manresa (2019):
+<https://github.com/tlamadon/blm-replicate>
 
@@
 
-# References {-}
+# References {.unnumbered}
 
-<div id = "refs"></div>
+::: {#refs}
+:::
 
-# Appendix: Chunk options {-}
+# Appendix: Chunk options {.unnumbered}
 

### 2021-07-25 article/articleCustomerDropoutMembership.tex
upd

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}
\title{\hfill\break
\hfill\break
\hfill\break
\vspace{1cm}Customer dropout membership\footnote{Corresponding address: \href{mailto:sobreiro@esdrm.ipsantarem.pt}{\nolinkurl{sobreiro@esdrm.ipsantarem.pt}}. The current template adapts part of the Rmd code by \href{https://github.com/paulcbauer/Writing_a_reproducable_paper_in_rmarkdown}{Paul C. Bauer}, Mannheim Centre for European Social Research.}\vspace{0.5cm}}
\author{Pedro Sobreiro, Javier Berrocal, Domingos Martinho, José Garcia Alonso\\
~\\
Extremadura University\\}
\date{\hfill\break
\hfill\break
25 julho, 2021\\
~\\}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
  \setsansfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin = 1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{dcolumn}
\usepackage{color}
\usepackage{pdfpages}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\begin{document}
\maketitle
\begin{abstract}
\noindent\setstretch{1}Abstract of the article. Here we can place more info.\vspace{.8cm}
\end{abstract}

\setstretch{1.2}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Research idea:

\begin{itemize}
\tightlist
\item
\end{itemize}

Context: An organization membership located in Portugal. The organization offers
an annual membership for the members, the service subscription has several
payment options:

\begin{itemize}
\tightlist
\item
  Men with a annual fee of 10€
\item
  Women annual fee of 6€
\item
  Correspondent fee 6€
\item
  Retired fee 5€
\item
  Student fee 2.5€
\item
  under-14 fee 1€
\end{itemize}

\clearpage

\renewcommand{\baselinestretch}{0.5}\normalsize

\renewcommand{\baselinestretch}{1.1}\normalsize

\clearpage

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(dplyr)}
\FunctionTok{library}\NormalTok{(dlookr)}
\FunctionTok{library}\NormalTok{(ggplot2)}

\CommentTok{\#eda\_report(nlswork,output\_dir  = }
\CommentTok{\#   "C:/Users/mangelo.EEG/Documents/GitHub/prjs/reports/",}
\CommentTok{\#   output\_file = "eda\_report.pdf")}

\DocumentationTok{\#\# The data}

\FunctionTok{names}\NormalTok{(airquality)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "Ozone"   "Solar.R" "Wind"    "Temp"    "Month"   "Day"
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#summary(nlswork)}

\DocumentationTok{\#\# Missing values}
\FunctionTok{library}\NormalTok{(visdat)}
\FunctionTok{vis\_dat}\NormalTok{(airquality)}
\end{Highlighting}
\end{Shaded}

\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(naniar)}
\FunctionTok{vis\_miss}\NormalTok{(airquality)}
\end{Highlighting}
\end{Shaded}

\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-2.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{gg\_miss\_upset}\NormalTok{(airquality)}
\end{Highlighting}
\end{Shaded}

\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-3.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\# GRAPHS}
\NormalTok{dplyr}\SpecialCharTok{::}\FunctionTok{glimpse}\NormalTok{(cars}\SpecialCharTok{$}\NormalTok{Ozone)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  NULL
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{d }\OtherTok{\textless{}{-}} \FunctionTok{density}\NormalTok{(airquality}\SpecialCharTok{$}\NormalTok{Temp)}
\FunctionTok{plot}\NormalTok{(d)}
\end{Highlighting}
\end{Shaded}

\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-4.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot}\NormalTok{(airquality}\SpecialCharTok{$}\NormalTok{Ozone, airquality}\SpecialCharTok{$}\NormalTok{Solar.R)}
\end{Highlighting}
\end{Shaded}

\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-5.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{ggplot}\NormalTok{(airquality, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Ozone, }\AttributeTok{y =}\NormalTok{ Solar.R)) }\SpecialCharTok{+}
\FunctionTok{geom\_miss\_point}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-6.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{ggplot}\NormalTok{(airquality, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Wind, }\AttributeTok{y =}\NormalTok{ Temp)) }\SpecialCharTok{+}
\FunctionTok{geom\_miss\_point}\NormalTok{() }\SpecialCharTok{+}
\FunctionTok{facet\_wrap}\NormalTok{(}\FunctionTok{vars}\NormalTok{(Month))}
\end{Highlighting}
\end{Shaded}

\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-7.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{stats }\OtherTok{\textless{}{-}} \FunctionTok{summary}\NormalTok{(airquality}\SpecialCharTok{$}\NormalTok{Temp)}
\NormalTok{stats}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
##   56.00   72.00   79.00   77.88   85.00   97.00
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{describe}\NormalTok{(airquality)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 6 x 26
##   variable     n    na   mean    sd se_mean   IQR skewness kurtosis   p00   p01
##   <chr>    <int> <int>  <dbl> <dbl>   <dbl> <dbl>    <dbl>    <dbl> <dbl> <dbl>
## 1 Ozone      116    37  42.1  33.0    3.06   45.2  1.24       1.29    1    4.3 
## 2 Solar.R    146     7 186.   90.1    7.45  143   -0.428     -0.968   7   10.2 
## 3 Wind       153     0   9.96  3.52   0.285   4.1  0.348      0.111   1.7  2.56
## 4 Temp       153     0  77.9   9.47   0.765  13   -0.378     -0.404  56   57   
## 5 Month      153     0   6.99  1.42   0.115   2   -0.00239   -1.30    5    5   
## 6 Day        153     0  15.8   8.86   0.717  15    0.00265   -1.20    1    1   
## # ... with 15 more variables: p05 <dbl>, p10 <dbl>, p20 <dbl>, p25 <dbl>,
## #   p30 <dbl>, p40 <dbl>, p50 <dbl>, p60 <dbl>, p70 <dbl>, p75 <dbl>,
## #   p80 <dbl>, p90 <dbl>, p95 <dbl>, p99 <dbl>, p100 <dbl>
\end{verbatim}

\hypertarget{experimental-results}{%
\section{Experimental Results}\label{experimental-results}}

\hypertarget{data-description}{%
\subsection{Data description}\label{data-description}}

\begin{verbatim}
##  [1] "Sócio"               "dataAdesao"          "ano"                
##  [4] "dataNascimento"      "idade"               "sexo"               
##  [7] "estadoCivil"         "categoria"           "quotaMensal"        
## [10] "profissao"           "codPostal"           "ultimaQuota"        
## [13] "ultimoPagamento"     "valorTotal"          "totalJogos"         
## [16] "jogosEpoca"          "diasUltimoPagamento" "mesesUP"            
## [19] "abandonou"           "anosSocio"           "idaEstadio"         
## [22] "escaloesTotalJogos"  "mes"
\end{verbatim}

\begin{verbatim}
##  [1] "num_socio"                 "dt_inscription"           
##  [3] "year"                      "birth_date"               
##  [5] "age"                       "sex"                      
##  [7] "marital_status"            "category"                 
##  [9] "monthly_fee"               "occupation"               
## [11] "zip_code"                  "dt_last_invoice"          
## [13] "dt_last_payment"           "total_amount"             
## [15] "total_matches"             "season_matches"           
## [17] "days_since_last_payment"   "months_since_last_payment"
## [19] "dropout"                   "years_membership"         
## [21] "stadium_access"            "quart_stadium_entries"    
## [23] "inscription_month"
\end{verbatim}

\begin{verbatim}
## tibble [25,316 x 14] (S3: tbl_df/tbl/data.frame)
##  $ year                     : num [1:25316] 1944 1944 1945 1945 1945 ...
##  $ age                      : num [1:25316] 83 88 73 97 97 91 88 95 88 78 ...
##  $ sex                      : chr [1:25316] "M" "M" "M" "M" ...
##  $ marital_status           : chr [1:25316] "casado" "solteiro" "nao definido" "casado" ...
##  $ monthly_fee              : num [1:25316] 10 10 10 5 10 5 5 5 10 10 ...
##  $ total_amount             : num [1:25316] 1906 1906 1553 790 1466 ...
##  $ total_matches            : num [1:25316] 0 0 0 0 0 20 74 0 154 0 ...
##  $ season_matches           : num [1:25316] 0 0 0 0 0 0 0 0 6 0 ...
##  $ months_since_last_payment: num [1:25316] 3 3 36 8 35 4 41 40 4 2 ...
##  $ dropout                  : num [1:25316] 0 0 1 0 1 0 1 1 0 0 ...
##  $ years_membership         : num [1:25316] 74 74 73 73 73 73 73 73 73 72 ...
##  $ stadium_access           : num [1:25316] 0 0 0 0 0 1 1 0 1 0 ...
##  $ quart_stadium_entries    : chr [1:25316] "ate 1" "ate 1" "ate 1" "ate 1" ...
##  $ inscription_month        : num [1:25316] 10 10 8 9 9 12 1 1 2 4 ...
\end{verbatim}

\includegraphics{articleCustomerDropoutMembership_files/figure-latex/get_data-1.pdf}

Table \ref{tab:tab1cars} shows data's summary statistics.\footnote{You can reference the table as \ref{tab1cars}.} \texttt{stargazer()} is
and excellent solution to export outputs.

Teste \ref{tab:summarytable}

\begin{table}

\caption{\label{tab:summarytable}Summary statistics}
\centering
\begin{tabular}[t]{ll}
\toprule
Characteristic & N = 25,316\\
\midrule
year & 2,007 (11)\\
age & 27 (20)\\
sex & \\
\hspace{1em}F & 32\%\\
\hspace{1em}M & 68\%\\
\addlinespace
marital\_status & \\
\hspace{1em}casado & 20\%\\
\hspace{1em}nao definido & 30\%\\
\hspace{1em}outro & 2.0\%\\
\hspace{1em}solteiro & 48\%\\
\addlinespace
monthly\_fee & \\
\hspace{1em}0 & <0.1\%\\
\hspace{1em}1 & 32\%\\
\hspace{1em}2.5 & 28\%\\
\hspace{1em}5 & 3.4\%\\
\addlinespace
\hspace{1em}6 & 12\%\\
\hspace{1em}10 & 24\%\\
total\_amount & 316 (494)\\
total\_matches & 27 (46)\\
season\_matches & 2.2 (4.1)\\
\addlinespace
months\_since\_last\_payment & 19 (32)\\
dropout & 22\%\\
years\_membership & 11 (11)\\
stadium\_access & 40\%\\
quart\_stadium\_entries & \\
\addlinespace
\hspace{1em}1 a 21 & 10\%\\
\hspace{1em}21 a 56 & 9.8\%\\
\hspace{1em}56 a 105 & 10.0\%\\
\hspace{1em}ate 1 & 60\%\\
\hspace{1em}mais 105 & 10.0\%\\
\addlinespace
inscription\_month & 6.9 (3.4)\\
\bottomrule
\multicolumn{2}{l}{\rule{0pt}{1em}\textsuperscript{1} Mean (SD); \%}\\
\end{tabular}
\end{table}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#from pysurvival.utils.display import correlation\_matrix}
\ImportTok{import}\NormalTok{ pandas }\ImportTok{as}\NormalTok{ pd}
\ImportTok{import}\NormalTok{ numpy }\ImportTok{as}\NormalTok{ np}

\NormalTok{col }\OperatorTok{=}\NormalTok{ [}\StringTok{\textquotesingle{}sex\textquotesingle{}}\NormalTok{,}\StringTok{\textquotesingle{}marital\_status\textquotesingle{}}\NormalTok{,}\StringTok{\textquotesingle{}quart\_stadium\_entries\textquotesingle{}}\NormalTok{]}

\NormalTok{df\_members }\OperatorTok{=}\NormalTok{ r.df\_members }\CommentTok{\#copy r dataframe to python}

\NormalTok{df\_members }\OperatorTok{=}\NormalTok{ pd.get\_dummies(df\_members, columns}\OperatorTok{=}\NormalTok{col,drop\_first}\OperatorTok{=}\VariableTok{True}\NormalTok{)}

\CommentTok{\# Creating the time and event columns}
\NormalTok{time\_column }\OperatorTok{=} \StringTok{\textquotesingle{}years\_membership\textquotesingle{}}
\NormalTok{event\_column }\OperatorTok{=} \StringTok{\textquotesingle{}dropout\textquotesingle{}}

\CommentTok{\# Extracting the features}
\NormalTok{features }\OperatorTok{=}\NormalTok{ np.setdiff1d(df\_members.columns, [time\_column, event\_column] ).tolist()}

\CommentTok{\#correlation\_matrix(df\_members, figure\_size=(10,10), text\_fontsize=6)}
\end{Highlighting}
\end{Shaded}

The average age in our data is 27.3.

\hypertarget{sec:tables}{%
\section{Tables}\label{sec:tables}}

R Markdown PDF is now able to produce good tables with our output. For
\texttt{stargazer} the label is contained in the function, while for \texttt{kable} it's
contained in the chunk name.

\hypertarget{stargazer-summary-and-regression-tables}{%
\subsection{stargazer(): Summary and regression tables}\label{stargazer-summary-and-regression-tables}}

Table \ref{tab2} reports regression outputs. Name the models as you can refer
to their names in the text (M1, M2, M3).

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(stargazer)}
\NormalTok{model1 }\OtherTok{\textless{}{-}} \FunctionTok{lm}\NormalTok{(speed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ dist, }\AttributeTok{data =}\NormalTok{ cars)}
\NormalTok{model2 }\OtherTok{\textless{}{-}} \FunctionTok{lm}\NormalTok{(speed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ dist, }\AttributeTok{data =}\NormalTok{ cars)}
\NormalTok{model3 }\OtherTok{\textless{}{-}} \FunctionTok{lm}\NormalTok{(dist }\SpecialCharTok{\textasciitilde{}}\NormalTok{ speed, }\AttributeTok{data =}\NormalTok{ cars)}
\FunctionTok{stargazer}\NormalTok{(model1, model2, model3,}
          \AttributeTok{title =} \StringTok{"Regression table with stargazer"}\NormalTok{,}
          \AttributeTok{label =} \StringTok{"tab2"}\NormalTok{,}
          \AttributeTok{table.placement =} \StringTok{"h"}\NormalTok{,}
          \AttributeTok{column.labels =} \FunctionTok{c}\NormalTok{(}\StringTok{"M1"}\NormalTok{, }\StringTok{"M2"}\NormalTok{, }\StringTok{"M3"}\NormalTok{),}
          \AttributeTok{model.numbers =} \ConstantTok{FALSE}\NormalTok{,}
          \AttributeTok{header =} \ConstantTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}[h] \centering 
  \caption{Regression table with stargazer} 
  \label{tab2} 
\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{3}{c}{\textit{Dependent variable:}} \\ 
\cline{2-4} 
\\[-1.8ex] & \multicolumn{2}{c}{speed} & dist \\ 
 & M1 & M2 & M3 \\ 
\hline \\[-1.8ex] 
 dist & 0.166$^{***}$ & 0.166$^{***}$ &  \\ 
  & (0.017) & (0.017) &  \\ 
  & & & \\ 
 speed &  &  & 3.932$^{***}$ \\ 
  &  &  & (0.416) \\ 
  & & & \\ 
 Constant & 8.284$^{***}$ & 8.284$^{***}$ & $-$17.579$^{**}$ \\ 
  & (0.874) & (0.874) & (6.758) \\ 
  & & & \\ 
\hline \\[-1.8ex] 
Observations & 50 & 50 & 50 \\ 
R$^{2}$ & 0.651 & 0.651 & 0.651 \\ 
Adjusted R$^{2}$ & 0.644 & 0.644 & 0.644 \\ 
Residual Std. Error (df = 48) & 3.156 & 3.156 & 15.380 \\ 
F Statistic (df = 1; 48) & 89.567$^{***}$ & 89.567$^{***}$ & 89.567$^{***}$ \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}

\hypertarget{figures}{%
\section{Figures}\label{figures}}

\hypertarget{graphs-with-r}{%
\subsection{Graphs with R}\label{graphs-with-r}}

You can insert figures like this. One would like to produce and insert them on
the fly in the \texttt{.rmd} file. Figure \ref{fig:fig-1} is such an example.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot}\NormalTok{(cars}\SpecialCharTok{$}\NormalTok{speed, cars}\SpecialCharTok{$}\NormalTok{dist)}
\end{Highlighting}
\end{Shaded}

\begin{figure}[h]

{\centering \includegraphics{articleCustomerDropoutMembership_files/figure-latex/Figures 1, fig-1-1}

}

\caption{Scatterplot of Speed and Distance}(\#fig:Figures 1, fig-1)
\end{figure}

However, in some cases it does not work.

\hypertarget{example-ggplot2-graphs}{%
\subsection{Example: ggplot2 graphs}\label{example-ggplot2-graphs}}

See the \texttt{ggplot2} output reported in Figure \ref{fig:fig-2}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{mtcars}\SpecialCharTok{$}\NormalTok{cyl }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(mtcars}\SpecialCharTok{$}\NormalTok{cyl) }\CommentTok{\# Convert cyl to factor}
\FunctionTok{library}\NormalTok{(ggplot2)}
\FunctionTok{ggplot}\NormalTok{(mtcars, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ wt, }\AttributeTok{y =}\NormalTok{ mpg, }\AttributeTok{shape =}\NormalTok{ cyl)) }\SpecialCharTok{+} \FunctionTok{geom\_point}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Weight (lb/1000)"}\NormalTok{, }\AttributeTok{y =} \StringTok{"Miles/(US) gallon"}\NormalTok{,}
       \AttributeTok{shape =} \StringTok{"Number of }\SpecialCharTok{\textbackslash{}n}\StringTok{ Cylinders"}\NormalTok{) }\SpecialCharTok{+} \FunctionTok{theme\_classic}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{figure}[h]

{\centering \includegraphics{articleCustomerDropoutMembership_files/figure-latex/Fig-2-1}

}

\caption{Miles per gallon according to the weight}\label{fig:Fig-2}
\end{figure}

\hypertarget{another-example-using-plotly}{%
\subsection{Another example using Plotly}\label{another-example-using-plotly}}

With \texttt{Plotly} we can produce interactive graphs which play well, for example,
once can embedded in html webpages (drop by
\href{https://paulcbauer.shinyapps.io/visualizing-causal-scenarios/}{here} for an
example). One can insert this type of graphs in R Markdown PDF using \texttt{Orca} (it
generates static images from Plotly graphs). Go
\href{https://github.com/plotly/orca\#installation}{here} to check how to install it.
See Figure \ref{fig:fig-3} for an example.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(plotly)}
\NormalTok{p }\OtherTok{\textless{}{-}} \FunctionTok{plot\_ly}\NormalTok{(cars, }\AttributeTok{type =} \StringTok{"scatter"}\NormalTok{, }\AttributeTok{mode =} \StringTok{"markers"}\NormalTok{,}
        \AttributeTok{x =} \SpecialCharTok{\textasciitilde{}}\NormalTok{speed,}
        \AttributeTok{y =} \SpecialCharTok{\textasciitilde{}}\NormalTok{dist)}
\CommentTok{\#Sys.setenv(\textquotesingle{}MAPBOX\_TOKEN\textquotesingle{} = \textquotesingle{}12423423\textquotesingle{}) \# set arbitrary token}
\CommentTok{\#orca(p, "logs/plotly{-}plot.pdf")}
\end{Highlighting}
\end{Shaded}

\begin{figure}[ht]
\centering
\caption{Example: export a Plotly figure using `orca`}\label{fig:fig-3}
    \includegraphics[width = 0.9\linewidth]{../figures/teste.png}
\begin{flushleft}
\end{flushleft}
\end{figure}
\vspace{-1.2cm}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Lets create a value for example}

\NormalTok{media }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(cars}\SpecialCharTok{$}\NormalTok{speed)}
\end{Highlighting}
\end{Shaded}

The criminal rate is 15.4\%o.

\vspace{0.3cm}

\hypertarget{miguels-tests}{%
\section{Miguel's tests}\label{miguels-tests}}

\hypertarget{r}{%
\subsection{R}\label{r}}

Example of an equation

\[\int_0^{2\pi} \sin x~dx\]

\emph{Example of a matrix}

\[
\mathbf{X} = \left[\begin{array}
{rrr}
1 & 2 & 3 \\
4 & 5 & 6 \\
7 & 8 & 9
\end{array}\right]
\]

or

\begin{equation}
f\left(k\right) = \binom{n}{k}p^k\left(1-p\right)^{n-k} \label{eq:binom}
\end{equation}

See Equation \eqref{eq:binom}.

\begin{align}
y_{ijt} = \beta x_{ijt} + \eta_i + \gamma_j + \lambda_t + \varepsilon_{ijt}
\end{align}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(stargazer)}
\FunctionTok{stargazer}\NormalTok{(cars,}
          \AttributeTok{title =} \StringTok{"Summary 24"}\NormalTok{,}
          \AttributeTok{label =} \StringTok{"tab24"}\NormalTok{,}
          \AttributeTok{table.placement =} \StringTok{"ht"}\NormalTok{,}
          \AttributeTok{header =} \ConstantTok{FALSE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}[ht] \centering 
  \caption{Summary 24} 
  \label{tab24} 
\begin{tabular}{@{\extracolsep{5pt}}lccccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Pctl(25)} & \multicolumn{1}{c}{Pctl(75)} & \multicolumn{1}{c}{Max} \\ 
\hline \\[-1.8ex] 
speed & 50 & 15.400 & 5.288 & 4 & 12 & 19 & 25 \\ 
dist & 50 & 42.980 & 25.769 & 2 & 26 & 56 & 120 \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}

\hypertarget{final-remarks}{%
\section{Final remarks}\label{final-remarks}}

Check the replication package for Bonhomme, Lamadon and Manresa (2019):
\url{https://github.com/tlamadon/blm-replicate}

\newpage

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\end{CSLReferences}

\hypertarget{appendix-chunk-options}{%
\section*{Appendix: Chunk options}\label{appendix-chunk-options}}
\addcontentsline{toc}{section}{Appendix: Chunk options}

\hypertarget{software-versioning}{%
\subsection{Software versioning}\label{software-versioning}}

\hypertarget{r-1}{%
\subsubsection{R}\label{r-1}}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{cat}\NormalTok{(}\FunctionTok{paste}\NormalTok{(}\StringTok{"\#"}\NormalTok{, }\FunctionTok{capture.output}\NormalTok{(}\FunctionTok{sessionInfo}\NormalTok{()), }\StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{, }\AttributeTok{collapse  =} \StringTok{""}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # R version 4.1.0 (2021-05-18) 
## # Platform: x86_64-w64-mingw32/x64 (64-bit) 
## # Running under: Windows 10 x64 (build 19043) 
## #  
## # Matrix products: default 
## #  
## # locale: 
## # [1] LC_COLLATE=Portuguese_Portugal.1252  LC_CTYPE=Portuguese_Portugal.1252    
## # [3] LC_MONETARY=Portuguese_Portugal.1252 LC_NUMERIC=C                         
## # [5] LC_TIME=Portuguese_Portugal.1252     
## #  
## # attached base packages: 
## # [1] stats     graphics  grDevices utils     datasets  methods   base      
## #  
## # other attached packages: 
## #  [1] plotly_4.9.4.1   kableExtra_1.3.4 gtsummary_1.4.2  readxl_1.3.1     
## #  [5] stargazer_5.2.2  naniar_0.6.1     visdat_0.5.3     ggplot2_3.3.4    
## #  [9] dlookr_0.4.5     dplyr_1.0.7      
## #  
## # loaded via a namespace (and not attached): 
## #   [1] webshot_0.5.2       RColorBrewer_1.1-2  httr_1.4.2          
## #   [4] UpSetR_1.4.0        tools_4.1.0         backports_1.2.1     
## #   [7] utf8_1.2.1          R6_2.5.0            rpart_4.1-15        
## #  [10] lazyeval_0.2.2      Hmisc_4.5-0         nortest_1.0-4       
## #  [13] colorspace_2.0-1    nnet_7.3-16         withr_2.4.2         
## #  [16] tidyselect_1.1.1    gridExtra_2.3       curl_4.3.1          
## #  [19] compiler_4.1.0      extrafontdb_1.0     cli_2.5.0           
## #  [22] rvest_1.0.0         gt_0.3.0            htmlTable_2.2.1     
## #  [25] xml2_1.3.2          sandwich_3.0-1      labeling_0.4.2      
## #  [28] bookdown_0.22       scales_1.1.1        checkmate_2.0.0     
## #  [31] mvtnorm_1.1-2       proxy_0.4-26        RcmdrMisc_2.7-1     
## #  [34] systemfonts_1.0.2   stringr_1.4.0       digest_0.6.27       
## #  [37] foreign_0.8-81      rmarkdown_2.9       svglite_2.0.0       
## #  [40] rio_0.5.27          base64enc_0.1-3     jpeg_0.1-8.1        
## #  [43] pkgconfig_2.0.3     htmltools_0.5.1.1   extrafont_0.17      
## #  [46] highr_0.9           htmlwidgets_1.5.3   rlang_0.4.11        
## #  [49] rstudioapi_0.13     prettydoc_0.4.1     farver_2.1.0        
## #  [52] generics_0.1.0      jsonlite_1.7.2      zoo_1.8-9           
## #  [55] crosstalk_1.1.1     zip_2.2.0           car_3.0-11          
## #  [58] magrittr_2.0.1      Formula_1.2-4       Matrix_1.3-3        
## #  [61] Rcpp_1.0.6          munsell_0.5.0       fansi_0.5.0         
## #  [64] reticulate_1.20     abind_1.4-5         gdtools_0.2.3       
## #  [67] partykit_1.2-13     lifecycle_1.0.0     stringi_1.6.1       
## #  [70] yaml_2.2.1          inum_1.0-4          carData_3.0-4       
## #  [73] MASS_7.3-54         plyr_1.8.6          grid_4.1.0          
## #  [76] hrbrthemes_0.8.0    forcats_0.5.1       crayon_1.4.1        
## #  [79] lattice_0.20-44     haven_2.4.1         splines_4.1.0       
## #  [82] hms_1.1.0           knitr_1.33          pillar_1.6.1        
## #  [85] glue_1.4.2          evaluate_0.14       latticeExtra_0.6-29 
## #  [88] broom.helpers_1.3.0 data.table_1.14.0   png_0.1-7           
## #  [91] vctrs_0.3.8         Rttf2pt1_1.3.8      cellranger_1.1.0    
## #  [94] tidyr_1.1.3         gtable_0.3.0        purrr_0.3.4         
## #  [97] xfun_0.24           openxlsx_4.2.4      libcoin_1.0-8       
## # [100] e1071_1.7-7         class_7.3-19        survival_3.2-11     
## # [103] viridisLite_0.4.0   tibble_3.1.2        cluster_2.1.2       
## # [106] corrplot_0.89       ellipsis_0.3.2
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
  \CommentTok{\# or use message() instead of cat()}
\end{Highlighting}
\end{Shaded}

\end{document}

### 2021-07-26 article/articleCustomerDropoutMembership.Rmd
upd

@@
   sub("CODE", code, html)
-
-##https://opensource.com/article/19/5/python-3-default-mac
-#Sys.setenv(RETICULATE_PYTHON = "C:/Users//AppData/Local/Microsoft/WindowsApps/python3")
-
-#install.packages("reticulate")
-#library(reticulate)
-#virtualenv_create("artigosurvival",python_version="3.7.10")
-#use_virtualenv("artigosurvival", required = TRUE)
 }   
@@
 ```{r RETICULATE,echo = FALSE}
+#Replace by your environment, usually which python solves the problem
+#conda env list also is a good option
 #set environment first
+switch(Sys.info()[['sysname']],
+       Windows = {
             Sys.setenv(RETICULATE_PYTHON="C:\\Users\\sobre\\anaconda3\\envs\\survival\\python.exe")
@@
             use_condaenv("survival", required = TRUE)
-#use_virtualenv("C:/Users/mangelo.EEG/Documents/python")
+       },
+       Linux = {
+            Sys.setenv(RETICULATE_PYTHON="/home/sobreiro/miniconda3/envs/survival/bin/python")
+            library(reticulate)
+            use_condaenv("survival", required = TRUE)
+       }
+)
 ```
@@
 library(kableExtra)
+library(labelled)
+
+var_label(df_members$year) <- "Inscription year"
+var_label(df_members$age) <- "Age in years"
+var_label(df_members$sex) <- "Male or female"
+var_label(df_members$marital_status) = "Single, married and other."
 
@@
     tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd})",
-                                 all_categorical() ~ "{p}%")
-               )
-as_kable_extra(tbl,booktabs=T,caption="Summary statistics")
+                                 all_categorical() ~ "{p}%"),
+                type = list(age ~ "continuous")
+               ) %>% add_stat_label()
+
+as_kable_extra(tbl,booktabs=T,caption="Summary statistics of features used")
 ```
@@
 correlation_matrix(df_members, figure_size=(10,10), text_fontsize=6)
+
+r.df_members = df_members
+
 ```
 
-The average age in our data is `r round(mean(df_members$age,na.rm = TRUE),1)`.
+The average age is `r round(mean(df_members$age,na.rm = TRUE),1)` &plusmn;
+`r round(sd(df_members$age,na.rm = TRUE),1)`.
+
+total_amount
+total_matches
+season_matches12%
+24%
+316 (494)
+27 (46)
+2.2 (4.1)
+months_since_last_payment 19 (32)
+dropout
+22%
+years_membership
+11 (11)
+
+```{r test}
+
+str(df_members)
+
+```
 

### 2021-07-27 article/articleCustomerDropoutMembership.Rmd
upd

@@
 abstract: \noindent\setstretch{1}Abstract of the article. Here we can place more info.\vspace{.8cm}
-bibliography: references.bib
+bibliography: refMembership.bib
 csl: american-sociological-association.csl
@@
     latex_engine: xelatex
+    number_sections: true
 mainfont: Times New Roman
@@
 
-```{r RETICULATE,echo = FALSE}
+```{r RETICULATE, message=FALSE, warning=FALSE, include=FALSE}
+# load essential libraries
+library(dplyr)
+library(dlookr)
+library(ggplot2)
+library(reticulate)
 #Replace by your environment, usually which python solves the problem
 #conda env list also is a good option
-#set environment first
-switch(Sys.info()[['sysname']],
+#set environment first then call reticulate library
+path_python_windows <- "C:\\Users\\sobre\\AppData\\Local\\r-miniconda\\envs\\rsurvival\\python.exe"
+path_python_linux <- "/home/sobreiro/miniconda3/envs/survival/bin/python"
+switch(Sys.info()[["sysname"]],
        Windows = {
-            Sys.setenv(RETICULATE_PYTHON="C:\\Users\\sobre\\anaconda3\\envs\\survival\\python.exe")
+            Sys.setenv(RETICULATE_PYTHON = path_python_windows)
             #call reticulate
@@
             #activate environment
-            use_condaenv("survival", required = TRUE)
+            use_condaenv("rsurvival", required = TRUE)
        },
        Linux = {
-            Sys.setenv(RETICULATE_PYTHON="/home/sobreiro/miniconda3/envs/survival/bin/python")
+            Sys.setenv(RETICULATE_PYTHON = path_python_linux)
             library(reticulate)
@@
 Research idea:
-
--
 
@@
 
-\clearpage
+Churn dropout prediction is a problem being addressed supported in the idea that
+the customers database is the most valuable asset that the organizations possess
+[@Athanassopoulos_2000], which requires determining customers that will attrite
+[@alboukaey_dynamic_2020]. Dropout implies in contractual business that the
+customer needs to renew their contracts to continue its usage
+[@Ascarza_Hardie_2013].
 
-\renewcommand{\baselinestretch}{0.5}\normalsize
+However, in contractual settings the customer dropout represents an explicit
+ending of a relationship more punitive than non contractual settings
+[@risselada_staying_2010] This has implications to the profitability of the
+organizations increasing marketing costs and reducing sales
+[@amin_customer_2017].
 
-\renewcommand{\baselinestretch}{1.1}\normalsize
+The anticipation of the dropout allows the development of countermeasures to
+reduce customer churn. Several studies address the problem related to customer
+retention trying to improve the profitability [@coussement_improving_2009;
+@garcia_intelligent_2017; @devriendt_why_2019]
 
-\clearpage
-
-```{r Stats1, echo = TRUE, message = FALSE}
-library(dplyr)
-library(dlookr)
-library(ggplot2)
-
-#eda_report(nlswork,output_dir  = 
-#	"C:/Users/mangelo.EEG/Documents/GitHub/prjs/reports/",
-#	output_file = "eda_report.pdf")
-
-## The data
-
-names(airquality)
-#summary(nlswork)
-
-## Missing values
-library(visdat)
-vis_dat(airquality)
-
-library(naniar)
-vis_miss(airquality)
-
-gg_miss_upset(airquality)
-## GRAPHS
-dplyr::glimpse(cars$Ozone)
-d <- density(airquality$Temp)
-plot(d)
-
-plot(airquality$Ozone, airquality$Solar.R)
-ggplot(airquality, aes(x = Ozone, y = Solar.R)) +
-geom_miss_point()
-
-ggplot(airquality, aes(x = Wind, y = Temp)) +
-geom_miss_point() +
-facet_wrap(vars(Month))
-
-stats <- summary(airquality$Temp)
-stats
-
-describe(airquality)
-
+``` {.markdown}
+If an organization can predict a possible dropout and
+develop countermeasures to avoid desertion, they can avoid customer defections that 
+lead to a loss of money. Reichheld (1996) evidenced that reducing dropout rates by 5% 
+(e.g., from 15% to 10% per year) could represent an increase in prots up to double, as 
+acquiring new customers costs 5 to 6 times more than retaining existing ones (Bhattacharya, 
+1998). Existing organizations are addressing this problem by shifting their target from 
+capturing new customers to preserving existing ones (García et al., 2017),
+as investments in retention strategies have higher returns than acquisitions (Coussement 
+& Van den Poel, 2009). The importance of customer retention to maintain organizational protability (Devriendt
+et al., 2019) leads to the problem of how to quantify the nancial impact of customer retention actions
+under the assumption that the organization goal should be related to the increase the lifetime of the
+customer to increase their prots. The customer lifetime value (CLV) allows us to measure this, as itr
 ```
 
-# Experimental Results
+1.  Address the global problem of customer dropout
+2.  The identification of approaches to predict dropout requires more than only,
+    address the prediction accuracy such as ... place existing studies
+    addressing this...
+3.  A lot of effort has been placed testing the accuracy of existing algorithms,
+    in this study we try to fill this gap and explore also a balancing between
+    the model interpretability, accuracy, and the investment required.
+4.  Dropout prediction problem related to the timings
 
-## Data description
+The approaches normally employing use a dependent variable representing dropout
+or non-dropout, without considering a dynamic perspective that the dropout risk
+changes over time [@Alboukaey_dynamic_2020]. The survival models try to solve
+this limitation [@routh_estimating_2020] capturing a temporal dimension of the
+customer dropout [@perianez_churn_2016]. @perianez_churn_2016 used survival
+analysis to predict also when the dropout will occur.
 
-```{r get_data, echo = FALSE, message = FALSE, warning=FALSE}
+Other studies proposed also the integration of several algorithms to improve the
+performance in the prediction of the dropout such the usage of clusters
+combinated with churn prediction [@hung_applying_2006; @gok_case_2015;
+@vijaya_sivasankar_2019]. The approach relies in the assumption that combining
+the customers in different clusters allows the improvement of the prediction
+accuracy. @vijaya_sivasankar_2019 suggested the adoption hybrid models combining
+more than one classier are achieving increased performance compared to those
+using single classifiers.
+
+There are several challenges around the timing related to dropout, or
+considering the dynamic behavior of the customer in the intent to drop out
+[@alboukaey_dynamic_2020]. The importance of understanding when dropout will
+occur and the risk when discarding the temporal perspective of the problem seems
+to be an element that should be addressed. Few studies considered this
+[@perianez_churn_2016; @burez_separating_2008]. This shows an opportunity to
+address the importance of the timeframe and its influence on the efficiency of
+the model and also evalute if the combination of clusters could improve the
+performance.
+
+In this study, we adopt random survival forests which have never been used in
+understanding factors affecting membership in a sport club using existing data
+in a Sport Club. The analysis is based on the use of random survival forests in
+the presence of covariates that do not necessarily satisfy the PH assumption.
+Additionly we also propose a new approach combining clusters with survival
+analysis.
+
+??? Add interpretability layer
+
+Random Survival Forests does not make the proportional hazards assumption
+[@Ehrlinger_2016] and has the flexibility to model survivor curves that are of
+dissimilar shapes for contrasting groups of subjects. Random Survival Forest is
+an extension of Random Forest allowing efficient non-parametric analysis of time
+to event data [@Breiman_2001]. This characteristics allow us to surpass the Cox
+Regression limitation of the proportional hazard assumption, requiring to
+exclude variables which not fulfill the model assumption. It was shown by
+@Breiman_2001 that ensemble learning can be further improved by injecting
+randomization into the base learning process - a method called Random Forests.
+
+# Methodology
+
+Dropout is a binary value where one represent churn and zero not churn. The dropout
+happens when a member does not have a payment ...
+
+## Dataset
+
+### Data description
+
+```{r get_data, echo = FALSE, message = FALSE, warning=FALSE, include = FALSE}
 library(stargazer)
@@
 
-Table \@ref(tab:tab1cars) shows data's summary statistics.[^1] `stargazer()` is
-and excellent solution to export outputs.
-
-[^1]: You can reference the table as \@ref(tab1cars).
-
-```{r Export Stats0, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
+Table \@ref(tab:summarytable) shows data's summary statistics. 
+```{r teste, eval=FALSE, message=FALSE, warning=FALSE, echo=FALSE, paged.print=FALSE}
 library(stargazer)
-df_summary <- summary(df_members, object)
+df_summary <- summary(df_members)
 stargazer(df_members,
@@
          )
-
 ```
-
-Teste \@ref(tab:summarytable)
-
 ```{r summarytable, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
@@
 ```
+The average age is `r round(mean(df_members$age,na.rm = TRUE),1)` ±
+`r round(sd(df_members$age,na.rm = TRUE),1)`, the members have an attendance 
+`r round(mean(df_members$total_matches,na.rm = TRUE),0)` ± `r 
+round(sd(df_members$total_matches,na.rm = TRUE),1)` with a membership
+`r round(mean(df_members$years_membership,na.rm = TRUE),0)` ± `r 
+round(sd(df_members$years_membership,na.rm = TRUE),1)`.
+Figure \@ref(fig:membershipyear) shows the distribution of the dropout considering the inscription year.
 
-```{python Py,eval=TRUE,echo=TRUE}
+```{r membershipyear, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Number of members by year"}
+members_year<-df_members%>% 
+  select(year,dropout) %>%
+  group_by(year,dropout) %>%
+  summarize(count = n())
+
+members_year$dropout = factor(members_year$dropout)  
+# scale_y_continuous(breaks=c(2,4,6,8,10))+
+g<-ggplot(data = members_year,mapping = aes(x = year,y = count,linetype = dropout)) 
+g+geom_line()+labs(x='year',y='Number of members') +
+  theme_classic()
+```
+
+## TODO
+
+The random survival forest was developed using the package PySurvival
+[@Fotso_others_2019]. The most relevant variables predicting the dropout are
+analysed using the log-rank test. The metric variables are transformed to
+categorical using the quartiles to provide a statistical comparison of groups.
+The survival analysis was conducted using the package Lifelines
+[@Davidson-Pilon_2021].
+
+PySurvival is an open source python package for Survival Analysis modeling - the
+modeling concept used to analyze or predict when an event is likely to happen.
+It is built upon the most commonly used machine learning packages such NumPy,
+SciPy and PyTorch. PySurvival is compatible with Python 2.7-3.7
+
+## Running the model
+
+    from pysurvival.utils.display import correlation_matrix
+    correlation_matrix(df[features], figure_size=(10,10), text_fontsize=8)
+
+```{python Py,eval=TRUE, echo=FALSE}
 from pysurvival.utils.display import correlation_matrix
@@
 
-The average age is `r round(mean(df_members$age,na.rm = TRUE),1)` &plusmn;
-`r round(sd(df_members$age,na.rm = TRUE),1)`.
+Removed the variables with greater correlations
 
-total_amount
-total_matches
-season_matches12%
-24%
-316 (494)
-27 (46)
-2.2 (4.1)
-months_since_last_payment 19 (32)
-dropout
-22%
-years_membership
-11 (11)
+    to_remove = ['totalJogos', 'idaEstadio']
+    features = np.setdiff1d(features, to_remove).tolist()
 
-```{r test}
+## Model building
 
-str(df_members)
+The model was built with with 60% of the data for training and 40% for testing.
+The survival model parameters where:
 
+    from pysurvival.models.survival_forest import RandomSurvivalForestModel
+    csf = RandomSurvivalForestModel(num_trees=200)
+    csf.fit(X_train, T_train, E_train, max_features='sqrt', max_depth=5, min_node_size=20)
+
+The model accuracy is very high in the first years. The prediction is very
+similar to the actual value.
+
+![Prediction accuracy](../analysis/predictionAccuracy.png)
+
+All the outputs are available [here](../analysis/01.survRandomForest.pdf)
+
+# Open questions - to remove
+
+```{=tex}
+\begin{itemize}
+    \item RQ1: What is the current state of the research being developed?
+    \item RQ2: What algorithms have been used to predict dropout?
+    \item RQ3: What are the features used to predict dropout?
+    \item RQ4: When does dropout occur?
+    \item RQ5: How is the accuracy of the machine learning algorithms in predicting dropout measured?
+\end{itemize}
+```
+From RQ1, it was possible to identify some business areas that are
+under-researched, such as the energy sector, education, logistics and
+hospitality. Compared to other business areas such telecom or the financial
+sector, research on the energy sector or water supply is lacking, considering
+the contractual settings that are assumed to provide such types of services.
+Considering the business model of many software companies as software as a
+services (SaS), the number of research works is also surprisingly low.
+
+RQ2 also provided an overall perspective related to the algorithms being used to
+predict customer dropout. The first could be the importance and wider adoption
+of decision trees and random forests [@antipov_applying_2010;
+@benoit_improving_2012; @burez_crm_2007], and logistic regression
+[@coussement_improved_2010], which could be due to its higher interpretability
+and flexibility [@keramati_improved_2014]. Interpretability is an important
+aspect for the marketing department in the extraction of valuable information
+from the model to develop effective retention strategies [@verbeke_new_2012].
+The problem arises in the balancing between interpretability and the higher
+performance of the algorithms inspired by nature (such as neural networks). From
+a business perspective, dropout prediction should also be considered as a
+business objective, which requires more than predicting if the customer will
+churn or not [@devriendt_why_2019], where higher interpretability provides
+better support in the development of retention strategies. The developed SLR
+also raises the possibility of integrating different algorithms using ensemble
+methods or integrating several models using a hybrid approach. None of the
+studies integrated the survival approach to predict customer dropout, for
+example, using a hybrid approach.
+
+It is considered positive if actions are developed to retain customers, but the
+problems should also be considered, such as the following: (1) customers who
+have greater risk of dropout should be targeted to provide a base for a better
+ROI in the retention strategies [@xie_customer_2009; @coussement_churn_2008] and
+(2) the retention strategies should be developed focused on customers with
+higher satisfaction, or its inclusion could be a reminder of the contractual
+agreement nearing an end and could lead to churn [@devriendt_why_2019].
+
+From RQ3, several types of features being used were able to be identified, such
+as demographic, behavioral, and economic indicators, pictorial data, network
+relationships or high cardinality features. The problem that arises is that some
+studies used data and features that were not described, and this creates a major
+issue, How can reproducibility be developed in a study without the availability
+of the data or the identification of the features used? Considering that science
+is driven by data, with the development of new technologies, the increasing
+complexity of research and the amount of data collected, the challenge is to
+ensure that research is available to all [@Hanson_Sugden_Alberts_2011]; this
+requires both availability of the data and the algorithms so that they can be
+explored by other researchers. The features are selected mainly to verify the
+performance of the models, and are essential to performance prediction,
+accuracy, and the steps for processing the data, which are fundamental to
+improve the model accuracy [@azeem_churn_2017].
+
+There are several challenges around the timing related to dropout, or
+considering the dynamic behavior of the customer in the intent to drop out
+[@alboukaey_dynamic_2020]. The importance of understanding when dropout will
+occur and the risk when discarding the temporal perspective of the problem seems
+to be an element that should be addressed. Few studies considered this
+[@perianez_churn_2016; @burez_separating_2008]. This shows an opportunity to
+address the importance of the timeframe and its influence on the efficiency of
+the model.
+
+According to each business model, the timeframe could be addressed considering
+the survival probability according to the customer relationship age, and dropout
+predictions could be developed according to these survival probabilities, as
+suggested by @esteves_churn_2016, to investigate which data timeframe produces
+the best result and how the efficiency of the models is influenced by this
+timeframe. Exploring the duration of the relation and the understanding of the
+features that increase or decrease that duration seems to be an important
+approach that could complement the existing approaches to predicting dropout.
+
+From RQ5, the literature analysis showed that different types of questions
+arise. Which are the best approaches to develop the analysis of the performance
+in predicting dropout? Several metrics are customer dropout is to improve the
+performance of organizations in retaining customers, which is a management
+problem in which data mining is adopted [@verbeke_new_2012]. The goals of the
+model should be formulated considering the context of the problem that is being
+addressed; in marketing retention strategies, the up-lift supports the
+development of proactive actions to minimize the investment in retention
+strategies [@coussement_churn_2008]. Some assumptions that underlie the adoption
+of uplift metrics consider that customers with a higher risk of churning could
+not be the best targets, as suggested by @ascarza_retention_2018. Other
+researchers addressed the problem using the top-decile lift to develop more
+proactive actions to retain the customers at risk of churning
+[@coussement_churn_2008;xie_customer_2009]. This approach considers the 10% of
+customers with more risk, and investments in retention strategies should be
+developed that distinguish the churners susceptible to marketing actions from
+those who will leave anyway [@coussement_comparative_2017]. Although uplift
+models seem to be good strategies, they should also used, such as AUC,
+sensitivity, specificity, recall, precision, and F-score. However, the goal
+ofconsider factors other than risk and customer satisfaction, as not taking this
+into consideration could be counterproductive and the model should be removed
+from the retention strategy.
+
+The true business objective is to reduce customer churn. Customers who are about
+to churn but cannot be retained should be excluded from the campaign, as
+targeting them will be a waste of scarce resources [@devriendt_why_2019]. Using
+these models seem to be a good strategy, as they can outperform predictive
+models that consider only accuracy from a profitability busshould be considered
+that customers with a higher risk of churning may not be the best targets to
+develop retention strategies. Those perspectives entail the dropout.
+
+that a business context, or the clarification of a business objective underlying
+the prediction of customer dropout, should be developed, to clarify which
+objectives should be achieved before employing the profitability of reducing g
+machine learning algorithms. Surprisingly, the analyzed studies did not address
+the customer lifetime value as an objective to optimize considerininess
+perspective.
+
+# Aspects to consider
+
+-   Interpretability from RQ2
+-   The business objective is to increase the number of members and organization
+    profits
+-   piping several algorithms to improve accuracy. Aka hybrid approach
+-   @Alboukaey_dynamic_2020 proposes ...
+-   grep the articles addressing hybrid: pdfgrep -ri "hybrid.{1,10} approach"
+
+# Results
+
+In this section, we present our experiments to validate the proposed models,
+comparing against other approaches. The models where optimized using the
+hyper-parameters Grid Search technique. The explored hyper-parameters and the
+best values of these parameters for every model are listed in (ref:table1).
+
+(ref:table1) Table 2
+
+```{=latex}
+\begin{table}[ht]
+    \footnotesize
+    \centering
+    \begin{tabular}{p{0.15\textwidth}p{0.55\textwidth}p{0.25\textwidth}}
+    \hline
+    \textbf{Model name} & \textbf{Explored parameters values} & \textbf{Best parameters} \\
+    \hline
+    Survival trees  & pysurvival random forest                  & a \\
+    Survival trees
+    with clusters   & pysurvival random forest with clusters    & a \\
+    Scikit survival 
+    trees           & scikit survival                           & a \\
+    Scikit survival 
+    with clusters   & scikit with clusters                      & a \\
+    Scikit survival 
+    gradient boost  & scikit survival gradient boost            & a \\
+    Scikit survival 
+    gradient boost 
+    with clusters   & scikit with clusters                      & a \\
+    \hline
+    \end{tabular}
+    \caption{Hyper-parameters best values}
+    \label{hyperparametersbestvalues}
+\end{table}
+```
+```{=html}
+<!--
+|        Model name        | Explored hyper-parameters' values                    | Best parameters |
+| ------------------------ | ---------------------------------------------------- | --------------- |
+| Survival trees           |                                                      |                 |   
+| without cluster          |                                                      |                 |  
+-->
 ```
 
-# Tables {#sec:tables}
+# Conclusion
 
-R Markdown PDF is now able to produce good tables with our output. For
-`stargazer` the label is contained in the function, while for `kable` it's
-contained in the chunk name.
+Article Ascarza
 
-## stargazer(): Summary and regression tables
+-   Retention Futility: Targeting High-Risk Customers Might be Ineffective
+    [@ascarza_retention_2018]
 
-Table \@ref(tab2) reports regression outputs. Name the models as you can refer
-to their names in the text (M1, M2, M3).
+Ascarza, E. (2018). Retention Futility: Targeting High-Risk Customers Might be
+Ineffective. Journal of Marketing Research, 55(1), 80-98. sim.
+<https://doi.org/10.1509/jmr.16.0163>
 
-```{r Export Regressions, echo = TRUE, message = FALSE, warning = FALSE, paged.print = FALSE, results = "asis"}
-library(stargazer)
-model1 <- lm(speed ~ dist, data = cars)
-model2 <- lm(speed ~ dist, data = cars)
-model3 <- lm(dist ~ speed, data = cars)
-stargazer(model1, model2, model3,
-          title = "Regression table with stargazer",
-          label = "tab2",
-          table.placement = "h",
-          column.labels = c("M1", "M2", "M3"),
-          model.numbers = FALSE,
-          header = FALSE)
+Example of Developed actions:
+
+``` {.markdown}
+Each month, the company identified the customers who were up for renewal and
+split them (randomly and evenly) between a treatment group that received a 
+"thank you" gift with the letter and a control group that received only the 
+renewal latter.
 ```
-
-# Figures
-
-## Graphs with R
-
-You can insert figures like this. One would like to produce and insert them on
-the fly in the `.rmd` file. Figure \@ref(fig:fig-1) is such an example.
-
-```{r Figures 1, fig-1, fig.align = "center", fig.cap = "Scatterplot of Speed and Distance", fig.pos = "h", message = FALSE, warning = FALSE, paged.print = FALSE}
-plot(cars$speed, cars$dist)
-```
-
-However, in some cases it does not work.
-
-## Example: ggplot2 graphs
-
-See the `ggplot2` output reported in Figure \@ref(fig:fig-2).
-
-```{r Fig-2, fig.align = "center", fig.cap = "Miles per gallon according to the weight", fig.pos = "h", fig.width = 6, fig.height = 3, message = FALSE, warning = FALSE, paged.print = FALSE}
-mtcars$cyl <- as.factor(mtcars$cyl) # Convert cyl to factor
-library(ggplot2)
-ggplot(mtcars, aes(x = wt, y = mpg, shape = cyl)) + geom_point() +
-  labs(x = "Weight (lb/1000)", y = "Miles/(US) gallon",
-       shape = "Number of \n Cylinders") + theme_classic()
-```
-
-## Another example using Plotly
-
-With `Plotly` we can produce interactive graphs which play well, for example,
-once can embedded in html webpages (drop by
-[here](https://paulcbauer.shinyapps.io/visualizing-causal-scenarios/) for an
-example). One can insert this type of graphs in R Markdown PDF using `Orca` (it
-generates static images from Plotly graphs). Go
-[here](https://github.com/plotly/orca#installation) to check how to install it.
-See Figure \@ref(fig:fig-3) for an example.
-
-```{r Fig-3, message = FALSE, warning = FALSE}
-library(plotly)
-p <- plot_ly(cars, type = "scatter", mode = "markers",
-        x = ~speed,
-        y = ~dist)
-#Sys.setenv('MAPBOX_TOKEN' = '12423423') # set arbitrary token
-#orca(p, "logs/plotly-plot.pdf")
-```
-
-```{=tex}
-\begin{figure}[ht]
-\centering
-\caption{Example: export a Plotly figure using `orca`}\label{fig:fig-3}
-    \includegraphics[width = 0.9\linewidth]{../figures/teste.png}
-\begin{flushleft}
-\end{flushleft}
-\end{figure}
-\vspace{-1.2cm}
-```
-```{r exemploMedia}
-# Lets create a value for example
-
-media <- mean(cars$speed)
-
-```
-
-The criminal rate is `r media`%o.
-
-\vspace{0.3cm}
-
-# Miguel's tests
-
-## R
-
-Example of an equation
-
-$$\int_0^{2\pi} \sin x~dx$$
-
-*Example of a matrix*
-
-$$
-\mathbf{X} = \left[\begin{array}
-{rrr}
-1 & 2 & 3 \\
-4 & 5 & 6 \\
-7 & 8 & 9
-\end{array}\right]
-$$
-
-or
-
-```{=tex}
-\begin{equation}
-f\left(k\right) = \binom{n}{k}p^k\left(1-p\right)^{n-k} \label{eq:binom}
-\end{equation}
-```
-See Equation \@ref(eq:binom).
-
-```{=tex}
-\begin{align}
-y_{ijt} = \beta x_{ijt} + \eta_i + \gamma_j + \lambda_t + \varepsilon_{ijt}
-\end{align}
-```
-```{r echo = TRUE, message = FALSE, warning = FALSE, paged.print = FALSE, results = "asis"}
-library(stargazer)
-stargazer(cars,
-          title = "Summary 24",
-          label = "tab24",
-          table.placement = "ht",
-          header = FALSE)
-```
-
-# Final remarks
-
-Check the replication package for Bonhomme, Lamadon and Manresa (2019):
-<https://github.com/tlamadon/blm-replicate>
-
-\newpage
 
@@
 ```
+
+### Other used tools
+
+-   [Visidata](https://www.visidata.org/) for quick exploratory. VisiData is a
+    free, open-source tool that lets you quickly open, explore, summarize, and
+    analyze datasets in your computer's terminal.

### 2021-07-27 article/articleCustomerDropoutMembership.Rmd
upd

@@
 Table \@ref(tab:summarytable) shows data's summary statistics. 
+The average age is `r round(mean(df_members$age,na.rm = TRUE),1)` ±
+`r round(sd(df_members$age,na.rm = TRUE),1)`, the members have an attendance 
+`r round(mean(df_members$total_matches,na.rm = TRUE),0)` ± `r 
+round(sd(df_members$total_matches,na.rm = TRUE),1)` with a membership
+`r round(mean(df_members$years_membership,na.rm = TRUE),0)` ± `r 
+round(sd(df_members$years_membership,na.rm = TRUE),1)`.
+
 ```{r teste, eval=FALSE, message=FALSE, warning=FALSE, echo=FALSE, paged.print=FALSE}
@@
 ```
-The average age is `r round(mean(df_members$age,na.rm = TRUE),1)` ±
-`r round(sd(df_members$age,na.rm = TRUE),1)`, the members have an attendance 
-`r round(mean(df_members$total_matches,na.rm = TRUE),0)` ± `r 
-round(sd(df_members$total_matches,na.rm = TRUE),1)` with a membership
-`r round(mean(df_members$years_membership,na.rm = TRUE),0)` ± `r 
-round(sd(df_members$years_membership,na.rm = TRUE),1)`.
 Figure \@ref(fig:membershipyear) shows the distribution of the dropout considering the inscription year.

### 2021-07-27 article/articleCustomerDropoutMembership.Rmd
upd

@@
   |
-  | \vspace{1cm}Customer dropout membership^[Corresponding address: sobreiro@esdrm.ipsantarem.pt.
-The current template adapts part of the Rmd code by [Paul C.
-Bauer](https://github.com/paulcbauer/Writing_a_reproducable_paper_in_rmarkdown), Mannheim Centre for European Social Research.]\vspace{0.5cm}
+  | \vspace{1cm}Customer dropout membership^[Corresponding address: sobreiro@esdrm.ipsantarem.pt. Quality of Life Research Centre, Polytechnic Institute of Santarém, Portugal]\vspace{0.5cm}
 author: |
@@
 
-Dropout is a binary value where one represent churn and zero not churn. The dropout
-happens when a member does not have a payment ...
+Dropout is a binary value where one represent churn and zero not churn. The
+dropout happens when a member does not have a payment ...
+
+The model performance was determined with the concordance probability (C-index),
+Brier Score (BS) and Mean Absolute Error (MAE) [@wangmachine2017]. The feature
+importance was determined calculating the difference between the true class
+label and noised data [@Breiman_2001].
 
 ## Dataset
-
-### Data description
 
@@
 # rename column labels
-names(df_members) <- c("num_socio", "dt_inscription", "year", "birth_date",
+names(df_members) <- c("num_socio", "dt_inscription", "year_inscription", "birth_date",
                        "age", "sex", "marital_status", "category",
@@
 df_members <- df_members %>%
-select(year, age, sex, marital_status, monthly_fee, total_amount, total_matches,
+select(age, sex, marital_status, monthly_fee, total_amount, total_matches,
        season_matches, months_since_last_payment, dropout, years_membership,
@@
 
-Table \@ref(tab:summarytable) shows data's summary statistics. 
-The average age is `r round(mean(df_members$age,na.rm = TRUE),1)` ±
-`r round(sd(df_members$age,na.rm = TRUE),1)`, the members have an attendance 
-`r round(mean(df_members$total_matches,na.rm = TRUE),0)` ± `r 
-round(sd(df_members$total_matches,na.rm = TRUE),1)` with a membership
-`r round(mean(df_members$years_membership,na.rm = TRUE),0)` ± `r 
-round(sd(df_members$years_membership,na.rm = TRUE),1)`.
+Table \@ref(tab:summarytable) shows data's summary statistics. The average age
+is `r round(mean(df_members$age,na.rm = TRUE),1)` ±
+`r round(sd(df_members$age,na.rm = TRUE),1)`, the members have an attendance of
+`r round(mean(df_members$total_matches,na.rm = TRUE),0)` ±
+`r  round(sd(df_members$total_matches,na.rm = TRUE),1)` with a membership of
+`r round(mean(df_members$years_membership,na.rm = TRUE),0)` ±
+`r  round(sd(df_members$years_membership,na.rm = TRUE),1)` years.
 
@@
 ```
+
 ```{r summarytable, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
@@
 
-var_label(df_members$year) <- "Inscription year"
 var_label(df_members$age) <- "Age in years"
@@
                                  all_categorical() ~ "{p}%"),
-                type = list(age ~ "continuous")
-               ) %>% add_stat_label()
+                type = list(age ~ "continuous")) %>%
+    add_stat_label()
 
-as_kable_extra(tbl,booktabs=T,caption="Summary statistics of features used")
+as_kable_extra(tbl, booktabs = T,
+               caption = "Summary statistics of features used")
 ```
-Figure \@ref(fig:membershipyear) shows the distribution of the dropout considering the inscription year.
+
+Figure \@ref(fig:membershipyear) shows the distribution of the dropout
+considering the number of years of membership.
 
 ```{r membershipyear, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Number of members by year"}
-members_year<-df_members%>% 
-  select(year,dropout) %>%
-  group_by(year,dropout) %>%
+members_year <- df_members %>%
+  select(years_membership, dropout) %>%
+  group_by(years_membership, dropout) %>%
   summarize(count = n())
 
-members_year$dropout = factor(members_year$dropout)  
-# scale_y_continuous(breaks=c(2,4,6,8,10))+
-g<-ggplot(data = members_year,mapping = aes(x = year,y = count,linetype = dropout)) 
-g+geom_line()+labs(x='year',y='Number of members') +
+members_year$dropout <- factor(members_year$dropout)
+g <- ggplot(data = members_year,
+            mapping = aes(x = years_membership, y = count,linetype = dropout))
+g + geom_line() + labs(x = "year", y = "Number of members") +
     theme_classic()
@@
 
-## TODO
+## Model construction
+
+Address the model construction... the categorical variables $sex$,
+$marital\_status$ and $quart\_stadium\_entries$ where converted to dummy
+variables.
 
@@
 
-## Running the model
+### Survival trees based model
 
-    from pysurvival.utils.display import correlation_matrix
-    correlation_matrix(df[features], figure_size=(10,10), text_fontsize=8)
+In this model... The survival trees based model uses pysurvival random forest
 
@@
 
-correlation_matrix(df_members, figure_size=(10,10), text_fontsize=6)
+correlation_matrix(df_members[features], figure_size=(10,10), text_fontsize=6)
 
 r.df_members = df_members
-
 ```
 
-Removed the variables with greater correlations
+Removed the variables with greater correlations $total\_matches$ and
+$quart\_stadium\_entries$
 
-    to_remove = ['totalJogos', 'idaEstadio']
-    features = np.setdiff1d(features, to_remove).tolist()
+```{python removeVar, eval=TRUE, echo=FALSE, warning=FALSE}
+to_remove = ['total_matches', 'quart_stadium_entries_ate 1']
 
-## Model building
+#consider also the features previously removed
+features = np.setdiff1d(df_members[features].columns, to_remove).tolist()
+
+df_members.drop(columns = to_remove, inplace=True)
+
+correlation_matrix(df_members[features], figure_size=(10,10), text_fontsize=6)
+```
+
+```{python createModel, eval=TRUE, echo=FALSE, cache = TRUE, warning = FALSE}
+from pysurvival.models.survival_forest import RandomSurvivalForestModel
+
+# Number of samples in the dataset
+N = df_members.shape[0]
+
+# Building training and testing sets
+from sklearn.model_selection import train_test_split
+index_train, index_test = train_test_split( range(N), test_size = 0.4)
+data_train = df_members.loc[index_train].reset_index( drop = True )
+data_test  = df_members.loc[index_test].reset_index( drop = True )
+
+# Creating the X, T and E inputs
+X_train, X_test = df_members[features], data_test[features]
+T_train, T_test = df_members[time_column], data_test[time_column]
+E_train, E_test = df_members[event_column], data_test[event_column]
+
+# Fitting the model
+csf = RandomSurvivalForestModel(num_trees=200)
+csf.fit(X_train, T_train, E_train, max_features='sqrt',
+        max_depth=5, min_node_size=20)
+```
+
+Table \@ref(tab:summarytable2) shows variables importance.
+
+```{r summarytable2, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
+tbl <- py$csf$variable_importance_table
+kbl(tbl, booktabs = T, caption = "Summary statistics of features used")
+```
+
+#### Model building
 
@@
 
-    from pysurvival.models.survival_forest import RandomSurvivalForestModel
-    csf = RandomSurvivalForestModel(num_trees=200)
-    csf.fit(X_train, T_train, E_train, max_features='sqrt', max_depth=5, min_node_size=20)
+```{python pysurvivalperformance, eval=TRUE, echo=FALSE, cache=TRUE, warning = FALSE}
+from pysurvival.utils.metrics import concordance_index
+from pysurvival.utils.display import integrated_brier_score
+from pysurvival.utils.display import compare_to_actual
+
+c_index = concordance_index(csf, X_test, T_test, E_test)
+
+ibs = integrated_brier_score(csf, X_test, T_test, E_test, t_max=12,
+    figure_size=(12,5))
+
+results = compare_to_actual(csf, X_test, T_test, E_test, is_at_risk = False,  
+figure_size=(12, 6), metrics = ['rmse', 'mean', 'median'])
+```
 
 The model accuracy is very high in the first years. The prediction is very
-similar to the actual value.
+similar to the actual value. The absolute error mean of \~37 customers.
 
-![Prediction accuracy](../analysis/predictionAccuracy.png)
+### Survival trees based model with clusters
 
-All the outputs are available [here](../analysis/01.survRandomForest.pdf)
+Here we are will create clusters and developed the optimization within each
+cluster...
+
+The calculation of he number of clusters used the package mclust
+[@scrucca2016] using the Bayesian Information Criterion (BIC). The model that
+gives the minimum BIC score can be selected as the best model [@schwarz1978]
+simplifying the problem related to choosing the number of components and
+identifying the structure of the covariance matrix, based on modelling with
+multivariate normal distributions for each component that forms the data set
+[@akogul2016].
+
+In multivariate models are available the following approaches:
+
+-   "EII"spherical, equal volume
+-   "EEE"ellipsoidal, equal volume, shape, and orientation
+-   "VII"spherical, unequal volume
+-   "VVV"ellipsoidal, varying volume, shape, and orientation, which is used as
+    default for initialization of EM algorithm
+-   VVI": diagonal, varying volume and shape t
+
+estava a aqui a confirmar os clusters, talvez seja melhor reduzir as variáveis...
+
+```{r clusters, echo=FALSE, message=FALSE, warning=FALSE}
+library(mclust)
+y <- scale(py$df_members)
+bic <- mclustBIC(y,modelNames = c("VEI"))
+plot(bic)
+
+temp = data.frame(bic[,'VEI'])
+temp$cluster = row.names(temp)
+
+names(temp) = c("val","cluster")
+
+kbl(temp, booktabs = T, caption = "Summary statistics of features used")
+
+mod1 <- Mclust(py$df_members,x=BIC)
+summary(mod1)
+
+#plot(mod1, what="uncertainty")
+```
+
+TODO:
 
@@
 ```
-
 # Conclusion

### 2021-07-30 article/articleCustomerDropoutMembership.Rmd
upd

@@
 performance in the prediction of the dropout such the usage of clusters
-combinated with churn prediction [@hung_applying_2006; @gok_case_2015;
+combined with churn prediction [@hung_applying_2006; @gok_case_2015;
 @vijaya_sivasankar_2019]. The approach relies in the assumption that combining
@@
 
-```{python pysurvivalperformance, eval=TRUE, echo=FALSE, cache=TRUE, warning = FALSE}
+```{python pysurvivalperformance, eval=TRUE, echo=FALSE, cache=FALSE, warning = FALSE}
 from pysurvival.utils.metrics import concordance_index
@@
 The model accuracy is very high in the first years. The prediction is very
-similar to the actual value. The absolute error mean of \~37 customers.
+similar to the actual value. The absolute error mean of 
+`r round(py$results$mean_absolute_error,0)` customers.
 
@@
 
+Estou a ter problemas com o cálculo dos clusters com o BIC... 
 estava a aqui a confirmar os clusters, talvez seja melhor reduzir as variáveis...
+testar a abordagem aos clusters no artigo dos vinhos...
 
-```{r clusters, echo=FALSE, message=FALSE, warning=FALSE}
+```{r clusters, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
 library(mclust)
@@
 
-TODO:
+```{r NbClust, eval=FALSE}
+library(NbClust)
+nb <- NbClust(y, diss=NULL, distance = "euclidean", 
+              min.nc=2, max.nc=5, method = "kmeans", 
+              index = "all", alphaBeale = 0.1)
+hist(nb$Best.nc[1,], breaks = max(na.omit(nb$Best.nc[1,])))
+
+```
+
+```{python elbowCalculation, eval=TRUE, echo=FALSE, cache=TRUE, warning = FALSE}
+from sklearn.cluster import KMeans
+from sklearn import preprocessing
+from scipy.spatial.distance import cdist
+import matplotlib.pyplot as plt
+
+import numpy as np
+#Finding optimal no. of clusters
+clusters=range(1,20)
+meanDistortions=[]
+
+for k in clusters:
+    model=KMeans(n_clusters=k)
+    model.fit(df_members)
+    prediction=model.predict(df_members)
+    meanDistortions.append(sum(np.min(cdist(df_members, model.cluster_centers_, 'euclidean'), axis=1)) / df_members.shape[0])
+
+plt.plot(clusters, meanDistortions, 'bx-')
+plt.xlabel('k')
+plt.ylabel('Average distortion')
+plt.show()
+
+```
+
+We are going to consider five clusters
+
+```{python kmeans calculation, eval=TRUE, echo=FALSE, cache=FALSE, warning = FALSE}
+cluster = KMeans(n_clusters=5)
+
+cluster.fit(df_members)
+df_members['cluster']=cluster.predict(df_members)
+print(df_members.cluster.value_counts())
+```
+
+```{python kmeans PCA, eval=TRUE, echo=FALSE, cache=FALSE, warning = FALSE}
+from sklearn.decomposition import PCA
+
+plt.rcParams['figure.figsize'] = [13, 4]
+
+pca=PCA(n_components=2)
+df_members['X_pca']=pca.fit_transform(df_members)[:,0] #todas as linhas da primeira coluna com redução
+df_members['Y_pca']=pca.fit_transform(df_members)[:,1] #todas as linhas da segunda coluna com redução
+
+fig, ax = plt.subplots()
+for cluster in df_members.cluster.unique():
+    x = df_members['X_pca'].loc[df_members.cluster == cluster]
+    y = df_members['Y_pca'].loc[df_members.cluster == cluster]
+    ax.scatter(x, y, label=cluster, alpha=1, edgecolors='none')
+
+ax.legend()
+ax.grid(True)
+ax.set_title('Clusters clientes')
+```
+
+TODO: fit the model for the five clusters... compare the performance with the model without
+clusters... corrigir acima para selecionar só as features
+
+```{python createModelClusters, eval=FALSE, echo=FALSE, cache = TRUE, warning = FALSE}
+# Building training and testing sets for the clusters
+for cluster in df_members.cluster.unique():
+    # Number of samples in the dataset
+    df_members_cluster = df_members[df_members.cluster == cluster]
+    N = df_members_cluster.shape[0]
+    index_train, index_test = train_test_split(range(N), test_size = 0.4)
+    data_train = df_members_cluster[features].loc[index_train].reset_index( drop = True )
+    data_test  = df_members_cluster[features].loc[index_test].reset_index( drop = True )
+
+    # Creating the X, T and E inputs
+    X_train, X_test = df_members_cluster[features], data_test[features]
+    T_train, T_test = df_members_cluster[time_column], data_test[time_column]
+    E_train, E_test = df_members_cluster[event_column], data_test[event_column]
+
+    # Fitting the model
+    csf = RandomSurvivalForestModel(num_trees=200)
+    csf.fit(X_train, T_train, E_train, max_features='sqrt',
+        max_depth=5, min_node_size=20)
+```
 

### 2021-09-04 article/articleCustomerDropoutMembership.Rmd
upd

@@
 output:
+  bookdown::word_document2:
+    latex_engine: xelatex
+    number_sections: true
   bookdown::pdf_document2:
@@
 
-```{python createModel, eval=TRUE, echo=FALSE, cache = TRUE, warning = FALSE}
+```{python createModel, eval=TRUE, echo=FALSE, cache = FALSE, warning = FALSE}
 from pysurvival.models.survival_forest import RandomSurvivalForestModel
+from sklearn.model_selection import train_test_split
+from pysurvival.utils.metrics import concordance_index
+from pysurvival.utils.display import integrated_brier_score
+from pysurvival.utils.display import compare_to_actual
 
-# Number of samples in the dataset
-N = df_members.shape[0]
+X = df_members.copy()
+t = df_members['years_membership']
+e = df_members['dropout']
+X.drop(axis=1, columns=['years_membership','dropout'])
 
-# Building training and testing sets
-from sklearn.model_selection import train_test_split
-index_train, index_test = train_test_split( range(N), test_size = 0.4)
-data_train = df_members.loc[index_train].reset_index( drop = True )
-data_test  = df_members.loc[index_test].reset_index( drop = True )
-
-# Creating the X, T and E inputs
-X_train, X_test = df_members[features], data_test[features]
-T_train, T_test = df_members[time_column], data_test[time_column]
-E_train, E_test = df_members[event_column], data_test[event_column]
+X_train, X_test, t_train, t_test, e_train, e_test = train_test_split(X, t, e, test_size=0.3, random_state=0)
 
@@
 csf = RandomSurvivalForestModel(num_trees=200)
-csf.fit(X_train, T_train, E_train, max_features='sqrt',
-        max_depth=5, min_node_size=20)
+csf.fit(X_train, t_train, e_train, max_features='sqrt', max_depth=5, min_node_size=20)
+
+c_index = concordance_index(csf, X_test, t_test, e_test)
+ibs = integrated_brier_score(csf, X_test, t_test, e_test, t_max=12, figure_size=(12,5))
+results = compare_to_actual(csf, X_test, t_test, e_test, is_at_risk = False,  figure_size=(12, 6), metrics = ['rmse', 'mean', 'median'])
 ```
@@
 
-The model was built with with 60% of the data for training and 40% for testing.
+The model was built with with 70% of the data for training and 30% for testing.
 The survival model parameters where:
-
-```{python pysurvivalperformance, eval=TRUE, echo=FALSE, cache=FALSE, warning = FALSE}
-from pysurvival.utils.metrics import concordance_index
-from pysurvival.utils.display import integrated_brier_score
-from pysurvival.utils.display import compare_to_actual
-
-c_index = concordance_index(csf, X_test, T_test, E_test)
-
-ibs = integrated_brier_score(csf, X_test, T_test, E_test, t_max=12,
-    figure_size=(12,5))
-
-results = compare_to_actual(csf, X_test, T_test, E_test, is_at_risk = False,  
-figure_size=(12, 6), metrics = ['rmse', 'mean', 'median'])
-```
 
@@
 
-```{python elbowCalculation, eval=TRUE, echo=FALSE, cache=TRUE, warning = FALSE}
+```{python elbowCalculation, eval=TRUE, echo=FALSE, cache=FALSE, warning = FALSE}
 from sklearn.cluster import KMeans
@@
 
-TODO: fit the model for the five clusters... compare the performance with the model without
-clusters... corrigir acima para selecionar só as features
+TODO: Estava aqui... fit the model for the five clusters... compare the performance with the model without
+clusters... corrigir acima para selecionar só as features.. Explorart-SNE for better visualization...
 

### 2021-09-22 article/articleCustomerDropoutMembership.Rmd
upd

@@
 plt.ylabel('Average distortion')
-plt.show()
+plt.show();
 
@@
 ax.grid(True)
-ax.set_title('Clusters clientes')
+ax.set_title('Clusters clientes');
 ```
@@
     csf = RandomSurvivalForestModel(num_trees=200)
-    csf.fit(X_train, T_train, E_train, max_features='sqrt',
-        max_depth=5, min_node_size=20)
+    csf.fit(X_train, T_train, E_train, max_features='sqrt', max_depth=5, min_node_size=20)
 ```

### 2021-09-22 article/articleCustomerDropoutMembership.Rmd
upd

@@
 results = compare_to_actual(csf, X_test, t_test, e_test, is_at_risk = False,  figure_size=(12, 6), metrics = ['rmse', 'mean', 'median'])
+
+print(results)
 ```
@@
     csf.fit(X_train, T_train, E_train, max_features='sqrt', max_depth=5, min_node_size=20)
+    c_index = concordance_index(csf, X_test, t_test, e_test)
+    ibs = integrated_brier_score(csf, X_test, t_test, e_test, t_max=12, figure_size=(12,5))
+    results_clusters = compare_to_actual(csf, X_test, t_test, e_test, is_at_risk = False,  figure_size=(12, 6), 
+                                metrics = ['rmse', 'mean', 'median'])
+    print(results_clusters)
 ```

### 2021-09-29 article/articleCustomerDropoutMembership.Rmd
upd

@@
 TODO: Estava aqui... fit the model for the five clusters... compare the performance with the model without
-clusters... corrigir acima para selecionar só as features.. Explorart-SNE for better visualization...
+clusters... corrigir acima para selecionar só as features.. Explorar t-SNE for better visualization...
 
-```{python createModelClusters, eval=FALSE, echo=FALSE, cache = TRUE, warning = FALSE}
+```{python createModelClusters, eval=TRUE, echo=FALSE, cache = TRUE, warning = FALSE}
 # Building training and testing sets for the clusters
@@
     df_members_cluster = df_members[df_members.cluster == cluster]
-    N = df_members_cluster.shape[0]
-    index_train, index_test = train_test_split(range(N), test_size = 0.4)
-    data_train = df_members_cluster[features].loc[index_train].reset_index( drop = True )
-    data_test  = df_members_cluster[features].loc[index_test].reset_index( drop = True )
-
-    # Creating the X, T and E inputs
-    X_train, X_test = df_members_cluster[features], data_test[features]
-    T_train, T_test = df_members_cluster[time_column], data_test[time_column]
-    E_train, E_test = df_members_cluster[event_column], data_test[event_column]
-
+    X = df_members_cluster.copy()
+    t = df_members_cluster['years_membership']
+    e = df_members_cluster['dropout']
+    X.drop(axis=1, columns=['years_membership','dropout'])
+    X_train, X_test, t_train, t_test, e_train, e_test = train_test_split(X, t, e, random_state=0)
+    print(f"The cluster {cluster} as a size of {X.shape[0]}")
     # Fitting the model
     csf = RandomSurvivalForestModel(num_trees=200)
-    csf.fit(X_train, T_train, E_train, max_features='sqrt', max_depth=5, min_node_size=20)
+    csf.fit(X_train, t_train, e_train, max_features='sqrt',
+        max_depth=5, min_node_size=20)
+
     c_index = concordance_index(csf, X_test, t_test, e_test)
+
     ibs = integrated_brier_score(csf, X_test, t_test, e_test, t_max=12, figure_size=(12,5))
-    results_clusters = compare_to_actual(csf, X_test, t_test, e_test, is_at_risk = False,  figure_size=(12, 6), 
-                                metrics = ['rmse', 'mean', 'median'])
-    print(results_clusters)
+
+    results = compare_to_actual(csf, X_test, t_test, e_test, is_at_risk = False,  
+                                figure_size=(12, 6), metrics = ['rmse', 'mean', 'median'])
+    print(results)
+
 ```

### 2021-09-29 article/articleCustomerDropoutMembership.Rmd
up

@@
 output:
-  bookdown::word_document2:
-    latex_engine: xelatex
-    number_sections: true
   bookdown::pdf_document2:
@@
     keep_tex: true
+    latex_engine: xelatex
+    number_sections: true
+  bookdown::word_document2:
     latex_engine: xelatex
@@
 
-```{python createModel, eval=TRUE, echo=FALSE, cache = FALSE, warning = FALSE}
+```{python createModel, eval=FALSE, echo=FALSE, cache = FALSE, warning = FALSE}
 from pysurvival.models.survival_forest import RandomSurvivalForestModel
@@
 df_members['cluster']=cluster.predict(df_members)
-print(df_members.cluster.value_counts())
+print(df_members.cluster.value_counts());
 ```
@@
     y = df_members['Y_pca'].loc[df_members.cluster == cluster]
-    ax.scatter(x, y, label=cluster, alpha=1, edgecolors='none')
+    ax.scatter(x, y, label=cluster, alpha=1, edgecolors='none');
 
@@
                                 figure_size=(12, 6), metrics = ['rmse', 'mean', 'median'])
-    print(results)
-
+    print(f"Cluster {cluster} RMSE {results['root_mean_squared_error']}")
+    print(f"Cluster {cluster} MAE {results['median_absolute_error']}")
+    print(f"Cluster {cluster} MAError {results['mean_absolute_error']}")
 ```
@@
 hyper-parameters Grid Search technique. The explored hyper-parameters and the
-best values of these parameters for every model are listed in (ref:table1).
-
-(ref:table1) Table 2
+best values of these parameters for every model are listed in (Table 3).
 
@@
 -->
+```
+
+The results of the performance of the models are available in table 4.
+Colocar o modelo. Resultados do modelo
+```{=latex}
+\begin{table}[ht]
+    \footnotesize
+    \centering
+    \begin{tabular}{p{0.15\textwidth}p{0.55\textwidth}p{0.25\textwidth}}
+    \hline
+    \textbf{Model name} & \textbf{Results} \\
+    \hline
+    Survival trees  & RMSE 2.77 Median Absolute Error 0.0 \\
+    Survival trees
+    with clusters   & Cluster 1: RMSE  2.775 Median Absolute Error  0.000 
+                      Cluster 3: RMSE  8.769 Median Absolute Error  1.276
+                      Cluster 2: RMSE 14.883 Median Absolute Error  2.030
+                      Cluster 4: RMSE  9.624 Median Absolute Error  4.748
+                      Cluster 0: RMSE 77.236 Median Absolute Error 26.066 \\
+    Scikit survival 
+    trees           & scikit survival                           \\
+    Scikit survival 
+    with clusters   & scikit with clusters                      \\
+    Scikit survival 
+    gradient boost  & scikit survival gradient boost           \\
+    Scikit survival 
+    gradient boost 
+    with clusters   & scikit with clusters                      \\
+    \hline
+    \end{tabular}
+    \caption{Hyper-parameters best values}
+    \label{hyperparametersbestvalues}
+\end{table}
 ```

### 2021-10-24 article/articleCustomerDropoutMembership.Rmd
upd

@@
 
-``` {.markdown}
+``` markdown
 If an organization can predict a possible dropout and
@@
 Other studies proposed also the integration of several algorithms to improve the
-performance in the prediction of the dropout such the usage of clusters
-combined with churn prediction [@hung_applying_2006; @gok_case_2015;
+performance in the prediction of the dropout such the usage of clusters combined
+with churn prediction [@hung_applying_2006; @gok_case_2015;
 @vijaya_sivasankar_2019]. The approach relies in the assumption that combining
@@
 considering the dynamic behavior of the customer in the intent to drop out
-[@alboukaey_dynamic_2020]. The importance of understanding when dropout will
+\[@alboukaey_dynamic_2020\]. The importance of understanding when dropout will
 occur and the risk when discarding the temporal perspective of the problem seems
 to be an element that should be addressed. Few studies considered this
-[@perianez_churn_2016; @burez_separating_2008]. This shows an opportunity to
+\[@perianez_churn_2016; @burez_separating_2008\]. This shows an opportunity to
 address the importance of the timeframe and its influence on the efficiency of
@@
 
-```{python createModel, eval=FALSE, echo=FALSE, cache = FALSE, warning = FALSE}
+```{python createModel, eval=TRUE, echo=FALSE, cache = FALSE, warning = FALSE}
 from pysurvival.models.survival_forest import RandomSurvivalForestModel
@@
 # Fitting the model
-csf = RandomSurvivalForestModel(num_trees=200)
-csf.fit(X_train, t_train, e_train, max_features='sqrt', max_depth=5, min_node_size=20)
+csf = RandomSurvivalForestModel(num_trees=20)
+csf.fit(X_train, t_train, e_train, max_features='sqrt', max_depth=5, min_node_size=20, seed = 1)
 
@@
 
-The calculation of he number of clusters used the package mclust
-[@scrucca2016] using the Bayesian Information Criterion (BIC). The model that
-gives the minimum BIC score can be selected as the best model [@schwarz1978]
-simplifying the problem related to choosing the number of components and
-identifying the structure of the covariance matrix, based on modelling with
-multivariate normal distributions for each component that forms the data set
-[@akogul2016].
+The calculation of he number of clusters used the package mclust [@scrucca2016]
+using the Bayesian Information Criterion (BIC). The model that gives the minimum
+BIC score can be selected as the best model [@schwarz1978] simplifying the
+problem related to choosing the number of components and identifying the
+structure of the covariance matrix, based on modelling with multivariate normal
+distributions for each component that forms the data set [@akogul2016].
 
@@
 
-Estou a ter problemas com o cálculo dos clusters com o BIC... 
-estava a aqui a confirmar os clusters, talvez seja melhor reduzir as variáveis...
-testar a abordagem aos clusters no artigo dos vinhos...
+Estou a ter problemas com o cálculo dos clusters com o BIC... estava a aqui a
+confirmar os clusters, talvez seja melhor reduzir as variáveis... testar a
+abordagem aos clusters no artigo dos vinhos...
 
@@
 
-TODO: Estava aqui... fit the model for the five clusters... compare the performance with the model without
-clusters... corrigir acima para selecionar só as features.. Explorar t-SNE for better visualization...
+TODO: Estava aqui... fit the model for the five clusters... compare the
+performance with the model without clusters... corrigir acima para selecionar só
+as features.. Explorar t-SNE for better visualization...
 
@@
     # Fitting the model
-    csf = RandomSurvivalForestModel(num_trees=200)
+    csf = RandomSurvivalForestModel(num_trees=20)
     csf.fit(X_train, t_train, e_train, max_features='sqrt',
-        max_depth=5, min_node_size=20)
+        max_depth=5, min_node_size=20, seed = 1)
 
@@
 performance of organizations in retaining customers, which is a management
-problem in which data mining is adopted [@verbeke_new_2012]. The goals of the
+problem in which data mining is adopted \[@verbeke_new_2012\]. The goals of the
 model should be formulated considering the context of the problem that is being
@@
 development of proactive actions to minimize the investment in retention
-strategies [@coussement_churn_2008]. Some assumptions that underlie the adoption
-of uplift metrics consider that customers with a higher risk of churning could
-not be the best targets, as suggested by @ascarza_retention_2018. Other
-researchers addressed the problem using the top-decile lift to develop more
-proactive actions to retain the customers at risk of churning
-[@coussement_churn_2008;xie_customer_2009]. This approach considers the 10% of
+strategies \[@coussement_churn_2008\]. Some assumptions that underlie the
+adoption of uplift metrics consider that customers with a higher risk of
+churning could not be the best targets, as suggested by @ascarza_retention_2018.
+Other researchers addressed the problem using the top-decile lift to develop
+more proactive actions to retain the customers at risk of churning
+\[@coussement_churn_2008;xie_customer_2009\]. This approach considers the 10% of
 customers with more risk, and investments in retention strategies should be
@@
 ```
-```{=html}
-<!--
-|        Model name        | Explored hyper-parameters' values                    | Best parameters |
-| ------------------------ | ---------------------------------------------------- | --------------- |
-| Survival trees           |                                                      |                 |   
-| without cluster          |                                                      |                 |  
--->
-```
+The results of the performance of the models are available in table 4. Colocar o
+modelo. Resultados do modelo
 
-The results of the performance of the models are available in table 4.
-Colocar o modelo. Resultados do modelo
 ```{=latex}
@@
     \centering
-    \begin{tabular}{p{0.15\textwidth}p{0.55\textwidth}p{0.25\textwidth}}
+    \begin{tabular}{p{0.15\textwidth}p{0.35\textwidth}r p{0.10\textwidth}}
     \hline
-    \textbf{Model name} & \textbf{Results} \\
+    \textbf{Model name} & \textbf{Results} & \textbf{n}\\
     \hline
-    Survival trees  & RMSE 2.77 Median Absolute Error 0.0 \\
-    Survival trees
-    with clusters   & Cluster 1: RMSE  2.775 Median Absolute Error  0.000 
-                      Cluster 3: RMSE  8.769 Median Absolute Error  1.276
-                      Cluster 2: RMSE 14.883 Median Absolute Error  2.030
-                      Cluster 4: RMSE  9.624 Median Absolute Error  4.748
-                      Cluster 0: RMSE 77.236 Median Absolute Error 26.066 \\
+    Survival trees  & RMSE 57.815                                     & 25316\\
+                    & MAE 18.966                                      & \\   
+                    & MEAE 38.557                                     & \\
+    Survival trees  & Cluster 0: RMSE  2.785 MAE  0.000 MEAE  0.899   &   930\\
+    with clusters   & Cluster 1: RMSE 78.141 MAE 28.537 MEAE 56.989   & 17067\\
+                    & Cluster 2: RMSE 15.081 MAE  1.995 MEAE  6.850   &  2423\\
+                    & Cluster 3: RMSE  9.354 MAE  2.506 MEAE  6.427   &  2816\\
+                    & Cluster 4: RMSE  8.805 MAE  1.349 MEAE  3.725   &  2080\\
+
     Scikit survival 
-    trees           & scikit survival                           \\
+    trees           & scikit survival                                 & \\
     Scikit survival 
-    with clusters   & scikit with clusters                      \\
+    with clusters   & scikit with clusters                            & \\
     Scikit survival 
-    gradient boost  & scikit survival gradient boost           \\
+    gradient boost  & scikit survival gradient boost                  & \\
     Scikit survival 
     gradient boost 
-    with clusters   & scikit with clusters                      \\
+    with clusters   & scikit with clusters                            & \\
     \hline
@@
     \label{hyperparametersbestvalues}
+    * Root Mean Square Error (RMSE), Mean Absolute Error (MAE), Median Absolute Error (MEAE)  
 \end{table}
@@
 
-Example of Developed actions:
+Example of Developed actions place in the discussion:
 
-``` {.markdown}
+``` markdown
 Each month, the company identified the customers who were up for renewal and

### 2021-10-24 article/articleCustomerDropoutMembership.tex
upd

@@
 \vspace{1cm}Customer dropout membership\footnote{Corresponding address: \href{mailto:sobreiro@esdrm.ipsantarem.pt}{\nolinkurl{sobreiro@esdrm.ipsantarem.pt}}.
-The current template adapts part of the Rmd code by \href{https://github.com/paulcbauer/Writing_a_reproducable_paper_in_rmarkdown}{Paul C.
-Bauer}, Mannheim Centre for European Social Research.}\vspace{0.5cm}}
+Quality of Life Research Centre, Polytechnic Institute of Santarém, Portugal}\vspace{0.5cm}}
 \author{Pedro Sobreiro, Javier Berrocal, Domingos Martinho, José Garcia Alonso\\
-~\\
+\strut \\
 Extremadura University\\}
@@
 \hfill\break
-25 julho, 2021\\
-~\\}
+9 outubro, 2021\\
+\strut \\}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-\end{itemize}
-
 Context: An organization membership located in Portugal. The organization offers
@@
 
-\clearpage
+Churn dropout prediction is a problem being addressed supported in the idea that
+the customers database is the most valuable asset that the organizations possess
+(\protect\hyperlink{ref-Athanassopoulos_2000}{Athanassopoulos 2000}), which requires determining customers that will attrite
+(\protect\hyperlink{ref-alboukaey_dynamic_2020}{Alboukaey, Joukhadar, and Ghneim 2020a}). Dropout implies in contractual business that the
+customer needs to renew their contracts to continue its usage
+(\protect\hyperlink{ref-Ascarza_Hardie_2013}{Ascarza and Hardie 2013}).
 
-\renewcommand{\baselinestretch}{0.5}\normalsize
+However, in contractual settings the customer dropout represents an explicit
+ending of a relationship more punitive than non contractual settings
+(\protect\hyperlink{ref-risselada_staying_2010}{Risselada, Verhoef, and Bijmolt 2010}) This has implications to the profitability of the
+organizations increasing marketing costs and reducing sales
+(\protect\hyperlink{ref-amin_customer_2017}{Amin et al. 2017}).
 
-\renewcommand{\baselinestretch}{1.1}\normalsize
-
-\clearpage
+The anticipation of the dropout allows the development of countermeasures to
+reduce customer churn. Several studies address the problem related to customer
+retention trying to improve the profitability (\protect\hyperlink{ref-coussement_improving_2009}{Coussement and Van den Poel 2009}; \protect\hyperlink{ref-devriendt_why_2019}{Devriendt, Berrevoets, and Verbeke 2019}; \protect\hyperlink{ref-garcia_intelligent_2017}{García, Nebot, and Vellido 2017})
 
@@
 \begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(dplyr)}
-\FunctionTok{library}\NormalTok{(dlookr)}
-\FunctionTok{library}\NormalTok{(ggplot2)}
-
-\CommentTok{\#eda\_report(nlswork,output\_dir  = }
-\CommentTok{\#   "C:/Users/mangelo.EEG/Documents/GitHub/prjs/reports/",}
-\CommentTok{\#   output\_file = "eda\_report.pdf")}
-
-\DocumentationTok{\#\# The data}
-
-\FunctionTok{names}\NormalTok{(airquality)}
+\NormalTok{If an organization can predict a possible dropout and}
+\NormalTok{develop countermeasures to avoid desertion, they can avoid customer defections that }
+\NormalTok{lead to a loss of money. Reichheld (1996) evidenced that reducing dropout rates by 5\% }
+\NormalTok{(e.g., from 15\% to 10\% per year) could represent an increase in prots up to double, as }
+\NormalTok{acquiring new customers costs 5 to 6 times more than retaining existing ones (Bhattacharya, }
+\NormalTok{1998). Existing organizations are addressing this problem by shifting their target from }
+\NormalTok{capturing new customers to preserving existing ones (García et al., 2017),}
+\NormalTok{as investments in retention strategies have higher returns than acquisitions (Coussement }
+\NormalTok{\& Van den Poel, 2009). The importance of customer retention to maintain organizational protability (Devriendt}
+\NormalTok{et al., 2019) leads to the problem of how to quantify the nancial impact of customer retention actions}
+\NormalTok{under the assumption that the organization goal should be related to the increase the lifetime of the}
+\NormalTok{customer to increase their prots. The customer lifetime value (CLV) allows us to measure this, as itr}
 \end{Highlighting}
@@
 
-\begin{verbatim}
-## [1] "Ozone"   "Solar.R" "Wind"    "Temp"    "Month"   "Day"
-\end{verbatim}
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Address the global problem of customer dropout
+\item
+  The identification of approaches to predict dropout requires more than only,
+  address the prediction accuracy such as \ldots{} place existing studies
+  addressing this\ldots{}
+\item
+  A lot of effort has been placed testing the accuracy of existing algorithms,
+  in this study we try to fill this gap and explore also a balancing between
+  the model interpretability, accuracy, and the investment required.
+\item
+  Dropout prediction problem related to the timings
+\end{enumerate}
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#summary(nlswork)}
+The approaches normally employing use a dependent variable representing dropout
+or non-dropout, without considering a dynamic perspective that the dropout risk
+changes over time (\protect\hyperlink{ref-Alboukaey_dynamic_2020}{Alboukaey, Joukhadar, and Ghneim 2020b}). The survival models try to solve
+this limitation (\protect\hyperlink{ref-routh_estimating_2020}{Routh, Roy, and Meyer 2020}) capturing a temporal dimension of the
+customer dropout (\protect\hyperlink{ref-perianez_churn_2016}{Perianez et al.
+2016}).
+\protect\hyperlink{ref-perianez_churn_2016}{Perianez et al.} (\protect\hyperlink{ref-perianez_churn_2016}{2016}) used survival
+analysis to predict also when the dropout will occur.
 
-\DocumentationTok{\#\# Missing values}
-\FunctionTok{library}\NormalTok{(visdat)}
-\FunctionTok{vis\_dat}\NormalTok{(airquality)}
-\end{Highlighting}
-\end{Shaded}
+Other studies proposed also the integration of several algorithms to improve the
+performance in the prediction of the dropout such the usage of clusters combined
+with churn prediction (\protect\hyperlink{ref-gok_case_2015}{Gök, Özyer, and Jida 2015}; \protect\hyperlink{ref-hung_applying_2006}{Hung, Yen, and Wang 2006}; \protect\hyperlink{ref-vijaya_sivasankar_2019}{Vijaya and Sivasankar 2019}).
+The approach relies in the assumption that combining
+the customers in different clusters allows the improvement of the prediction
+accuracy. \protect\hyperlink{ref-vijaya_sivasankar_2019}{Vijaya and Sivasankar} (\protect\hyperlink{ref-vijaya_sivasankar_2019}{2019}) suggested the adoption hybrid models combining
+more than one classier are achieving increased performance compared to those
+using single classifiers.
 
-\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-1.pdf}
+There are several challenges around the timing related to dropout, or
+considering the dynamic behavior of the customer in the intent to drop out
+\[@alboukaey_dynamic_2020\]. The importance of understanding when dropout will
+occur and the risk when discarding the temporal perspective of the problem seems
+to be an element that should be addressed. Few studies considered this
+\[@perianez_churn_2016; @burez_separating_2008\]. This shows an opportunity to
+address the importance of the timeframe and its influence on the efficiency of
+the model and also evalute if the combination of clusters could improve the
+performance.
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(naniar)}
-\FunctionTok{vis\_miss}\NormalTok{(airquality)}
-\end{Highlighting}
-\end{Shaded}
+In this study, we adopt random survival forests which have never been used in
+understanding factors affecting membership in a sport club using existing data
+in a Sport Club. The analysis is based on the use of random survival forests in
+the presence of covariates that do not necessarily satisfy the PH assumption.
+Additionly we also propose a new approach combining clusters with survival
+analysis.
 
-\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-2.pdf}
+??? Add interpretability layer
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{gg\_miss\_upset}\NormalTok{(airquality)}
-\end{Highlighting}
-\end{Shaded}
+Random Survival Forests does not make the proportional hazards assumption
+(\protect\hyperlink{ref-Ehrlinger_2016}{Ehrlinger 2016}) and has the flexibility to model survivor curves that are of
+dissimilar shapes for contrasting groups of subjects. Random Survival Forest is
+an extension of Random Forest allowing efficient non-parametric analysis of time
+to event data (\protect\hyperlink{ref-Breiman_2001}{Breiman 2001}). This characteristics allow us to surpass the Cox
+Regression limitation of the proportional hazard assumption, requiring to
+exclude variables which not fulfill the model assumption. It was shown by
+\protect\hyperlink{ref-Breiman_2001}{Breiman} (\protect\hyperlink{ref-Breiman_2001}{2001}) that ensemble learning can be further improved by injecting
+randomization into the base learning process - a method called Random Forests.
 
-\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-3.pdf}
+\hypertarget{methodology}{%
+\section{Methodology}\label{methodology}}
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\DocumentationTok{\#\# GRAPHS}
-\NormalTok{dplyr}\SpecialCharTok{::}\FunctionTok{glimpse}\NormalTok{(cars}\SpecialCharTok{$}\NormalTok{Ozone)}
-\end{Highlighting}
-\end{Shaded}
+Dropout is a binary value where one represent churn and zero not churn. The
+dropout happens when a member does not have a payment \ldots{}
 
-\begin{verbatim}
-##  NULL
-\end{verbatim}
+The model performance was determined with the concordance probability (C-index),
+Brier Score (BS) and Mean Absolute Error (MAE) (\protect\hyperlink{ref-wangmachine2017}{Wang, Li, and Reddy 2017}). The feature
+importance was determined calculating the difference between the true class
+label and noised data (\protect\hyperlink{ref-Breiman_2001}{Breiman 2001}).
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{d }\OtherTok{\textless{}{-}} \FunctionTok{density}\NormalTok{(airquality}\SpecialCharTok{$}\NormalTok{Temp)}
-\FunctionTok{plot}\NormalTok{(d)}
-\end{Highlighting}
-\end{Shaded}
+\hypertarget{dataset}{%
+\subsection{Dataset}\label{dataset}}
 
-\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-4.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{plot}\NormalTok{(airquality}\SpecialCharTok{$}\NormalTok{Ozone, airquality}\SpecialCharTok{$}\NormalTok{Solar.R)}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-5.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{ggplot}\NormalTok{(airquality, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Ozone, }\AttributeTok{y =}\NormalTok{ Solar.R)) }\SpecialCharTok{+}
-\FunctionTok{geom\_miss\_point}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-6.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{ggplot}\NormalTok{(airquality, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Wind, }\AttributeTok{y =}\NormalTok{ Temp)) }\SpecialCharTok{+}
-\FunctionTok{geom\_miss\_point}\NormalTok{() }\SpecialCharTok{+}
-\FunctionTok{facet\_wrap}\NormalTok{(}\FunctionTok{vars}\NormalTok{(Month))}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{articleCustomerDropoutMembership_files/figure-latex/Stats1-7.pdf}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{stats }\OtherTok{\textless{}{-}} \FunctionTok{summary}\NormalTok{(airquality}\SpecialCharTok{$}\NormalTok{Temp)}
-\NormalTok{stats}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##    Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-##   56.00   72.00   79.00   77.88   85.00   97.00
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{describe}\NormalTok{(airquality)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## # A tibble: 6 x 26
-##   variable     n    na   mean    sd se_mean   IQR skewness kurtosis   p00   p01
-##   <chr>    <int> <int>  <dbl> <dbl>   <dbl> <dbl>    <dbl>    <dbl> <dbl> <dbl>
-## 1 Ozone      116    37  42.1  33.0    3.06   45.2  1.24       1.29    1    4.3 
-## 2 Solar.R    146     7 186.   90.1    7.45  143   -0.428     -0.968   7   10.2 
-## 3 Wind       153     0   9.96  3.52   0.285   4.1  0.348      0.111   1.7  2.56
-## 4 Temp       153     0  77.9   9.47   0.765  13   -0.378     -0.404  56   57   
-## 5 Month      153     0   6.99  1.42   0.115   2   -0.00239   -1.30    5    5   
-## 6 Day        153     0  15.8   8.86   0.717  15    0.00265   -1.20    1    1   
-## # ... with 15 more variables: p05 <dbl>, p10 <dbl>, p20 <dbl>, p25 <dbl>,
-## #   p30 <dbl>, p40 <dbl>, p50 <dbl>, p60 <dbl>, p70 <dbl>, p75 <dbl>,
-## #   p80 <dbl>, p90 <dbl>, p95 <dbl>, p99 <dbl>, p100 <dbl>
-\end{verbatim}
-
-\hypertarget{experimental-results}{%
-\section{Experimental Results}\label{experimental-results}}
-
-\hypertarget{data-description}{%
-\subsection{Data description}\label{data-description}}
-
-\begin{verbatim}
-##  [1] "Sócio"               "dataAdesao"          "ano"                
-##  [4] "dataNascimento"      "idade"               "sexo"               
-##  [7] "estadoCivil"         "categoria"           "quotaMensal"        
-## [10] "profissao"           "codPostal"           "ultimaQuota"        
-## [13] "ultimoPagamento"     "valorTotal"          "totalJogos"         
-## [16] "jogosEpoca"          "diasUltimoPagamento" "mesesUP"            
-## [19] "abandonou"           "anosSocio"           "idaEstadio"         
-## [22] "escaloesTotalJogos"  "mes"
-\end{verbatim}
-
-\begin{verbatim}
-##  [1] "num_socio"                 "dt_inscription"           
-##  [3] "year"                      "birth_date"               
-##  [5] "age"                       "sex"                      
-##  [7] "marital_status"            "category"                 
-##  [9] "monthly_fee"               "occupation"               
-## [11] "zip_code"                  "dt_last_invoice"          
-## [13] "dt_last_payment"           "total_amount"             
-## [15] "total_matches"             "season_matches"           
-## [17] "days_since_last_payment"   "months_since_last_payment"
-## [19] "dropout"                   "years_membership"         
-## [21] "stadium_access"            "quart_stadium_entries"    
-## [23] "inscription_month"
-\end{verbatim}
-
-\begin{verbatim}
-## tibble [25,316 x 14] (S3: tbl_df/tbl/data.frame)
-##  $ year                     : num [1:25316] 1944 1944 1945 1945 1945 ...
-##  $ age                      : num [1:25316] 83 88 73 97 97 91 88 95 88 78 ...
-##  $ sex                      : chr [1:25316] "M" "M" "M" "M" ...
-##  $ marital_status           : chr [1:25316] "casado" "solteiro" "nao definido" "casado" ...
-##  $ monthly_fee              : num [1:25316] 10 10 10 5 10 5 5 5 10 10 ...
-##  $ total_amount             : num [1:25316] 1906 1906 1553 790 1466 ...
-##  $ total_matches            : num [1:25316] 0 0 0 0 0 20 74 0 154 0 ...
-##  $ season_matches           : num [1:25316] 0 0 0 0 0 0 0 0 6 0 ...
-##  $ months_since_last_payment: num [1:25316] 3 3 36 8 35 4 41 40 4 2 ...
-##  $ dropout                  : num [1:25316] 0 0 1 0 1 0 1 1 0 0 ...
-##  $ years_membership         : num [1:25316] 74 74 73 73 73 73 73 73 73 72 ...
-##  $ stadium_access           : num [1:25316] 0 0 0 0 0 1 1 0 1 0 ...
-##  $ quart_stadium_entries    : chr [1:25316] "ate 1" "ate 1" "ate 1" "ate 1" ...
-##  $ inscription_month        : num [1:25316] 10 10 8 9 9 12 1 1 2 4 ...
-\end{verbatim}
-
-\includegraphics{articleCustomerDropoutMembership_files/figure-latex/get_data-1.pdf}
-
-Table \ref{tab:tab1cars} shows data's summary statistics.\footnote{You can reference the table as \ref{tab1cars}.} \texttt{stargazer()} is
-and excellent solution to export outputs.
-
-Teste \ref{tab:summarytable}
+Table \ref{tab:summarytable} shows data's summary statistics. The average age
+is 27.3 ±
+20.1, the members have an attendance of
+27 ±
+45.8 with a membership of
+11 ±
+10.9 years.
 
@@
 
-\caption{\label{tab:summarytable}Summary statistics}
+\caption{\label{tab:summarytable}Summary statistics of features used}
 \centering
@@
 \midrule
-year & 2,007 (11)\\
-age & 27 (20)\\
-sex & \\
+Age in years, Mean (SD) & 27 (20)\\
+Male or female, \% & \\
 \hspace{1em}F & 32\%\\
 \hspace{1em}M & 68\%\\
+Single, married and other., \% & \\
 \addlinespace
-marital\_status & \\
 \hspace{1em}casado & 20\%\\
@@
 \hspace{1em}solteiro & 48\%\\
+monthly\_fee, \% & \\
 \addlinespace
-monthly\_fee & \\
 \hspace{1em}0 & <0.1\%\\
@@
 \hspace{1em}5 & 3.4\%\\
+\hspace{1em}6 & 12\%\\
 \addlinespace
-\hspace{1em}6 & 12\%\\
 \hspace{1em}10 & 24\%\\
-total\_amount & 316 (494)\\
-total\_matches & 27 (46)\\
-season\_matches & 2.2 (4.1)\\
+total\_amount, Mean (SD) & 316 (494)\\
+total\_matches, Mean (SD) & 27 (46)\\
+season\_matches, Mean (SD) & 2.2 (4.1)\\
+months\_since\_last\_payment, Mean (SD) & 19 (32)\\
 \addlinespace
-months\_since\_last\_payment & 19 (32)\\
-dropout & 22\%\\
-years\_membership & 11 (11)\\
-stadium\_access & 40\%\\
-quart\_stadium\_entries & \\
+dropout, \% & 22\%\\
+years\_membership, Mean (SD) & 11 (11)\\
+stadium\_access, \% & 40\%\\
+quart\_stadium\_entries, \% & \\
+\hspace{1em}1 a 21 & 10\%\\
 \addlinespace
-\hspace{1em}1 a 21 & 10\%\\
 \hspace{1em}21 a 56 & 9.8\%\\
@@
 \hspace{1em}mais 105 & 10.0\%\\
-\addlinespace
-inscription\_month & 6.9 (3.4)\\
+inscription\_month, Mean (SD) & 6.9 (3.4)\\
 \bottomrule
-\multicolumn{2}{l}{\rule{0pt}{1em}\textsuperscript{1} Mean (SD); \%}\\
 \end{tabular}
@@
 
+Figure \ref{fig:membershipyear} shows the distribution of the dropout
+considering the number of years of membership.
+
+\begin{figure}
+\centering
+\includegraphics{articleCustomerDropoutMembership_files/figure-latex/membershipyear-1.pdf}
+\caption{\label{fig:membershipyear}Number of members by year}
+\end{figure}
+
+\hypertarget{model-construction}{%
+\subsection{Model construction}\label{model-construction}}
+
+Address the model construction\ldots{} the categorical variables \(sex\),
+\(marital\_status\) and \(quart\_stadium\_entries\) where converted to dummy
+variables.
+
+The random survival forest was developed using the package PySurvival
+(\protect\hyperlink{ref-Fotso_others_2019}{Fotso et al. 2019}). The most relevant variables predicting the dropout are
+analysed using the log-rank test. The metric variables are transformed to
+categorical using the quartiles to provide a statistical comparison of groups.
+The survival analysis was conducted using the package Lifelines
+(\protect\hyperlink{ref-Davidson-Pilon_2021}{Davidson-Pilon 2021}).
+
+PySurvival is an open source python package for Survival Analysis modeling - the
+modeling concept used to analyze or predict when an event is likely to happen.
+It is built upon the most commonly used machine learning packages such NumPy,
+SciPy and PyTorch. PySurvival is compatible with Python 2.7-3.7
+
+\hypertarget{survival-trees-based-model}{%
+\subsubsection{Survival trees based model}\label{survival-trees-based-model}}
+
+In this model\ldots{} The survival trees based model uses pysurvival random forest
+
+Removed the variables with greater correlations \(total\_matches\) and
+\(quart\_stadium\_entries\)
+
+\includegraphics{articleCustomerDropoutMembership_files/figure-latex/removeVar-1.pdf}
+
+\begin{verbatim}
+##         age  ...  quart_stadium_entries_mais 105
+## 0      83.0  ...                               0
+## 1      88.0  ...                               0
+## 2      73.0  ...                               0
+## 3      97.0  ...                               0
+## 4      97.0  ...                               0
+## ...     ...  ...                             ...
+## 25311   7.0  ...                               0
+## 25312   8.0  ...                               0
+## 25313   2.0  ...                               0
+## 25314  14.0  ...                               0
+## 25315  28.0  ...                               0
+## 
+## [25316 rows x 14 columns]
+\end{verbatim}
+
+\begin{verbatim}
+## RandomSurvivalForestModel
+\end{verbatim}
+
+\includegraphics{articleCustomerDropoutMembership_files/figure-latex/createModel-3.pdf} \includegraphics{articleCustomerDropoutMembership_files/figure-latex/createModel-4.pdf}
+
+\begin{verbatim}
+## {'root_mean_squared_error': 57.814697481046416, 'median_absolute_error': 18.96621852166039, 'mean_absolute_error': 38.55719494742953}
+\end{verbatim}
+
+Table \ref{tab:summarytable2} shows variables importance.
+
+\begin{table}
+
+\caption{\label{tab:summarytable2}Summary statistics of features used}
+\centering
+\begin{tabular}[t]{lrr}
+\toprule
+feature & importance & pct\_importance\\
+\midrule
+dropout & 6.0669405 & 0.2528415\\
+months\_since\_last\_payment & 4.9581586 & 0.2066327\\
+season\_matches & 2.3851555 & 0.0994020\\
+total\_amount & 1.8950237 & 0.0789757\\
+quart\_stadium\_entries\_21 a 56 & 1.4643279 & 0.0610263\\
+\addlinespace
+quart\_stadium\_entries\_56 a 105 & 1.3455565 & 0.0560765\\
+stadium\_access & 1.2980882 & 0.0540982\\
+inscription\_month & 1.2790920 & 0.0533065\\
+marital\_status\_nao definido & 1.0339551 & 0.0430904\\
+sex\_M & 0.8881565 & 0.0370142\\
+\addlinespace
+age & 0.7472122 & 0.0311403\\
+monthly\_fee & 0.6333690 & 0.0263958\\
+marital\_status\_solteiro & -0.2702218 & 0.0000000\\
+marital\_status\_outro & -1.0259784 & 0.0000000\\
+quart\_stadium\_entries\_mais 105 & -1.2762422 & 0.0000000\\
+\addlinespace
+years\_membership & -4.1258616 & 0.0000000\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\hypertarget{model-building}{%
+\paragraph{Model building}\label{model-building}}
+
+The model was built with with 70\% of the data for training and 30\% for testing.
+The survival model parameters where:
+
+The model accuracy is very high in the first years. The prediction is very
+similar to the actual value. The absolute error mean of
+39 customers.
+
+\hypertarget{survival-trees-based-model-with-clusters}{%
+\subsubsection{Survival trees based model with clusters}\label{survival-trees-based-model-with-clusters}}
+
+Here we are will create clusters and developed the optimization within each
+cluster\ldots{}
+
+The calculation of he number of clusters used the package mclust (\protect\hyperlink{ref-scrucca2016}{Scrucca et al. 2016})
+using the Bayesian Information Criterion (BIC). The model that gives the minimum
+BIC score can be selected as the best model (\protect\hyperlink{ref-schwarz1978}{Schwarz 1978}) simplifying the
+problem related to choosing the number of components and identifying the
+structure of the covariance matrix, based on modelling with multivariate normal
+distributions for each component that forms the data set (\protect\hyperlink{ref-akogul2016}{Akogul and Erisoglu 2016}).
+
+In multivariate models are available the following approaches:
+
+\begin{itemize}
+\tightlist
+\item
+  ``EII''spherical, equal volume
+\item
+  ``EEE''ellipsoidal, equal volume, shape, and orientation
+\item
+  ``VII''spherical, unequal volume
+\item
+  ``VVV''ellipsoidal, varying volume, shape, and orientation, which is used as
+  default for initialization of EM algorithm
+\item
+  VVI'': diagonal, varying volume and shape t
+\end{itemize}
+
+Estou a ter problemas com o cálculo dos clusters com o BIC\ldots{} estava a aqui a
+confirmar os clusters, talvez seja melhor reduzir as variáveis\ldots{} testar a
+abordagem aos clusters no artigo dos vinhos\ldots{}
+
 \begin{Shaded}
 \begin{Highlighting}[]
-\CommentTok{\#from pysurvival.utils.display import correlation\_matrix}
-\ImportTok{import}\NormalTok{ pandas }\ImportTok{as}\NormalTok{ pd}
-\ImportTok{import}\NormalTok{ numpy }\ImportTok{as}\NormalTok{ np}
-
-\NormalTok{col }\OperatorTok{=}\NormalTok{ [}\StringTok{\textquotesingle{}sex\textquotesingle{}}\NormalTok{,}\StringTok{\textquotesingle{}marital\_status\textquotesingle{}}\NormalTok{,}\StringTok{\textquotesingle{}quart\_stadium\_entries\textquotesingle{}}\NormalTok{]}
-
-\NormalTok{df\_members }\OperatorTok{=}\NormalTok{ r.df\_members }\CommentTok{\#copy r dataframe to python}
-
-\NormalTok{df\_members }\OperatorTok{=}\NormalTok{ pd.get\_dummies(df\_members, columns}\OperatorTok{=}\NormalTok{col,drop\_first}\OperatorTok{=}\VariableTok{True}\NormalTok{)}
-
-\CommentTok{\# Creating the time and event columns}
-\NormalTok{time\_column }\OperatorTok{=} \StringTok{\textquotesingle{}years\_membership\textquotesingle{}}
-\NormalTok{event\_column }\OperatorTok{=} \StringTok{\textquotesingle{}dropout\textquotesingle{}}
-
-\CommentTok{\# Extracting the features}
-\NormalTok{features }\OperatorTok{=}\NormalTok{ np.setdiff1d(df\_members.columns, [time\_column, event\_column] ).tolist()}
-
-\CommentTok{\#correlation\_matrix(df\_members, figure\_size=(10,10), text\_fontsize=6)}
+\FunctionTok{library}\NormalTok{(NbClust)}
+\NormalTok{nb }\OtherTok{\textless{}{-}} \FunctionTok{NbClust}\NormalTok{(y, }\AttributeTok{diss=}\ConstantTok{NULL}\NormalTok{, }\AttributeTok{distance =} \StringTok{"euclidean"}\NormalTok{, }
+              \AttributeTok{min.nc=}\DecValTok{2}\NormalTok{, }\AttributeTok{max.nc=}\DecValTok{5}\NormalTok{, }\AttributeTok{method =} \StringTok{"kmeans"}\NormalTok{, }
+              \AttributeTok{index =} \StringTok{"all"}\NormalTok{, }\AttributeTok{alphaBeale =} \FloatTok{0.1}\NormalTok{)}
+\FunctionTok{hist}\NormalTok{(nb}\SpecialCharTok{$}\NormalTok{Best.nc[}\DecValTok{1}\NormalTok{,], }\AttributeTok{breaks =} \FunctionTok{max}\NormalTok{(}\FunctionTok{na.omit}\NormalTok{(nb}\SpecialCharTok{$}\NormalTok{Best.nc[}\DecValTok{1}\NormalTok{,])))}
 \end{Highlighting}
@@
 
-The average age in our data is 27.3.
+\begin{verbatim}
+## KMeans(n_clusters=1)
+## KMeans(n_clusters=2)
+## KMeans(n_clusters=3)
+## KMeans(n_clusters=4)
+## KMeans(n_clusters=5)
+## KMeans(n_clusters=6)
+## KMeans(n_clusters=7)
+## KMeans()
+## KMeans(n_clusters=9)
+## KMeans(n_clusters=10)
+## KMeans(n_clusters=11)
+## KMeans(n_clusters=12)
+## KMeans(n_clusters=13)
+## KMeans(n_clusters=14)
+## KMeans(n_clusters=15)
+## KMeans(n_clusters=16)
+## KMeans(n_clusters=17)
+## KMeans(n_clusters=18)
+## KMeans(n_clusters=19)
+\end{verbatim}
 
-\hypertarget{sec:tables}{%
-\section{Tables}\label{sec:tables}}
+\includegraphics{articleCustomerDropoutMembership_files/figure-latex/elbowCalculation-1.pdf}
 
-R Markdown PDF is now able to produce good tables with our output. For
-\texttt{stargazer} the label is contained in the function, while for \texttt{kable} it's
-contained in the chunk name.
+We are going to consider five clusters
 
-\hypertarget{stargazer-summary-and-regression-tables}{%
-\subsection{stargazer(): Summary and regression tables}\label{stargazer-summary-and-regression-tables}}
+\begin{verbatim}
+## KMeans(n_clusters=5)
+\end{verbatim}
 
-Table \ref{tab2} reports regression outputs. Name the models as you can refer
-to their names in the text (M1, M2, M3).
+\begin{verbatim}
+## 1    17068
+## 3     2818
+## 2     2420
+## 4     2080
+## 0      930
+## Name: cluster, dtype: int64
+\end{verbatim}
+
+TODO: Estava aqui\ldots{} fit the model for the five clusters\ldots{} compare the
+performance with the model without clusters\ldots{} corrigir acima para selecionar só
+as features.. Explorar t-SNE for better visualization\ldots{}
+
+\begin{verbatim}
+##        age  monthly_fee  total_amount  ...  cluster        X_pca     Y_pca
+## 0     83.0         10.0        1906.0  ...        0  1591.733817  8.137428
+## 1     88.0         10.0        1906.0  ...        0  1591.880351  8.064681
+## 9     78.0         10.0        1901.0  ...        0  1586.569739  7.062815
+## 11    71.0         10.0        1841.0  ...        0  1526.408306  7.138363
+## 16    85.0         10.0        1846.0  ...        0  1531.809369  5.877420
+## ...    ...          ...           ...  ...      ...          ...       ...
+## 8395  38.0         10.0        1566.0  ...        0  1249.711939  0.672902
+## 8448  76.0         10.0        1618.0  ...        0  1302.665956  9.866317
+## 8527  46.0         10.0        1594.0  ...        0  1277.941830  1.814926
+## 8820  35.0         10.0        1569.0  ...        0  1252.586502  1.828636
+## 9618  50.0         10.0        1743.0  ...        0  1426.869737  8.847541
+## 
+## [930 rows x 17 columns]
+## The cluster 0 as a size of 930
+## RandomSurvivalForestModel
+## Cluster 0 RMSE 2.7847832470895475
+## Cluster 0 MAE 0.0
+## Cluster 0 MAError 0.8993908970392324
+##         age  monthly_fee  total_amount  ...  cluster        X_pca      Y_pca
+## 2      73.0         10.0        1553.0  ...        4  1238.195691  36.216307
+## 4      97.0         10.0        1466.0  ...        4  1151.973917  33.646937
+## 8      88.0         10.0        1357.5  ...        4  1043.740627   1.126672
+## 19     75.0         10.0        1330.0  ...        4  1015.805346  -0.159127
+## 20     71.0         10.0        1202.0  ...        4   887.122902  43.078919
+## ...     ...          ...           ...  ...      ...          ...        ...
+## 13979  41.0         10.0        1055.0  ...        4   739.066765  -5.630017
+## 14539  52.0         10.0        1160.0  ...        4   844.316227  -3.458902
+## 14561  63.0         10.0        1054.0  ...        4   738.751223  -6.257051
+## 14590  72.0         10.0        1054.0  ...        4   739.011833  -6.363162
+## 14749  74.0         10.0        1176.0  ...        4   860.951124  -4.457038
+## 
+## [2080 rows x 17 columns]
+## The cluster 4 as a size of 2080
+## RandomSurvivalForestModel
+## Cluster 4 RMSE 8.804563964516955
+## Cluster 4 MAE 1.3491183838782987
+## Cluster 4 MAError 3.724668251115418
+##         age  monthly_fee  total_amount  ...  cluster       X_pca      Y_pca
+## 3      97.0          5.0         790.0  ...        2  476.781177  -2.878081
+## 5      91.0          5.0         805.0  ...        2  491.654217  -6.574167
+## 6      88.0          5.0         615.0  ...        2  301.161262  27.733634
+## 12     89.0          5.0         725.0  ...        2  411.433840   5.239050
+## 13     92.0          5.0         825.0  ...        2  511.634603  -7.411714
+## ...     ...          ...           ...  ...      ...         ...        ...
+## 18969  36.0         10.0         592.5  ...        2  276.710745 -13.295363
+## 18995  29.0         10.0         605.0  ...        2  288.882087  -5.939539
+## 19065  65.0         10.0         618.0  ...        2  303.058407 -13.402052
+## 19138  52.0         10.0         594.0  ...        2  278.671061 -13.549406
+## 19416  34.0         10.0         646.0  ...        2  330.084474 -11.496563
+## 
+## [2423 rows x 17 columns]
+## The cluster 2 as a size of 2423
+## RandomSurvivalForestModel
+## Cluster 2 RMSE 15.081809157987264
+## Cluster 2 MAE 1.9948683372618339
+## Cluster 2 MAError 6.859660105558099
+##         age  monthly_fee  total_amount  ...  cluster       X_pca      Y_pca
+## 7      95.0          5.0         340.0  ...        3   26.565346  22.744681
+## 10     95.0          5.0         560.0  ...        3  246.434943  22.798580
+## 18     86.0          5.0         580.0  ...        3  266.594554 -10.865739
+## 31     89.0          5.0         580.0  ...        3  266.622206 -10.015085
+## 43     83.0          5.0         530.0  ...        3  216.549443 -11.924744
+## ...     ...          ...           ...  ...      ...         ...        ...
+## 22880  55.0         10.0         215.0  ...        3 -100.027020 -17.220179
+## 22889  23.0         10.0         215.0  ...        3 -101.020945 -13.716401
+## 22979  39.0         10.0         215.0  ...        3 -100.531227 -16.812108
+## 23367  31.0         10.0         210.0  ...        3 -105.731921 -17.904987
+## 23594  60.0         10.0         225.0  ...        3  -89.927650 -17.890924
+## 
+## [2816 rows x 17 columns]
+## The cluster 3 as a size of 2816
+## RandomSurvivalForestModel
+## Cluster 3 RMSE 9.353790515157675
+## Cluster 3 MAE 2.5063338932739687
+## Cluster 3 MAError 6.426789920472172
+##         age  monthly_fee  total_amount  ...  cluster       X_pca       Y_pca
+## 1315   54.0         10.0           0.0  ...        1 -314.381393  -22.760221
+## 1607   58.0         10.0           5.0  ...        1 -311.253206  111.949308
+## 1722   50.0         10.0           0.0  ...        1 -314.586559  -22.906591
+## 1723   43.0         10.0           0.0  ...        1 -314.792159  -22.808044
+## 1724   47.0         10.0           0.0  ...        1 -314.674673  -22.864357
+## ...     ...          ...           ...  ...      ...         ...         ...
+## 25311   7.0          1.0          17.0  ...        1 -299.395699  -21.087144
+## 25312   8.0          1.0          12.0  ...        1 -304.362982  -21.171984
+## 25313   2.0          1.0          17.0  ...        1 -299.542555  -21.016754
+## 25314  14.0          1.0          17.0  ...        1 -299.190099  -21.185691
+## 25315  28.0         10.0           0.0  ...        1 -315.721037  -21.659490
+## 
+## [17067 rows x 17 columns]
+## The cluster 1 as a size of 17067
+## RandomSurvivalForestModel
+## Cluster 1 RMSE 78.14078277695158
+## Cluster 1 MAE 28.53685333800375
+## Cluster 1 MAError 56.988744683732826
+\end{verbatim}
+
+\includegraphics{articleCustomerDropoutMembership_files/figure-latex/createModelClusters-3.pdf} \includegraphics{articleCustomerDropoutMembership_files/figure-latex/createModelClusters-4.pdf} \includegraphics{articleCustomerDropoutMembership_files/figure-latex/createModelClusters-5.pdf} \includegraphics{articleCustomerDropoutMembership_files/figure-latex/createModelClusters-6.pdf} \includegraphics{articleCustomerDropoutMembership_files/figure-latex/createModelClusters-7.pdf} \includegraphics{articleCustomerDropoutMembership_files/figure-latex/createModelClusters-8.pdf} \includegraphics{articleCustomerDropoutMembership_files/figure-latex/createModelClusters-9.pdf} \includegraphics{articleCustomerDropoutMembership_files/figure-latex/createModelClusters-10.pdf} \includegraphics{articleCustomerDropoutMembership_files/figure-latex/createModelClusters-11.pdf} \includegraphics{articleCustomerDropoutMembership_files/figure-latex/createModelClusters-12.pdf}
+
+\hypertarget{open-questions---to-remove}{%
+\section{Open questions - to remove}\label{open-questions---to-remove}}
+
+\begin{itemize}
+    \item RQ1: What is the current state of the research being developed?
+    \item RQ2: What algorithms have been used to predict dropout?
+    \item RQ3: What are the features used to predict dropout?
+    \item RQ4: When does dropout occur?
+    \item RQ5: How is the accuracy of the machine learning algorithms in predicting dropout measured?
+\end{itemize}
+
+From RQ1, it was possible to identify some business areas that are
+under-researched, such as the energy sector, education, logistics and
+hospitality. Compared to other business areas such telecom or the financial
+sector, research on the energy sector or water supply is lacking, considering
+the contractual settings that are assumed to provide such types of services.
+Considering the business model of many software companies as software as a
+services (SaS), the number of research works is also surprisingly low.
+
+RQ2 also provided an overall perspective related to the algorithms being used to
+predict customer dropout. The first could be the importance and wider adoption
+of decision trees and random forests (\protect\hyperlink{ref-antipov_applying_2010}{Antipov and Pokryshevskaya 2010}; \protect\hyperlink{ref-benoit_improving_2012}{Benoit and Van den Poel 2012}; \protect\hyperlink{ref-burez_crm_2007}{Burez and Van den Poel 2007}), and logistic regression
+(\protect\hyperlink{ref-coussement_improved_2010}{Coussement, Benoit, and Van den Poel 2010}), which could be due to its higher interpretability
+and flexibility (\protect\hyperlink{ref-keramati_improved_2014}{Keramati et al. 2014}). Interpretability is an important
+aspect for the marketing department in the extraction of valuable information
+from the model to develop effective retention strategies (\protect\hyperlink{ref-verbeke_new_2012}{Verbeke et al. 2012}).
+The problem arises in the balancing between interpretability and the higher
+performance of the algorithms inspired by nature (such as neural networks). From
+a business perspective, dropout prediction should also be considered as a
+business objective, which requires more than predicting if the customer will
+churn or not (\protect\hyperlink{ref-devriendt_why_2019}{Devriendt et al. 2019}), where higher interpretability provides
+better support in the development of retention strategies. The developed SLR
+also raises the possibility of integrating different algorithms using ensemble
+methods or integrating several models using a hybrid approach. None of the
+studies integrated the survival approach to predict customer dropout, for
+example, using a hybrid approach.
+
+It is considered positive if actions are developed to retain customers, but the
+problems should also be considered, such as the following: (1) customers who
+have greater risk of dropout should be targeted to provide a base for a better
+ROI in the retention strategies (\protect\hyperlink{ref-coussement_churn_2008}{Coussement and Van den Poel 2008}; \protect\hyperlink{ref-xie_customer_2009}{Xie et al. 2009}) and
+(2) the retention strategies should be developed focused on customers with
+higher satisfaction, or its inclusion could be a reminder of the contractual
+agreement nearing an end and could lead to churn (\protect\hyperlink{ref-devriendt_why_2019}{Devriendt et al. 2019}).
+
+From RQ3, several types of features being used were able to be identified, such
+as demographic, behavioral, and economic indicators, pictorial data, network
+relationships or high cardinality features. The problem that arises is that some
+studies used data and features that were not described, and this creates a major
+issue, How can reproducibility be developed in a study without the availability
+of the data or the identification of the features used? Considering that science
+is driven by data, with the development of new technologies, the increasing
+complexity of research and the amount of data collected, the challenge is to
+ensure that research is available to all (\protect\hyperlink{ref-Hanson_Sugden_Alberts_2011}{Hanson, Sugden, and Alberts 2011}); this
+requires both availability of the data and the algorithms so that they can be
+explored by other researchers. The features are selected mainly to verify the
+performance of the models, and are essential to performance prediction,
+accuracy, and the steps for processing the data, which are fundamental to
+improve the model accuracy (\protect\hyperlink{ref-azeem_churn_2017}{Azeem, Usman, and Fong 2017}).
+
+There are several challenges around the timing related to dropout, or
+considering the dynamic behavior of the customer in the intent to drop out
+\[@alboukaey_dynamic_2020\]. The importance of understanding when dropout will
+occur and the risk when discarding the temporal perspective of the problem seems
+to be an element that should be addressed. Few studies considered this
+\[@perianez_churn_2016; @burez_separating_2008\]. This shows an opportunity to
+address the importance of the timeframe and its influence on the efficiency of
+the model.
+
+According to each business model, the timeframe could be addressed considering
+the survival probability according to the customer relationship age, and dropout
+predictions could be developed according to these survival probabilities, as
+suggested by \protect\hyperlink{ref-esteves_churn_2016}{Esteves and Mendes-Moreira} (\protect\hyperlink{ref-esteves_churn_2016}{2016}), to investigate which data timeframe produces
+the best result and how the efficiency of the models is influenced by this
+timeframe. Exploring the duration of the relation and the understanding of the
+features that increase or decrease that duration seems to be an important
+approach that could complement the existing approaches to predicting dropout.
+
+From RQ5, the literature analysis showed that different types of questions
+arise. Which are the best approaches to develop the analysis of the performance
+in predicting dropout? Several metrics are customer dropout is to improve the
+performance of organizations in retaining customers, which is a management
+problem in which data mining is adopted \[@verbeke_new_2012\]. The goals of the
+model should be formulated considering the context of the problem that is being
+addressed; in marketing retention strategies, the up-lift supports the
+development of proactive actions to minimize the investment in retention
+strategies \[@coussement_churn_2008\]. Some assumptions that underlie the
+adoption of uplift metrics consider that customers with a higher risk of
+churning could not be the best targets, as suggested by \protect\hyperlink{ref-ascarza_retention_2018}{Ascarza} (\protect\hyperlink{ref-ascarza_retention_2018}{2018}).
+Other researchers addressed the problem using the top-decile lift to develop
+more proactive actions to retain the customers at risk of churning
+\[@coussement_churn_2008;xie_customer_2009\]. This approach considers the 10\% of
+customers with more risk, and investments in retention strategies should be
+developed that distinguish the churners susceptible to marketing actions from
+those who will leave anyway (\protect\hyperlink{ref-coussement_comparative_2017}{Coussement, Lessmann, and Verstraeten 2017}). Although uplift
+models seem to be good strategies, they should also used, such as AUC,
+sensitivity, specificity, recall, precision, and F-score. However, the goal
+ofconsider factors other than risk and customer satisfaction, as not taking this
+into consideration could be counterproductive and the model should be removed
+from the retention strategy.
+
+The true business objective is to reduce customer churn. Customers who are about
+to churn but cannot be retained should be excluded from the campaign, as
+targeting them will be a waste of scarce resources (\protect\hyperlink{ref-devriendt_why_2019}{Devriendt et al. 2019}). Using
+these models seem to be a good strategy, as they can outperform predictive
+models that consider only accuracy from a profitability busshould be considered
+that customers with a higher risk of churning may not be the best targets to
+develop retention strategies. Those perspectives entail the dropout.
+
+that a business context, or the clarification of a business objective underlying
+the prediction of customer dropout, should be developed, to clarify which
+objectives should be achieved before employing the profitability of reducing g
+machine learning algorithms. Surprisingly, the analyzed studies did not address
+the customer lifetime value as an objective to optimize considerininess
+perspective.
+
+\hypertarget{aspects-to-consider}{%
+\section{Aspects to consider}\label{aspects-to-consider}}
+
+\begin{itemize}
+\tightlist
+\item
+  Interpretability from RQ2
+\item
+  The business objective is to increase the number of members and organization
+  profits
+\item
+  piping several algorithms to improve accuracy. Aka hybrid approach
+\item
+  \protect\hyperlink{ref-Alboukaey_dynamic_2020}{Alboukaey et al.} (\protect\hyperlink{ref-Alboukaey_dynamic_2020}{2020b}) proposes \ldots{}
+\item
+  grep the articles addressing hybrid: pdfgrep -ri ``hybrid.\{1,10\} approach''
+\end{itemize}
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+In this section, we present our experiments to validate the proposed models,
+comparing against other approaches. The models where optimized using the
+hyper-parameters Grid Search technique. The explored hyper-parameters and the
+best values of these parameters for every model are listed in (Table 3).
+
+\begin{table}[ht]
+    \footnotesize
+    \centering
+    \begin{tabular}{p{0.15\textwidth}p{0.55\textwidth}p{0.25\textwidth}}
+    \hline
+    \textbf{Model name} & \textbf{Explored parameters values} & \textbf{Best parameters} \\
+    \hline
+    Survival trees  & pysurvival random forest                  & a \\
+    Survival trees
+    with clusters   & pysurvival random forest with clusters    & a \\
+    Scikit survival 
+    trees           & scikit survival                           & a \\
+    Scikit survival 
+    with clusters   & scikit with clusters                      & a \\
+    Scikit survival 
+    gradient boost  & scikit survival gradient boost            & a \\
+    Scikit survival 
+    gradient boost 
+    with clusters   & scikit with clusters                      & a \\
+    \hline
+    \end{tabular}
+    \caption{Hyper-parameters best values}
+    \label{hyperparametersbestvalues}
+\end{table}
+
+The results of the performance of the models are available in table 4. Colocar o
+modelo. Resultados do modelo
+
+\begin{table}[ht]
+    \footnotesize
+    \centering
+    \begin{tabular}{p{0.15\textwidth}p{0.35\textwidth}r p{0.10\textwidth}}
+    \hline
+    \textbf{Model name} & \textbf{Results} & \textbf{n}\\
+    \hline
+    Survival trees  & RMSE 57.815                                     & 25316\\
+                    & MAE 18.966                                      & \\   
+                    & MEAE 38.557                                     & \\
+    Survival trees  & Cluster 0: RMSE  2.785 MAE  0.000 MEAE  0.899   &   930\\
+    with clusters   & Cluster 1: RMSE 78.141 MAE 28.537 MEAE 56.989   & 17067\\
+                    & Cluster 2: RMSE 15.081 MAE  1.995 MEAE  6.850   &  2423\\
+                    & Cluster 3: RMSE  9.354 MAE  2.506 MEAE  6.427   &  2816\\
+                    & Cluster 4: RMSE  8.805 MAE  1.349 MEAE  3.725   &  2080\\
+
+    Scikit survival 
+    trees           & scikit survival                                 & \\
+    Scikit survival 
+    with clusters   & scikit with clusters                            & \\
+    Scikit survival 
+    gradient boost  & scikit survival gradient boost                  & \\
+    Scikit survival 
+    gradient boost 
+    with clusters   & scikit with clusters                            & \\
+    \hline
+    \end{tabular}
+    \caption{Hyper-parameters best values}
+    \label{hyperparametersbestvalues}
+    * Root Mean Square Error (RMSE), Mean Absolute Error (MAE), Median Absolute Error (MEAE)  
+\end{table}
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+Article Ascarza
+
+\begin{itemize}
+\tightlist
+\item
+  Retention Futility: Targeting High-Risk Customers Might be Ineffective
+  (\protect\hyperlink{ref-ascarza_retention_2018}{Ascarza 2018})
+\end{itemize}
+
+Ascarza, E. (2018). Retention Futility: Targeting High-Risk Customers Might be
+Ineffective. Journal of Marketing Research, 55(1), 80-98. sim.
+\url{https://doi.org/10.1509/jmr.16.0163}
+
+Example of Developed actions place in the discussion:
 
@@
 \begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(stargazer)}
-\NormalTok{model1 }\OtherTok{\textless{}{-}} \FunctionTok{lm}\NormalTok{(speed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ dist, }\AttributeTok{data =}\NormalTok{ cars)}
-\NormalTok{model2 }\OtherTok{\textless{}{-}} \FunctionTok{lm}\NormalTok{(speed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ dist, }\AttributeTok{data =}\NormalTok{ cars)}
-\NormalTok{model3 }\OtherTok{\textless{}{-}} \FunctionTok{lm}\NormalTok{(dist }\SpecialCharTok{\textasciitilde{}}\NormalTok{ speed, }\AttributeTok{data =}\NormalTok{ cars)}
-\FunctionTok{stargazer}\NormalTok{(model1, model2, model3,}
-          \AttributeTok{title =} \StringTok{"Regression table with stargazer"}\NormalTok{,}
-          \AttributeTok{label =} \StringTok{"tab2"}\NormalTok{,}
-          \AttributeTok{table.placement =} \StringTok{"h"}\NormalTok{,}
-          \AttributeTok{column.labels =} \FunctionTok{c}\NormalTok{(}\StringTok{"M1"}\NormalTok{, }\StringTok{"M2"}\NormalTok{, }\StringTok{"M3"}\NormalTok{),}
-          \AttributeTok{model.numbers =} \ConstantTok{FALSE}\NormalTok{,}
-          \AttributeTok{header =} \ConstantTok{FALSE}\NormalTok{)}
+\NormalTok{Each month, the company identified the customers who were up for renewal and}
+\NormalTok{split them (randomly and evenly) between a treatment group that received a }
+\NormalTok{"thank you" gift with the letter and a control group that received only the }
+\NormalTok{renewal latter.}
 \end{Highlighting}
 \end{Shaded}
-
-\begin{table}[h] \centering 
-  \caption{Regression table with stargazer} 
-  \label{tab2} 
-\begin{tabular}{@{\extracolsep{5pt}}lccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
- & \multicolumn{3}{c}{\textit{Dependent variable:}} \\ 
-\cline{2-4} 
-\\[-1.8ex] & \multicolumn{2}{c}{speed} & dist \\ 
- & M1 & M2 & M3 \\ 
-\hline \\[-1.8ex] 
- dist & 0.166$^{***}$ & 0.166$^{***}$ &  \\ 
-  & (0.017) & (0.017) &  \\ 
-  & & & \\ 
- speed &  &  & 3.932$^{***}$ \\ 
-  &  &  & (0.416) \\ 
-  & & & \\ 
- Constant & 8.284$^{***}$ & 8.284$^{***}$ & $-$17.579$^{**}$ \\ 
-  & (0.874) & (0.874) & (6.758) \\ 
-  & & & \\ 
-\hline \\[-1.8ex] 
-Observations & 50 & 50 & 50 \\ 
-R$^{2}$ & 0.651 & 0.651 & 0.651 \\ 
-Adjusted R$^{2}$ & 0.644 & 0.644 & 0.644 \\ 
-Residual Std. Error (df = 48) & 3.156 & 3.156 & 15.380 \\ 
-F Statistic (df = 1; 48) & 89.567$^{***}$ & 89.567$^{***}$ & 89.567$^{***}$ \\ 
-\hline 
-\hline \\[-1.8ex] 
-\textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
-\end{tabular} 
-\end{table}
-
-\hypertarget{figures}{%
-\section{Figures}\label{figures}}
-
-\hypertarget{graphs-with-r}{%
-\subsection{Graphs with R}\label{graphs-with-r}}
-
-You can insert figures like this. One would like to produce and insert them on
-the fly in the \texttt{.rmd} file. Figure \ref{fig:fig-1} is such an example.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{plot}\NormalTok{(cars}\SpecialCharTok{$}\NormalTok{speed, cars}\SpecialCharTok{$}\NormalTok{dist)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{articleCustomerDropoutMembership_files/figure-latex/Figures 1, fig-1-1}
-
-}
-
-\caption{Scatterplot of Speed and Distance}(\#fig:Figures 1, fig-1)
-\end{figure}
-
-However, in some cases it does not work.
-
-\hypertarget{example-ggplot2-graphs}{%
-\subsection{Example: ggplot2 graphs}\label{example-ggplot2-graphs}}
-
-See the \texttt{ggplot2} output reported in Figure \ref{fig:fig-2}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{mtcars}\SpecialCharTok{$}\NormalTok{cyl }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(mtcars}\SpecialCharTok{$}\NormalTok{cyl) }\CommentTok{\# Convert cyl to factor}
-\FunctionTok{library}\NormalTok{(ggplot2)}
-\FunctionTok{ggplot}\NormalTok{(mtcars, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ wt, }\AttributeTok{y =}\NormalTok{ mpg, }\AttributeTok{shape =}\NormalTok{ cyl)) }\SpecialCharTok{+} \FunctionTok{geom\_point}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Weight (lb/1000)"}\NormalTok{, }\AttributeTok{y =} \StringTok{"Miles/(US) gallon"}\NormalTok{,}
-       \AttributeTok{shape =} \StringTok{"Number of }\SpecialCharTok{\textbackslash{}n}\StringTok{ Cylinders"}\NormalTok{) }\SpecialCharTok{+} \FunctionTok{theme\_classic}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}[h]
-
-{\centering \includegraphics{articleCustomerDropoutMembership_files/figure-latex/Fig-2-1}
-
-}
-
-\caption{Miles per gallon according to the weight}\label{fig:Fig-2}
-\end{figure}
-
-\hypertarget{another-example-using-plotly}{%
-\subsection{Another example using Plotly}\label{another-example-using-plotly}}
-
-With \texttt{Plotly} we can produce interactive graphs which play well, for example,
-once can embedded in html webpages (drop by
-\href{https://paulcbauer.shinyapps.io/visualizing-causal-scenarios/}{here} for an
-example). One can insert this type of graphs in R Markdown PDF using \texttt{Orca} (it
-generates static images from Plotly graphs). Go
-\href{https://github.com/plotly/orca\#installation}{here} to check how to install it.
-See Figure \ref{fig:fig-3} for an example.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(plotly)}
-\NormalTok{p }\OtherTok{\textless{}{-}} \FunctionTok{plot\_ly}\NormalTok{(cars, }\AttributeTok{type =} \StringTok{"scatter"}\NormalTok{, }\AttributeTok{mode =} \StringTok{"markers"}\NormalTok{,}
-        \AttributeTok{x =} \SpecialCharTok{\textasciitilde{}}\NormalTok{speed,}
-        \AttributeTok{y =} \SpecialCharTok{\textasciitilde{}}\NormalTok{dist)}
-\CommentTok{\#Sys.setenv(\textquotesingle{}MAPBOX\_TOKEN\textquotesingle{} = \textquotesingle{}12423423\textquotesingle{}) \# set arbitrary token}
-\CommentTok{\#orca(p, "logs/plotly{-}plot.pdf")}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}[ht]
-\centering
-\caption{Example: export a Plotly figure using `orca`}\label{fig:fig-3}
-    \includegraphics[width = 0.9\linewidth]{../figures/teste.png}
-\begin{flushleft}
-\end{flushleft}
-\end{figure}
-\vspace{-1.2cm}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Lets create a value for example}
-
-\NormalTok{media }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(cars}\SpecialCharTok{$}\NormalTok{speed)}
-\end{Highlighting}
-\end{Shaded}
-
-The criminal rate is 15.4\%o.
-
-\vspace{0.3cm}
-
-\hypertarget{miguels-tests}{%
-\section{Miguel's tests}\label{miguels-tests}}
-
-\hypertarget{r}{%
-\subsection{R}\label{r}}
-
-Example of an equation
-
-\[\int_0^{2\pi} \sin x~dx\]
-
-\emph{Example of a matrix}
-
-\[
-\mathbf{X} = \left[\begin{array}
-{rrr}
-1 & 2 & 3 \\
-4 & 5 & 6 \\
-7 & 8 & 9
-\end{array}\right]
-\]
-
-or
-
-\begin{equation}
-f\left(k\right) = \binom{n}{k}p^k\left(1-p\right)^{n-k} \label{eq:binom}
-\end{equation}
-
-See Equation \eqref{eq:binom}.
-
-\begin{align}
-y_{ijt} = \beta x_{ijt} + \eta_i + \gamma_j + \lambda_t + \varepsilon_{ijt}
-\end{align}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(stargazer)}
-\FunctionTok{stargazer}\NormalTok{(cars,}
-          \AttributeTok{title =} \StringTok{"Summary 24"}\NormalTok{,}
-          \AttributeTok{label =} \StringTok{"tab24"}\NormalTok{,}
-          \AttributeTok{table.placement =} \StringTok{"ht"}\NormalTok{,}
-          \AttributeTok{header =} \ConstantTok{FALSE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{table}[ht] \centering 
-  \caption{Summary 24} 
-  \label{tab24} 
-\begin{tabular}{@{\extracolsep{5pt}}lccccccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
-Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St.
-Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Pctl(25)} & \multicolumn{1}{c}{Pctl(75)} & \multicolumn{1}{c}{Max} \\
-\hline \\[-1.8ex] 
-speed & 50 & 15.400 & 5.288 & 4 & 12 & 19 & 25 \\ 
-dist & 50 & 42.980 & 25.769 & 2 & 26 & 56 & 120 \\ 
-\hline \\[-1.8ex] 
-\end{tabular} 
-\end{table}
-
-\hypertarget{final-remarks}{%
-\section{Final remarks}\label{final-remarks}}
-
-Check the replication package for Bonhomme, Lamadon and Manresa (2019):
-\url{https://github.com/tlamadon/blm-replicate}
-
-\newpage
 
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-akogul2016}{}}%
+Akogul, Serkan and Murat Erisoglu.
+2016.
+{``A Comparison of Information Criteria in Clustering Based on Mixture of Multivariate Normal Distributions.''} \emph{Mathematical and Computational Applications} 21(3):34.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Alboukaey_dynamic_2020}{}}%
+Alboukaey, Nadia, Ammar Joukhadar, and Nada Ghneim. 2020b. {``Dynamic Behavior Based Churn Prediction in Mobile Telecom.''} \emph{Expert Systems with Applications} 162:113779.
+
+\leavevmode\vadjust pre{\hypertarget{ref-alboukaey_dynamic_2020}{}}%
+Alboukaey, Nadia, Ammar Joukhadar, and Nada Ghneim. 2020a. {``Dynamic Behavior Based Churn Prediction in Mobile Telecom.''} \emph{Expert Systems with Applications} 162:113779.
+
+\leavevmode\vadjust pre{\hypertarget{ref-amin_customer_2017}{}}%
+Amin, Adnan, Sajid Anwar, Awais Adnan, Muhammad Nawaz, Khalid Alawfi, Amir Hussain, and Kaizhu Huang.
+2017.
+{``Customer Churn Prediction in the Telecommunication Sector Using a Rough Set Approach.''} \emph{Neurocomputing} 237:242--54.
+
+\leavevmode\vadjust pre{\hypertarget{ref-antipov_applying_2010}{}}%
+Antipov, Evgeny and Elena Pokryshevskaya.
+2010.
+{``Applying {CHAID} for Logistic Regression Diagnostics and Classification Accuracy Improvement.''} \emph{Journal of Targeting, Measurement and Analysis for Marketing} 18(2):109--17.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ascarza_retention_2018}{}}%
+Ascarza, Eva. 2018. {``Retention Futility: Targeting High-Risk Customers Might Be Ineffective.''} \emph{Journal of Marketing Research} 55(1):80--98.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Ascarza_Hardie_2013}{}}%
+Ascarza, Eva and Bruce G. S. Hardie. 2013. {``A Joint Model of Usage and Churn in Contractual Settings.''} \emph{Marketing Science} 32(4):570--90.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Athanassopoulos_2000}{}}%
+Athanassopoulos, Antreas D. 2000. {``Customer Satisfaction Cues to Support Market Segmentation and Explain Switching Behavior.''} \emph{Journal of Business Research} 47(3):191--207.
+
+\leavevmode\vadjust pre{\hypertarget{ref-azeem_churn_2017}{}}%
+Azeem, Muhammad, Muhammad Usman, and A. C. M. Fong. 2017. {``A Churn Prediction Model for Prepaid Customers in Telecom Using Fuzzy Classifiers.''} \emph{Telecommunication Systems} 66(4):603--14.
+
+\leavevmode\vadjust pre{\hypertarget{ref-benoit_improving_2012}{}}%
+Benoit, Dries F. and Dirk Van den Poel. 2012. {``Improving Customer Retention in Financial Services Using Kinship Network Information.''} \emph{Expert Systems with Applications} 39(13):11435--42.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Breiman_2001}{}}%
+Breiman, Leo. 2001. {``Random Forests.''} \emph{Machine Learning} 45(1):5--32.
+
+\leavevmode\vadjust pre{\hypertarget{ref-burez_crm_2007}{}}%
+Burez, Jonathan and Dirk Van den Poel.
+2007.
+{``{CRM} at a Pay-{TV} Company: {Using} Analytical Models to Reduce Customer Attrition by Targeted Marketing for Subscription Services.''} \emph{Expert Systems with Applications} 32(2):277--88.
+
+\leavevmode\vadjust pre{\hypertarget{ref-coussement_improved_2010}{}}%
+Coussement, Kristof, Dries F.
+Benoit, and Dirk Van den Poel.
+2010.
+{``Improved Marketing Decision Making in a Customer Churn Prediction Context Using Generalized Additive Models.''} \emph{Expert Systems with Applications} 37(3):2132--43.
+
+\leavevmode\vadjust pre{\hypertarget{ref-coussement_comparative_2017}{}}%
+Coussement, Kristof, Stefan Lessmann, and Geert Verstraeten.
+2017.
+{``A Comparative Analysis of Data Preparation Algorithms for Customer Churn Prediction: {A} Case Study in the Telecommunication Industry.''} \emph{Decision Support Systems} 95:27--36.
+
+\leavevmode\vadjust pre{\hypertarget{ref-coussement_churn_2008}{}}%
+Coussement, Kristof and Dirk Van den Poel.
+2008.
+{``Churn Prediction in Subscription Services: {An} Application of Support Vector Machines While Comparing Two Parameter-Selection Techniques.''} \emph{Expert Systems with Applications} 34(1):313--27.
+
+\leavevmode\vadjust pre{\hypertarget{ref-coussement_improving_2009}{}}%
+Coussement, Kristof and Dirk Van den Poel.
+2009.
+{``Improving Customer Attrition Prediction by Integrating Emotions from Client/Company Interaction Emails and Evaluating Multiple Classifiers.''} \emph{Expert Systems with Applications} 36(3, Part 2):6127--34.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Davidson-Pilon_2021}{}}%
+Davidson-Pilon, Cameron. 2021. \emph{CamDavidsonPilon/Lifelines}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-devriendt_why_2019}{}}%
+Devriendt, Floris, Jeroen Berrevoets, and Wouter Verbeke. 2019. {``Why You Should Stop Predicting Customer Churn and Start Using Uplift Models.''} \emph{Information Sciences}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Ehrlinger_2016}{}}%
+Ehrlinger, John. 2016. {``ggRandomForests: Exploring Random Forest Survival.''} \emph{arXiv:1612.08974 {[}Stat{]}}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-esteves_churn_2016}{}}%
+Esteves, Georgina and Joao Mendes-Moreira.
+2016.
+{``Churn Perdiction in the Telecom Business.''} Pp.
+254--59 in \emph{2016 {Eleventh} {International} {Conference} on {Digital} {Information} {Management} ({ICDIM})}.
+Porto, Portugal: IEEE.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Fotso_others_2019}{}}%
+Fotso, Stephane and others. 2019. \emph{PySurvival: Open Source Package for Survival Analysis Modeling}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-garcia_intelligent_2017}{}}%
+García, David L., Angela Nebot, and Alfredo Vellido. 2017. {``Intelligent Data Analysis Approaches to Churn as a Business Problem: A Survey.''} \emph{Knowledge and Information Systems} 51(3):719--74.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gok_case_2015}{}}%
+Gök, Mehmet, Tansel Özyer, and Jamal Jida.
+2015.
+{``A {Case} {Study} for the {Churn} {Prediction} in {Turksat} {Internet} {Service} {Subscription}.''} Pp.
+1220--24 in \emph{Proceedings of the 2015 {IEEE}/{ACM} {International} {Conference} on {Advances} in {Social} {Networks} {Analysis} and {Mining} 2015 - {ASONAM} '15}.
+Paris, France: ACM Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hanson_Sugden_Alberts_2011}{}}%
+Hanson, Brooks, Andrew Sugden, and Bruce Alberts. 2011. {``Making Data Maximally Available.''} \emph{Science (New York, N.Y.)} 331(6018):649.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hung_applying_2006}{}}%
+Hung, Shin-Yuan, David C. Yen, and Hsiu-Yu Wang. 2006. {``Applying Data Mining to Telecom Churn Management.''} \emph{Expert Systems with Applications} 31(3):515--24.
+
+\leavevmode\vadjust pre{\hypertarget{ref-keramati_improved_2014}{}}%
+Keramati, A., R.
+Jafari-Marandi, M.
+Aliannejadi, I.
+Ahmadian, M.
+Mozaffari, and U.
+Abbasi.
+2014.
+{``Improved Churn Prediction in Telecommunication Industry Using Data Mining Techniques.''} \emph{Applied Soft Computing} 24:994--1012.
+
+\leavevmode\vadjust pre{\hypertarget{ref-perianez_churn_2016}{}}%
+Perianez, Africa, Alain Saas, Anna Guitart, and Colin Magne.
+2016.
+{``Churn {Prediction} in {Mobile} {Social} {Games}: {Towards} a {Complete} {Assessment} {Using} {Survival} {Ensembles}.''} Pp.
+564--73 in \emph{2016 {IEEE} {International} {Conference} on {Data} {Science} and {Advanced} {Analytics} ({DSAA})}.
+Montreal, QC, Canada: IEEE.
+
+\leavevmode\vadjust pre{\hypertarget{ref-risselada_staying_2010}{}}%
+Risselada, Hans, Peter C. Verhoef, and Tammo H. A. Bijmolt. 2010. {``Staying {Power} of {Churn} {Prediction} {Models}.''} \emph{Journal of Interactive Marketing} 24(3):198--208.
+
+\leavevmode\vadjust pre{\hypertarget{ref-routh_estimating_2020}{}}%
+Routh, Pallav, Arkajyoti Roy, and Jeff Meyer. 2020. {``Estimating Customer Churn Under Competing Risks.''} \emph{Journal of the Operational Research Society} 1--18.
+
+\leavevmode\vadjust pre{\hypertarget{ref-schwarz1978}{}}%
+Schwarz, Gideon. 1978. {``Estimating the Dimension of a Model.''} \emph{The Annals of Statistics} 6(2):461--64.
+
+\leavevmode\vadjust pre{\hypertarget{ref-scrucca2016}{}}%
+Scrucca, Luca, Michael Fop, T.
+,Brendan Murphy, and Adrian,E.
+Raftery.
+2016.
+{``Mclust 5: Clustering, Classification and Density Estimation Using Gaussian Finite Mixture Models.''} \emph{The R Journal} 8(1):289.
+
+\leavevmode\vadjust pre{\hypertarget{ref-verbeke_new_2012}{}}%
+Verbeke, Wouter, Karel Dejaeger, David Martens, Joon Hur, and Bart Baesens.
+2012.
+{``New Insights into Churn Prediction in the Telecommunication Sector: {A} Profit Driven Data Mining Approach.''} \emph{European Journal of Operational Research} 218(1):211--29.
+
+\leavevmode\vadjust pre{\hypertarget{ref-vijaya_sivasankar_2019}{}}%
+Vijaya, J.
+and E.
+Sivasankar.
+2019.
+{``An Efficient System for Customer Churn Prediction Through Particle Swarm Optimization Based Feature Selection Model with Simulated Annealing.''} \emph{Cluster Computing} 22(S5):10757--68.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wangmachine2017}{}}%
+Wang, Ping, Yan Li, and Chandan K. Reddy. 2017. {``Machine Learning for Survival Analysis: A Survey.''} \emph{arXiv:1708.04649 {[}Cs, Stat{]}}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-xie_customer_2009}{}}%
+Xie, Yaya, Xiu Li, E. W. T. Ngai, and Weiyun Ying. 2009. {``Customer Churn Prediction Using Improved Balanced Random Forests.''} \emph{Expert Systems with Applications} 36(3, Part 1):5445--49.
+
 \end{CSLReferences}
@@
 
-\hypertarget{r-1}{%
-\subsubsection{R}\label{r-1}}
+\hypertarget{r}{%
+\subsubsection{R}\label{r}}
 
@@
 ## # Platform: x86_64-w64-mingw32/x64 (64-bit) 
-## # Running under: Windows 10 x64 (build 19043) 
+## # Running under: Windows 10 x64 (build 22000) 
 ## #  
@@
 ## # other attached packages: 
-## #  [1] plotly_4.9.4.1   kableExtra_1.3.4 gtsummary_1.4.2  readxl_1.3.1     
-## #  [5] stargazer_5.2.2  naniar_0.6.1     visdat_0.5.3     ggplot2_3.3.4    
-## #  [9] dlookr_0.4.5     dplyr_1.0.7      
+## #  [1] labelled_2.8.0   kableExtra_1.3.4 gtsummary_1.4.2  visdat_0.5.3     
+## #  [5] readxl_1.3.1     stargazer_5.2.2  reticulate_1.22  ggplot2_3.3.5    
+## #  [9] dlookr_0.5.1     dplyr_1.0.7      
 ## #  
 ## # loaded via a namespace (and not attached): 
-## #   [1] webshot_0.5.2       RColorBrewer_1.1-2  httr_1.4.2          
-## #   [4] UpSetR_1.4.0        tools_4.1.0         backports_1.2.1     
-## #   [7] utf8_1.2.1          R6_2.5.0            rpart_4.1-15        
-## #  [10] lazyeval_0.2.2      Hmisc_4.5-0         nortest_1.0-4       
-## #  [13] colorspace_2.0-1    nnet_7.3-16         withr_2.4.2         
-## #  [16] tidyselect_1.1.1    gridExtra_2.3       curl_4.3.1          
-## #  [19] compiler_4.1.0      extrafontdb_1.0     cli_2.5.0           
-## #  [22] rvest_1.0.0         gt_0.3.0            htmlTable_2.2.1     
-## #  [25] xml2_1.3.2          sandwich_3.0-1      labeling_0.4.2      
-## #  [28] bookdown_0.22       scales_1.1.1        checkmate_2.0.0     
-## #  [31] mvtnorm_1.1-2       proxy_0.4-26        RcmdrMisc_2.7-1     
-## #  [34] systemfonts_1.0.2   stringr_1.4.0       digest_0.6.27       
-## #  [37] foreign_0.8-81      rmarkdown_2.9       svglite_2.0.0       
-## #  [40] rio_0.5.27          base64enc_0.1-3     jpeg_0.1-8.1        
-## #  [43] pkgconfig_2.0.3     htmltools_0.5.1.1   extrafont_0.17      
-## #  [46] highr_0.9           htmlwidgets_1.5.3   rlang_0.4.11        
-## #  [49] rstudioapi_0.13     prettydoc_0.4.1     farver_2.1.0        
-## #  [52] generics_0.1.0      jsonlite_1.7.2      zoo_1.8-9           
-## #  [55] crosstalk_1.1.1     zip_2.2.0           car_3.0-11          
-## #  [58] magrittr_2.0.1      Formula_1.2-4       Matrix_1.3-3        
-## #  [61] Rcpp_1.0.6          munsell_0.5.0       fansi_0.5.0         
-## #  [64] reticulate_1.20     abind_1.4-5         gdtools_0.2.3       
-## #  [67] partykit_1.2-13     lifecycle_1.0.0     stringi_1.6.1       
-## #  [70] yaml_2.2.1          inum_1.0-4          carData_3.0-4       
-## #  [73] MASS_7.3-54         plyr_1.8.6          grid_4.1.0          
-## #  [76] hrbrthemes_0.8.0    forcats_0.5.1       crayon_1.4.1        
-## #  [79] lattice_0.20-44     haven_2.4.1         splines_4.1.0       
-## #  [82] hms_1.1.0           knitr_1.33          pillar_1.6.1        
-## #  [85] glue_1.4.2          evaluate_0.14       latticeExtra_0.6-29 
-## #  [88] broom.helpers_1.3.0 data.table_1.14.0   png_0.1-7           
-## #  [91] vctrs_0.3.8         Rttf2pt1_1.3.8      cellranger_1.1.0    
-## #  [94] tidyr_1.1.3         gtable_0.3.0        purrr_0.3.4         
-## #  [97] xfun_0.24           openxlsx_4.2.4      libcoin_1.0-8       
-## # [100] e1071_1.7-7         class_7.3-19        survival_3.2-11     
-## # [103] viridisLite_0.4.0   tibble_3.1.2        cluster_2.1.2       
-## # [106] corrplot_0.89       ellipsis_0.3.2
+## #  [1] httr_1.4.2          tidyr_1.1.4         jsonlite_1.7.2      
+## #  [4] viridisLite_0.4.0   splines_4.1.0       Formula_1.2-4       
+## #  [7] shiny_1.7.1         highr_0.9           cellranger_1.1.0    
+## # [10] yaml_2.2.1          gdtools_0.2.3       Rttf2pt1_1.3.9      
+## # [13] pillar_1.6.3        lattice_0.20-44     glue_1.4.2          
+## # [16] extrafontdb_1.0     digest_0.6.28       promises_1.2.0.1    
+## # [19] rvest_1.0.1         colorspace_2.0-2    htmltools_0.5.2     
+## # [22] httpuv_1.6.3        Matrix_1.3-3        hrbrthemes_0.8.0    
+## # [25] pkgconfig_2.0.3     haven_2.4.3         bookdown_0.24       
+## # [28] purrr_0.3.4         xtable_1.8-4        mvtnorm_1.1-2       
+## # [31] scales_1.1.1        webshot_0.5.2       svglite_2.0.0       
+## # [34] later_1.3.0         tibble_3.1.4        farver_2.1.0        
+## # [37] generics_0.1.0      ellipsis_0.3.2      withr_2.4.2         
+## # [40] cli_3.0.1           survival_3.2-11     magrittr_2.0.1      
+## # [43] crayon_1.4.1        mime_0.12           pagedown_0.15       
+## # [46] evaluate_0.14       reactable_0.2.3     fansi_0.5.0         
+## # [49] broom.helpers_1.4.0 forcats_0.5.1       xml2_1.3.2          
+## # [52] tools_4.1.0         hms_1.1.1           lifecycle_1.0.1     
+## # [55] stringr_1.4.0       munsell_0.5.0       compiler_4.1.0      
+## # [58] inum_1.0-4          systemfonts_1.0.2   rlang_0.4.11        
+## # [61] grid_4.1.0          gt_0.3.1            rstudioapi_0.13     
+## # [64] rappdirs_0.3.3      htmlwidgets_1.5.4   labeling_0.4.2      
+## # [67] rmarkdown_2.11      partykit_1.2-15     gtable_0.3.0        
+## # [70] R6_2.5.1            gridExtra_2.3       knitr_1.36          
+## # [73] fastmap_1.1.0       extrafont_0.17      utf8_1.2.2          
+## # [76] libcoin_1.0-9       stringi_1.7.4       Rcpp_1.0.7          
+## # [79] vctrs_0.3.8         rpart_4.1-15        png_0.1-7           
+## # [82] tidyselect_1.1.1    xfun_0.26
 \end{verbatim}
@@
 
+\hypertarget{other-used-tools}{%
+\subsubsection{Other used tools}\label{other-used-tools}}
+
+\begin{itemize}
+\tightlist
+\item
+  \href{https://www.visidata.org/}{Visidata} for quick exploratory. VisiData is a
+  free, open-source tool that lets you quickly open, explore, summarize, and
+  analyze datasets in your computer's terminal.
+\end{itemize}
+
 \end{document}

### 2024-01-11 article/customer_fitness_artigo.Rmd
update

---
title: |
  |
  |
  |
  | \vspace{1cm}Health club customer dropout membership^[Corresponding address: sobreiro@esdrm.ipsantarem.pt. Quality of Life Research Centre, Polytechnic Institute of Santarém, Portugal]\vspace{0.5cm}
author: |
  | Pedro Sobreiro, Javier Berrocal, Domingos Martinho, José Garcia Alonso
  |
  | Extremadura University
  |
date: |
  |
  |
  | `r gsub("^0", "", format(Sys.time(), "%d %B, %Y"))`
  |
  |
linestretch: 1.2
colorlinks: true
abstract: \vspace{1cm} Customer retention is fundamental to improve the organizations profits. Existing approaches normaly explore static models predicting when the customer will dropout. In this paper we propose a different approach considering that the customer dropout risk changes over time using clusters to improve the model performance. We explore a survival model using random forests with and with out clusters in a dataset of 5209 customers. The model using clusters improved the performance significantly. This paper shows that the use of clusters should be considered to identify dropout patterns to support the timming when the dropout occurs considering the cluster where the customer is. Improving the performance as less errors and is expected to contribute to the identification when should be developed retention strategies. \vspace{.8cm} 
bibliography: references.bib
csl: american-sociological-association.csl
output:
  bookdown::pdf_document2:
    includes:
    toc: no
    keep_tex: true
    latex_engine: xelatex
    number_sections: true
  bookdown::word_document2:
    latex_engine: xelatex
    number_sections: true
mainfont: Times New Roman
sansfont: Times New Roman
fontsize: 12pt
link-citations: true
documentclass: article
geometry: margin = 1in
always_allow_html: yes
header-includes:
   - \usepackage{dcolumn}
   - \usepackage{color}
   - \usepackage{pdfpages}
   - \usepackage{amsmath}
   - \usepackage{booktabs}
   - \usepackage{makecell}
   - \usepackage{hyperref}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
editor_options: 
  markdown: 
    wrap: 80
---

```{r Setup, include = FALSE}
knitr::opts_chunk$set(cache = FALSE, warning = FALSE, message = FALSE)
# Use cache = TRUE if you want to speed up compilation
# set path
# get rmarkdown directory
caminho <- getwd()
# set working directory
setwd(caminho)
print(caminho)
# A function to allow for showing some of the inline code
rinline <- function(code) {
  html <- '<code  class = "r">``` `r CODE` ```</code>'
  sub("CODE", code, html)
}
```

```{r RETICULATE, message=FALSE, warning=FALSE, include=FALSE}
# load essential libraries
library(dplyr)
library(dlookr)
library(ggplot2)
library(reticulate)
#Replace by your environment, usually which python solves the problem
#conda env list also is a good option
#set environment first then call reticulate library
path_python_windows <- "C:\\Users\\sobre\\AppData\\Local\\r-miniconda\\envs\\rsurvival\\python.exe"
path_python_linux <- "/home/sobreiro/miniconda3/envs/survival/bin/python"
switch(Sys.info()[["sysname"]],
       Windows = {
            Sys.setenv(RETICULATE_PYTHON = path_python_windows)
            #call reticulate
            library(reticulate)
            #activate environment
            use_condaenv("rsurvival", required = TRUE)
            caminho_figuras <- "c:/nuvem/Dropbox/doutoramento/tese/9.tese_documento/PhD-Pedro-Sobreiro/figures/"
       },
       Linux = {
            Sys.setenv(RETICULATE_PYTHON = path_python_linux)
            library(reticulate)
            use_condaenv("survival", required = TRUE)
            matplotlib <- import("matplotlib")
            matplotlib$use("Agg", force = TRUE)
            # https://community.rstudio.com/t/reticulate-rstudio-matplotlib/33219
            caminho_figuras <- "/mnt/c/nuvem/Dropbox/doutoramento/tese/9.tese_documento/PhD-Pedro-Sobreiro/figures/"
       }
)
```

# Introduction

Customer retention is a problem is a problem being addressed using the dropout 
prediction as an insight to identify customers that could dropout.
The customers database is the most valuable asset that the organizations possess
[@Athanassopoulos_2000].
The problem of retention is not new, in a seminal paper @Copeland_1923 address
the problem of brand loyalty and in marketing research [@Mellens_Dekimpe_Steenkamp_1996].

The advantage of developing some strategies in retention are supported in the idea
that the costs of customer retention are lower than customer acquisition 
[@Fornell_Wernerfelt_1987; @Edward_Sahadev_2011].
The increase in the profits with the reduction of 5\% of the dropout could represent
almost a duplication of the profits [@Reichheld_1996].

The development of a customer retention strategy could be supported in the 
identification of the customers that will dropout [@alboukaey_dynamic_2020]. 
However, dropout has two underlying scenarios contractual and non-contractual 
settings {@Gupta2006Modeling; @ascarza_retention_2018},
in a contractual business the customer needs to renew their contracts to continue its usage
[@Ascarza_Hardie_2013], against non contractual where the firm has to infer if 
the customer is still active
However, in contractual settings the customer dropout represents an explicit
ending of a relationship which is more penalizing than non contractual settings
[@risselada_staying_2010]. 
This has implications to the profitability of the organizations increasing marketing 
costs and reducing sales [@amin_customer_2017].

The anticipation of the dropout allows the development of countermeasures to
reduce customer churn. Several studies address the problem related to customer
retention trying to improve the profitability [@coussement_improving_2009;
@garcia_intelligent_2017; @devriendt_why_2019].
Existing organizations are addressing this problem by shifting their target from 
capturing new customers to preserving existing ones [@garcia_intelligent_2017],
considering that investments in retention strategies are more profitable 
than acquiring new customers [@coussement_improving_2009].

The approaches normally employed use a dependent variable representing dropout
or non-dropout, without considering a dynamic perspective that the dropout risk
changes over time [@Alboukaey_dynamic_2020]. The survival models try to solve
this limitation [@routh_estimating_2020] capturing a temporal dimension of the
customer dropout [@perianez_churn_2016]. @perianez_churn_2016 used survival
analysis to predict also when the dropout will occur.

Other studies proposed also the integration of several algorithms to improve the
performance in the prediction of the dropout such the usage of clusters combined
with churn prediction [@hung_applying_2006; @gok_case_2015;
@vijaya_sivasankar_2019]. The approach relies in the assumption that combining
the customers in different clusters allows the improvement of the prediction
accuracy. @vijaya_sivasankar_2019 suggested the adoption hybrid models combining
more than one classier are achieving increased performance compared to those
using single classifiers.

There are several challenges around the timing related to dropout, or
considering the dynamic behavior of the customer in the intent to drop out
[@alboukaey_dynamic_2020]. The importance of understanding when dropout will
occur and the risk when discarding the temporal perspective of the problem seems
to be an element that should be addressed. Few studies considered this
[@perianez_churn_2016; @burez_separating_2008]. This shows an opportunity to
address the importance of the timeframe and its influence on the efficiency of
the model and also evalute if the combination of clusters could improve the
performance.

Survival analysis, which origin as stands in biomedical statistics, its especially 
well-suited to studying the timing of events in longitudinal data [@Singer_Willett_1993]. 
Survival analysis is a class of statistical methods modelling the occurrence and timing of 
an event, such as the customer dropout. 
Survival analysis allow us to examine not only if an event occurred but also how long it 
took to occur. 
A primary value of survival analysis, however, is to compare dropout probability for 
individuals classified with theoretically relevant variables. 
The survival methods have enjoyed and increasing popularity in several disciplines ranging 
from medicine to economics [@Singer_Willett_1993].

Random Survival Forests does not make the proportional hazards assumption
[@Ehrlinger_2016] and has the flexibility to model survivor curves that are of
dissimilar shapes for contrasting groups of subjects. Random Survival Forest is
an extension of Random Forest allowing efficient non-parametric analysis of time
to event data [@Breiman_2001]. This characteristics allow us to surpass the Cox
Regression limitation of the proportional hazard assumption, requiring to
exclude variables which not fulfill the model assumption. It was shown by
@Breiman_2001 that ensemble learning can be further improved by injecting
randomization into the base learning process - a method called Random Forests.

Some studies employed also a combination of clusters with the churn prediction 
[@hung_applying_2006;@gok_case_2015;@vijaya_sivasankar_2019], where the customers
where grouped in clusters to improve the accuracy within each cluster.
Clusters are approaches using unsupervised algorithms to group elements
with similar characteristics. 
This unsupervised method has been widely used employing approaches
such Hierarchical Clustering [@Saunders_1980], K-Means [@vijaya_sivasankar_2019], or
Random Forest Clustering [@Breiman_2001].
@JafariMarandi_Denton_Idris_Smith_Keramati_2020 explored an approach combining
clustering methods in parallel to a classification.

In this study, we investigate if an hybrid approach using clusters and random survival forests 
which have never been used in to predict membership in a health club using existing data 
improves the prediction accuracy. 
Our paper is organized as follows. In the next section, we address our research methodology,
using survival analysis to identify the survival probability along the time, how 
we address the problem of determining the clusters and the performance of the model 
to predict customer survival within each identified cluster.

# Survival analysis

Survival analysis focus in the analysis of the time until an event of interest,
and exploring its relationship with different factors.
The main advantage is related to the concept of censoring, indicating
that observations that are not complete related to the event of interest, e.g.
customers that didn't dropout yet, which are incorporated in the analysis.
This means that there are customers still active for which we don't know if 
the event of dropout has occurred, which is called censorship.
Survival models take censoring into account and incorporate this uncertainty, 
instead of predicting the time of event such in regression models, the survival models
allow to predict the probability of an event happens at a particular time.

The time of dropout is represented by T, which is a non-negative random variable, 
indicating the time period of the event occurring for a randomly selected individual 
from the population, representing the probability of an event to occur each time period given 
that has not already occurred in a previous time period, known as discrete-time hazard function 
[@Singer_Willett_1993]. 
The survival function represents the probability of an individual surviving after time t, 
S(t) = P(T $>$ t), t $\geq$ 0, with the properties S(0) = 1,S($\infty$) = 0. 
The distribution function is represented with F, defined as F(t) = P(T $\leq$ 0), for t $\geq$ 0. 
The function of probability density represented with f where:

\begin{equation}
f(t) = \lim_{dt{\to 0}} \frac{P[t \le T < t + dt ]}{dt}
\end{equation}

$f(t)dt$ represents the probability of an event occurring in the moment t. 
The need to represent the distribution evolution of the death probability along the time, 
uses to the hazard function, represented as:

\begin{equation}
\lambda(t) = \lim_{dt \to 0} \frac{P[t \le T < t + dt | T \ge t ]}{dt}
\end{equation}

The determination of the survival curves is based in the following elements: 
(1) the total value of observations removed during the time period (e.g. days, months or years), 
    either by dropout or by censorship; 
(2) observations that composed the sample of the study; 
(3) customers who had not yet dropped out at any given time. 
The survival probability until the time period ii ($p_i$) is calculated with:

\begin{equation}
p_i=\frac{r_i-d_i}{r_i}
\end{equation}

Where $r_i$ is the number of individuals that survived at the beginning of the period, 
$d_i$ the number of individuals who left during the period. 
The survival time estimate was also taken considering the month in which it is found (estimated). 
Cox's allow test difference between survival times. 
The advantage in using survival analysis was that allow us to detect if the risk of an event differs 
systematically across different people, using specific predictors. 
The coefficients in a Cox regression were related to the hazard, where a positive value 
represents a worse prognosis and the opposite, negative value a better prognosis.
The advantage of survival analysis was that allow us to include information of covariates
that were censored up to the censoring event.

The Cox PH model assumes the covariates to be time independent, for example
gender and age when where retrieved do not change over time [@Schober2018Survival]
Because the Cox model requires the hazards in both groups to be proportional, 
researchers are often asked to "test" whether hazards are proportional
[@Stensrud_Hernán_2020].
Considering this we explored other approach that allow us to develop this analysis without 
the proportional hazard assumptions, the survival trees.

# Survival Trees

Survival trees are methods based in tree based models based on Random Forest [@Breiman_2001].
Random survival forests is an ensemble method for analysis of right-censored
data [@Ishwaran_Kogalur_Blackstone_Lauer_2008], using randomization to 
improve the performance.
Random survival forests follows this framework [@Ishwaran_Kogalur_Blackstone_Lauer_2008]:

1. Draw B random samples of the same size from the original dataset with replacement. The samples that are not drawn are said to be out-of-bag (OOB). Grow a survival tree on each of the b = 1 , . . . , B samples.
2. At each node, select a random subset of predictor variables and find the best predictor and splitting value that provide two subsets (the daughter nodes) which maximizes the difference in the objective function.
3. Repeat step 2 recursively on each daughter node until a stopping criterion is met.
4. Calculate a cumulative hazard function (CHF) for each tree and average over all CHFs for the B trees to obtain the ensemble CHF.
5. Compute the prediction error for the ensemble CHF using only the OOB data.

In each node is selected a predictor $x$ from a random selected predicted variables
and split value $c$ (one unique value of $x$).
Each sample $i$ if assigned the daughter right node if $x_i \le c$ or left if 
$x_i \ge c$, then is calculated the logrank such as:

\begin{equation}
L(x, c) = \frac{ \sum^{N}_{i=1} \left(  d_{i, 1} - Y_{i,1} \frac{d_i}{Y_i} \right)  }
               { \sqrt{  \sum^{N}_{i=1}  \frac{Y_{i,1}}{Y_i} \left( 1 - \frac{Y_{i,1}}{Y_i} \right) \left( \frac{Y_i-d_i}{Y_i-1} \right) d_i  } }
\end{equation}

Where:

* $j$: Daughter node, $j \in \{1,2\}$
* $d_{i,j}$: Number of events at time $t_i$ in daughter node $j$
* $Y_{i,j}$: Number of elements that had the event or are in risk at time $t_i$
           in daughter node $j$
* $d_i$: Number of events at time $t_i$, such $d_i=\sum_j{d_{i,j}}$
* $Y_i$: Number of elements that experienced an event or are at risk at
      $t_i$ so $Y_i=\sum_j Y_{i,j}$

We loop every $x$ and $c$ until find $x^*$ that satisfy $|L(x^{*}, c^{*})| \geq |L(x, c)|$
for every $x$ and $c$.
The model performance was determined with the concordance probability (C-index),
Brier Score (BS) and Mean Absolute Error (MAE) [@wangmachine2017]. The feature
importance was determined calculating the difference between the true class
label and noised data [@Breiman_2001].

The BS is used to evaluate the predicted accuracy of the survival function
at a given time $t$. 
Representing the average square distance between the survival status and the
predicted survival probability, where the value 0 is the best possible outcome.

\begin{equation}
BS(t) =  \frac{1}{N} \sum_{i = 1}^{N} \left( \frac{\left( 0 - \hat{S}(t, \vec{x}_i)\right)^2 \cdot \mathbb{1}_{T_i \leq t, \delta_i = 1}}{ \hat{G}(T_i^-)} + \frac{ \left( 1 - \hat{S}(t, \vec{x}_i)\right)^2 \cdot \mathbb{1}_{T_i > t}}{ \hat{G}(t)} \right)
\end{equation}

The model should have a Brier score below $0.25$. 
Considering that if $\forall i \in [\![1, N]\!], \hat{S}(t, \vec{x}_i) = 0.5$, then $BS(t) = 0.25$.

# Methodology

To simplify the analysis, the survival probabilities are presented as a survival curve. 
The survival curve is a representation of the survival probabilities corresponding to a 
time where the events are observed [@Bland_Altman_1998].
The survival analysis was conducted using the package Lifelines [@Pilon_2021].
Where dropout is a binary value where one represent churn and zero not churn. The
dropout happens when a member does not have a payment.

The random survival forest was developed using the package PySurvival
[@Fotso_others_2019]. PySurvival is an open source python package for 
Survival Analysis modeling - the modeling concept used to analyze or predict 
when an event is likely to happen.
The model was built with with 70\% of the data for training and 30\% for testing.

The model performance was determined with the concordance probability (C-index),
Brier Score (BS) and Mean Absolute Error (MAE) [@wangmachine2017]. The feature
importance was determined calculating the difference between the true class
label and noised data [@Breiman_2001].

The BS measure the average discrepancies between the status (dropout/non-dropout)
and the estimated probabilities at a given time.
The Integrated Brier Score (IBS) was used to calculate the performance 
in all available times (from $t_1$ to $t_{max}$) as:

\begin{equation}
IBS = \int_{t_1}^{t_\text{max}} {BS}^c(t) d w(t)
\end{equation}

Representing the average square distance between the survival status and the
predicted survival probability, where the value 0 is the best possible outcome.

The calculation of he number of clusters used the package mclust [@scrucca2016]
using the Bayesian Information Criterion (BIC). The model that gives the minimum
BIC score can be selected as the best model [@schwarz1978] simplifying the
problem related to choosing the number of components and identifying the
structure of the covariance matrix, based on modelling with multivariate normal
distributions for each component that forms the data set [@akogul2016].

The hybrid approach was develop as follows:

1. Identify the optimal number of clusters using [@scrucca2016]
2. Fit the model using the identified number of clusters
3. Estimate for each element the cluster 
4. for each cluster follow the framework proposed by @Ishwaran_Kogalur_Blackstone_Lauer_2008
   to calculate the random survival model

# Dataset

In this study, data from 5,209 fitness customers was analysed (mean age = 27.88, SD=11.80 years) 
from a Portuguese fitness centre. 
The data was collected from software e\@sport (Cedis, Portugal) between 2014 and 2017. 
The information retrieved was: Age of the participants in years; Sex (0-female, 1-male); 
Non-attendance days before dropout; Total amount billed; Average number of visits per week;
Total number of visits; Weekly contracted accesses; Number of registration renewals; 
Number of customer referrals; Registration month; Customer enrolment duration; and status 
(dropout/non-dropout).
Dropout event occur when customer communicate the intention to terminate the contract or 
did not pay the monthly fee during 60 days.

```{r get_data, echo = FALSE, message = FALSE, warning=FALSE, include = FALSE}
library(stargazer)
library(readxl)
library(dplyr)
library(visdat)

df_members <- read_excel("data/fitness_customers.xlsx")

# rename column labels
names(df_members) <- c("id", "age", "sex", "dayswfreq", "tbilled",
                       "maccess", "freeuse", "nentries", "cfreq", "nrenewals",
                       "cref", "start_date", "months", "dropout")

# remove null values
df_members <- df_members[complete.cases(df_members), ]

names(df_members)

# select relevant variables
df_members <- df_members %>%
select(age, sex, dayswfreq, tbilled, maccess, freeuse,
       nentries, cfreq, months, dropout)

str(df_members)
vis_dat(df_members) #check
```

Table \@ref(tab:summarytable) shows data's summary statistics. The average age
is `r round(mean(df_members$age,na.rm = TRUE),1)` ±
`r round(sd(df_members$age,na.rm = TRUE),1)`, the entries are 
`r round(mean(df_members$nentries,na.rm = TRUE),0)` ±
`r  round(sd(df_members$nentries,na.rm = TRUE),1)` with an inscription period of
`r round(mean(df_members$months,na.rm = TRUE),0)` ±
`r  round(sd(df_members$months,na.rm = TRUE),1)` months.

```{r teste, eval=FALSE, message=FALSE, warning=FALSE, echo=FALSE, paged.print=FALSE}
library(stargazer)
df_summary <- summary(df_members)
stargazer(df_members,
          title = "Summary statistics",
          label = "tab1cars",
          table.placement = "h",
          header = FALSE,
          summary = TRUE,
          summary.stat = c("min", "p25", "median", "p75", "max", "median", "sd")
         )
```

```{r summarytable, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
library(gtsummary)
library(kableExtra)
library(labelled)

var_label(df_members$age) <- "Age in years"
var_label(df_members$sex) <- "Male or female"

tbl <- df_members %>%
    tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd})",
                                 all_categorical() ~ "{p}%"),
                type = list(age ~ "continuous")) %>%
    add_stat_label()

as_kable_extra(tbl, booktabs = T,
               caption = "Summary statistics of features used")
```
Figure \@ref(fig:membershipmonth) shows the distribution of the dropout
considering the number of months of membership.

```{r, label="membershipmonth", eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Number of members by month"}
members_month <- df_members %>%
  select(months, dropout) %>%
  group_by(months, dropout) %>%
  summarize(count = n())

members_month$dropout <- factor(members_month$dropout)
g <- ggplot(data = members_month,
            mapping = aes(x = months, y = count, linetype = dropout))
g <- g + geom_line() + labs(x = "months", y = "Number of members") +
    theme_classic()
g

# Export the file 
# ggsave(file="/mnt/c/nuvem/Dropbox/doutoramento/tese/9.tese_documento/PhD-Pedro-Sobreiro/figures/membershipyear.pdf",device="pdf")
```

```{python correlationMatrix, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "Correlation matrix"}
from pysurvival.utils.display import correlation_matrix
import matplotlib.pyplot as plt
import pandas as pd
import numpy as np

col = ['sex']

df_members = r.df_members #copy r dataframe to python

# convert to int
df_members['age']=df_members['age'].astype(int)
df_members['dayswfreq']=df_members['dayswfreq'].astype(int)
df_members['cfreq']=df_members['cfreq'].astype(int)
df_members['months']=df_members['months'].astype(int)
df_members['dropout']=df_members['dropout'].astype(int)
df_members['sex']=df_members['sex'].astype(int)

df_members = pd.get_dummies(df_members, columns=col,drop_first=True)

# Creating the time and event columns
time_column = 'months'
event_column = 'dropout'

# Extracting the features
features = np.setdiff1d(df_members.columns, [time_column, event_column] ).tolist()

correlation_matrix(df_members[features], figure_size=(10,10), text_fontsize=6)

r.df_members = df_members

#plt.plot()
```

# Results

The table \@ref(tab:survivalprobabilities) depicts the data of the survival time of the customers 
during the first months, the results showed that the customers have a survival probability of 24.44\% 
at 12 months (column $p_i$ - likelihood probability) with a median survival time of 10 months
(column estimated_survival).
The survival probability at 6 months was 54.5\%, representing an risk of dropout of 
45.5\% with a estimated survival of 6 months.

```{python lifelines, include=FALSE}
from lifelines import KaplanMeierFitter
kmf = KaplanMeierFitter()
T = df_members['months']
C = df_members['dropout']

kmf.fit(T,C,label="Customers")

kmf.event_table.reset_index()
kmf.conditional_time_to_event_

survival_table = pd.concat([kmf.event_table.reset_index(),
                            kmf.conditional_time_to_event_.reset_index(),
                            kmf.survival_function_.reset_index()],axis=1)

survival_table.drop(['timeline'],axis=1,inplace=True)
survival_table.columns = ['event_at', 'removed', 'observed', 'censored', 
                          'entrance', 'at_risk', 'estimated_survival', 
                          'prob']
```

```{r survivalprobabilities, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
library(gtsummary)
library(kableExtra)
library(labelled)
survival_table <- py$survival_table
var_label(survival_table$event_at) <- "Event Month"
var_label(survival_table$removed) <- "Removed"
var_label(survival_table$observed) <- "Dropout"
var_label(survival_table$censored) <- "Censored"
var_label(survival_table$at_risk) <- "Risk of dropout"
var_label(survival_table$estimated_survival) <- "Estimated survival (months)"
var_label(survival_table$prob) <- "Survival Probability"

tbl <- head(survival_table, 25)

kbl(tbl, booktabs = T,
    format = "latex",
    caption = "Determination of the survival time probabilities",
    digits = 3) %>%
    footnote(general = paste("Removed – the sum of customers with dropout and that are",
                             "censored; Censored – the event did not occur during the",
                             "period of this data, collection; Risk of Dropout – ",
                             "number of customers at risk of, dropout; pi – survival",
                             "probability; Estimated Survival - months to survive in",
                             "the sports facility.", sep = " "),
                              threeparttable = TRUE)
```

Figure \@ref(fig:lifelinesplot) shows the Kaplan Meier survival curve customers 
considering the number of months of membership (x axis) and survival probability 
(y axis).
The customer dropout is very high in the first 12 months, ranging from a survival
probability of 54\% after the first 6 months until 24\% after 12 months.

```{python lifelinesplot, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Survival probabilities"}
import matplotlib.pyplot as plt
import matplotlib.font_manager as font_manager

plt.rcParams['figure.figsize'] = [12, 7]

fontL = font_manager.FontProperties(family='Times New Roman',weight=None,style='normal', size=10)

# fontAxis = {'family': 'Times New Roman'}

ax = kmf.plot(color='black')
ax.legend(prop=fontL)
#ax.set_xlabel('Months',fontdict=fontAxis,fontsize=12)
ax.set_xlabel('Months')
#ax.set_ylabel('Survival probability',fontdict=fontAxis,fontsize=12)
ax.set_ylabel('Survival probability')

ax.axvline(x=6,ymax=0.54,linestyle='--',color='black');ax.axhline(y=0.54,xmax=0.14,linestyle='--',color='black')
ax.annotate("0.54",xy=(6, 0.54), xytext=(7, 0.7))

ax.axvline(x=12,ymax=0.24,linestyle='--',color='black');ax.axhline(y=0.24,xmax=0.254,linestyle='--',color='black')
ax.annotate("0.24",xy=(12, 0.24), xytext=(12, 0.40))

ax.axvline(x=18,ymax=0.15,linestyle='--',color='black');ax.axhline(y=0.15,xmax=0.38,linestyle='--',color='black')
ax.annotate("0.15",xy=(18, 0.15), xytext=(18, 0.30))

plt.savefig(r.caminho_figuras+'fitnessLifelinesPlot.png', dpi=300)
plt.show()

plt.close()
```

Figure \@ref(fig:logrank) shows the survival by gender.
The survival curves by gender are very similar, both types of 
customers present a behavior that is not very different.

```{python logrank, results='hide', echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "Survival by gender"}
import matplotlib.pyplot as plt

ax = plt.subplot(111)

plt.rcParams['figure.figsize'] = [12, 7]

for gen in df_members['sex_1'].unique():
    ix = df_members['sex_1'] == gen
    kmf.fit(T.loc[ix], C.loc[ix], label=str(gen))
    ax = kmf.plot(ax=ax)

plt.savefig(r.caminho_figuras+'fitnessLogrankGender.png', dpi=300)

plt.show()
plt.close()
```

Figure \@ref(fig:logrankcfreq) shows the survival by contracted frequency.
Customers with contracted frequency of 6 and 4 times a week have higher survival
probabilities, against lower of customers with contracted frequencies of 7 and 
2 times a week.
Survival curves allow to explore tendencies related to survival to extract
actionable knowledge.

```{python logrankcfreq, results='hide', echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "Survival by contracted frequency"}
import matplotlib.pyplot as plt

ax = plt.subplot(111)

plt.rcParams['figure.figsize'] = [12, 7]

for item in df_members['cfreq'].unique():
    ix = df_members['cfreq'] == item
    kmf.fit(T.loc[ix], C.loc[ix], label=str(item))
    ax = kmf.plot(ax=ax)

plt.savefig(r.caminho_figuras+'fitnessLogrankCfreq.png', dpi=300)

plt.show()
plt.close()
```

```{python coxRegression, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE, fig.cap = "Cox regression"}
vars= ['age', 'dayswfreq', 'tbilled', 'maccess', 'freeuse', 'nentries',
       'cfreq', 'months', 'sex_1', 'dropout']
df_regression = df_members[vars].copy()

from lifelines import CoxPHFitter
cph = CoxPHFitter()
cph.fit(df_regression,duration_col='months',event_col='dropout')

cph.check_assumptions(df_regression)
```

The proportional hazard assumptions failed in the following variables:
age p<0.01, cfreq p<0.01, dayswfreq p<0.01, tbilled p<0.01, freeuse 
p<0.01, nentries p<0.01.

## Survival trees

To evaluate the performance of the random survival forest we have calculated
the concordance probability (C-index), IBS and Mean Absolute Error (MAE).
The IBS presents an accuracy along the 12 months of 0.08 (figure \@ref(fig:createModel1)).

```{python createModel1, results='hide', echo=FALSE, message = FALSE, warning = FALSE, fig.cap = "Model performance"}
from pysurvival.models.survival_forest import RandomSurvivalForestModel
from sklearn.model_selection import train_test_split
from pysurvival.utils.metrics import concordance_index
from pysurvival.utils.display import integrated_brier_score
from pysurvival.utils.display import compare_to_actual

X = df_members.copy()
t = df_members['months']
e = df_members['dropout']
X.drop(axis=1, columns=['months','dropout'],inplace=True)

X_train, X_test, t_train, t_test, e_train, e_test = train_test_split(X, t, e, test_size=0.3, random_state=0)

# Fitting the model
csf = RandomSurvivalForestModel(num_trees=20)
csf.fit(X_train, t_train, e_train, max_features='sqrt', max_depth=5, min_node_size=20, seed = 1)

c_index = concordance_index(csf, X_test, t_test, e_test)
ibs = integrated_brier_score(csf, X_test, t_test, e_test, t_max=12, figure_size=(12,5))
#plt.savefig(r.caminho_figuras+'create_model1_fitness.png', dpi=300)
```

The actual versus predicted model presents the actual and predicted customers which dropout 
during the 40 months, which as an average absolute error of 7.5 customers (figure \@ref(fig:createModel2)).

```{python createModel2, results='hide', echo=FALSE, message = FALSE, warning = FALSE, fig.cap = "Conditional survival forest"}

results = compare_to_actual(csf, X_test, t_test, e_test, is_at_risk = False,  figure_size=(12, 6), metrics = ['rmse', 'mean', 'median'])
```

```{r previsaoglobal, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
# gravar a figura com a previsão global
#ggsave(file="/mnt/c/nuvem/Dropbox/doutoramento/tese/9.tese_documento/PhD-Pedro-Sobreiro/figures/previsao_global.pdf",device="pdf")

```

Table \@ref(tab:summarytable2) shows features importance calculated according [@Breiman_2001],
where the percent increase in misclassification rate as compared to the out-of-bag rate (with all 
variables intact), out-of-bag is a bootstrap aggregating (subsampling with replacement to create 
training samples for the model to learn from) where two independent sets are created. 
One set, the bootstrap sample, data chosen to be "in-the-bag" by sampling with replacement and the 
out-of-bag is all data not chosen in the sampling process.
The most important variable is the $dayswfreq$, followed by $tbilled$ and
$nentries$, compared with the $cfreq$, $age$, and $sex$.

```{r summarytable2, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
tbl <- py$csf$variable_importance_table
kbl(tbl, booktabs = T, caption = "Features importance in the survival model")
```

The prediction is very similar to the actual value. 
The model accuracy is very high with a root mean square error of
`r round(py$results$root_mean_squared_error)`.
The mean absolute error mean was `r round(py$results$mean_absolute_error,2)` customers,
and the median absolute error was `r round(py$results$median_absolute_error,2)`.

## Survival trees based model with clusters

In this approach we have created clusters and applied the survival trees within 
each cluster.
The determination of the clusters using the BIC criterion where the EEV model: 7 clusters -57159.24; 
6 clusters -63937.59; and 4 clusters -77088.81 figure \@ref(fig:clusters) shows the 
determination of the number of clusters using BIC, also the elbow analysis available in 
figure \@ref(fig:elbowCalculation). 
An optimal number of clusters was considered of five. 
Considering that was the value after the average distortion was flattened.

```{r clusters, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Analysis number of clusters"}
library(mclust)
#y <- scale(py$df_members)
y <- scale(py$df_members)

set.seed(0) # to make reproducible

#bic <- mclustBIC(y,modelNames = c("VEI"))
bic <- mclustBIC(y)
# Best model using the BIC criteria
#bic
plot(bic, what = "BIC")

```

```{r NbClust, echo = FALSE, message = FALSE, warning = FALSE, eval=FALSE}
library(NbClust)
nb <- NbClust(y, diss = NULL, distance = "euclidean",
              min.nc = 2, max.nc = 5, method = "kmeans",
              index = "all", alphaBeale = 0.1)
hist(nb$Best.nc[1, ], breaks = max(na.omit(nb$Best.nc[1, ])))

```

```{python elbowCalculation, results='hide', echo=FALSE, message=FALSE, warning = FALSE, fig.cap="Elbow analysis"}
from sklearn.cluster import KMeans
from sklearn import preprocessing
from scipy.spatial.distance import cdist
import matplotlib.pyplot as plt
import numpy as np

#Finding optimal no. of clusters
clusters=range(1,20)
meanDistortions=[]

for k in clusters:
    model=KMeans(n_clusters=k)
    model.fit(df_members)
    prediction=model.predict(df_members)
    meanDistortions.append(sum(np.min(cdist(df_members, model.cluster_centers_, 'euclidean'), axis=1)) / df_members.shape[0])

plt.plot(clusters, meanDistortions, 'bx-')
plt.xlabel('k')
plt.ylabel('Average distortion')
plt.show();

```

The calculation of the clusters to each member in the dataset was developed considering,
7,6 and 4 clusters. 
However after the executions KMeans only performed with 3 clusters.

```{python kmeanscalculation, results='hide', echo=FALSE, message=FALSE, warning=FALSE}
# we are going to use random_state = 0 for the centroid
# initialization being deterministic allowing a better
# reproducibility
cluster = KMeans(n_clusters=3, random_state=0)

cluster.fit(df_members)
df_members['cluster']=cluster.predict(df_members)
#print(df_members.cluster.value_counts());
```

```{python kmeans PCA, echo=FALSE, message = FALSE, warning = FALSE}
from sklearn.decomposition import PCA

plt.rcParams['figure.figsize'] = [13, 4]

pca=PCA(n_components=2)
df_members['X_pca']=pca.fit_transform(df_members)[:,0] #todas as linhas da primeira coluna com redução
df_members['Y_pca']=pca.fit_transform(df_members)[:,1] #todas as linhas da segunda coluna com redução

fig, ax = plt.subplots()
for cluster in df_members.cluster.unique():
    x = df_members['X_pca'].loc[df_members.cluster == cluster]
    y = df_members['Y_pca'].loc[df_members.cluster == cluster]
    ax.scatter(x, y, label=cluster, alpha=1, edgecolors='none');

ax.legend()
ax.grid(True)
ax.set_title('Customers clusters');
plt.show();
```

The model accuracy is very high in the first years. The prediction is very
similar to the actual value. The absolute error mean of
6 customers.

```{python createModelClusters, results='hide', include=TRUE, echo=FALSE, message = FALSE, warning=FALSE}
# Building training and testing sets for the clusters

df_resultados = pd.DataFrame(columns=['cluster','rmse','mean','median'])

for cluster in df_members.cluster.unique():
    # Number of samples in the dataset
    df_members_cluster = df_members[df_members.cluster == cluster].copy()
    X = df_members_cluster.copy()
    t = df_members_cluster['months']
    e = df_members_cluster['dropout']
    X.drop(axis=1, columns=['months','dropout'], inplace=True)
    X_train, X_test, t_train, t_test, e_train, e_test = train_test_split(X, t, 
                                                        e, random_state=0)
    # Fitting the model
    csf = RandomSurvivalForestModel(num_trees=20)
    csf.fit(X_train, t_train, e_train, max_features='sqrt', max_depth=5, 
            min_node_size=20, seed = 1)

    results_cluster = compare_to_actual(csf, X_test, t_test, e_test, 
                                is_at_risk = False, figure_size=(12, 6), 
                                metrics = ['rmse', 'mean', 'median'])

    #print(f"The cluster {cluster} as a size of {X.shape[0]}")
    #print(results_cluster)
    #print(f"Cluster {cluster} RMSE {results_cluster['root_mean_squared_error']}")
    #print(f"Cluster {cluster} MAE {results_cluster['median_absolute_error']}")
    #print(f"Cluster {cluster} MAError {results_cluster['mean_absolute_error']}")

    df_resultados = df_resultados.append({'cluster' : cluster, 
                                          'rmse' : results_cluster['root_mean_squared_error'], 
                                          'mean' : results_cluster['median_absolute_error'], 
                                          'median' : results_cluster['mean_absolute_error']}, 
                                          ignore_index = True)
```

The performance of the cluster 1 the IBS presents an accuracy of 0.06 
(figure \@ref(fig:createModelCluster1IBS)) along all time.
The actual versus predicted model presents the actual and predicted customers which dropout during
the 40 months, which as an mean absolute error of 1.5 customers, the mean median absolute
error was 0.61 and the Root Mean Square Error of 2.8 (figure \@ref(fig:createModelCluster1SF)).

The features importance in the survival model cluster 1 (table \@ref(tab:summarytableCluster1))
identify the three most relevant features to predict survival $maccess$, 
$tbilled$, and $dayswfreq$. The features with lower relevance were $freeuse$, $sex$ and $cfreq$.

```{python createModelCluster1IBS, results='hide', echo=FALSE, message = FALSE, warning = FALSE, fig.cap = "Model performance cluster 1"}
df_members_cluster = df_members[df_members.cluster == 0].copy()
X = df_members_cluster.copy()
t = df_members_cluster['months']
e = df_members_cluster['dropout']
X.drop(axis=1, columns=['months','dropout','X_pca','Y_pca','cluster'], inplace=True)
X_train, X_test, t_train, t_test, e_train, e_test = train_test_split(X, t, 
                                                    e, random_state=0)
# Fitting the model
csf = RandomSurvivalForestModel(num_trees=20)
csf.fit(X_train, t_train, e_train, max_features='sqrt', max_depth=5, 
        min_node_size=20, seed = 1)

c_index = concordance_index(csf, X_test, t_test, e_test)
ibs = integrated_brier_score(csf, X_test, t_test, e_test, t_max=12, 
                             figure_size=(12,5))

```

```{python createModelCluster1SF, results='hide', echo=FALSE, message = FALSE, warning = FALSE, fig.cap = "Conditional survival forest cluster 1"}
compare_to_actual(csf, X_test, t_test, e_test, is_at_risk = False, figure_size=(12, 6), 
                  metrics = ['rmse', 'mean', 'median'])

```

```{r summarytableCluster1, echo=FALSE, message=FALSE, warning=FALSE}
tbl <- py$csf$variable_importance_table
kbl(tbl, booktabs = T,
    caption = "Features importance in the survival model with cluster 1")
```

The performance of the cluster 2 the IBS presents an accuracy along time 0.09 
(figure \@ref(fig:createModelCluster2IBS)) along all time.
The actual versus predicted model presents the actual and predicted customers which dropout during
the 40 months, which as an mean absolute error of 6.5 customers, the mean median absolute
error was 2.3 and the Root Mean Square Error of 11.79 (figure \@ref(fig:createModelCluster2SF)).
The features importance in the survival model cluster 2 (table \@ref(tab:summarytableCluster2))
identify the three most relevant features to predict survival $dayswfreq$, 
$tbilled$, and $freeuse$. The least relevant were $nentries$, $cfreq$, and $age$.

```{python createModelCluster2IBS, results='hide', echo=FALSE, message = FALSE, warning = FALSE, fig.cap = "Model performance cluster 2"}
df_members_cluster = df_members[df_members.cluster == 1].copy()
X = df_members_cluster.copy()
t = df_members_cluster['months']
e = df_members_cluster['dropout']
X.drop(axis=1, columns=['months','dropout','X_pca','Y_pca','cluster'], inplace=True)
X_train, X_test, t_train, t_test, e_train, e_test = train_test_split(X, t, 
                                                    e, random_state=0)
# Fitting the model
csf = RandomSurvivalForestModel(num_trees=20)
csf.fit(X_train, t_train, e_train, max_features='sqrt', max_depth=5, 
        min_node_size=20, seed = 1)

c_index = concordance_index(csf, X_test, t_test, e_test)
ibs = integrated_brier_score(csf, X_test, t_test, e_test, t_max=12, 
                             figure_size=(12,5))

```

```{python createModelCluster2SF, results='hide', echo=FALSE, message = FALSE, warning = FALSE, fig.cap = "Conditional survival forest cluster 2"}
compare_to_actual(csf, X_test, t_test, e_test, is_at_risk = False, figure_size=(12, 6), 
                  metrics = ['rmse', 'mean', 'median'])

```

The performance of the cluster 3 the IBS presents an accuracy along time 0.01 
(figure \@ref(fig:createModelCluster3IBS)) along all time.
The actual versus predicted model presents the actual and predicted customers which dropout during
the 40 months, which as an mean absolute error of 1.5 customers, the mean median absolute
error was 1.2 and the Root Mean Square Error of 2.08 (figure \@ref(fig:createModelCluster3SF)).
The features importance in the survival model cluster 3 (table \@ref(tab:summarytableCluster3))
identify the three most relevant features to predict survival $tbilled$, 
$dayswfreq$, and $nentries$. 
The least relevant were $sex$, $age$, and $cfreq$.

```{r summarytableCluster2, echo=FALSE, message=FALSE, warning=FALSE}
tbl <- py$csf$variable_importance_table
kbl(tbl, booktabs = T,
    caption = "Features importance in the survival model with cluster 2")
```

```{python createModelCluster3IBS, results='hide', echo=FALSE, message = FALSE, warning = FALSE, fig.cap = "Model performance cluster 3"}
df_members_cluster = df_members[df_members.cluster == 2].copy()
X = df_members_cluster.copy()
t = df_members_cluster['months']
e = df_members_cluster['dropout']
X.drop(axis=1, columns=['months','dropout','X_pca','Y_pca','cluster'], inplace=True)
X_train, X_test, t_train, t_test, e_train, e_test = train_test_split(X, t, 
                                                    e, random_state=0)
# Fitting the model
csf = RandomSurvivalForestModel(num_trees=20)
csf.fit(X_train, t_train, e_train, max_features='sqrt', max_depth=5, 
        min_node_size=20, seed = 1)

c_index = concordance_index(csf, X_test, t_test, e_test)
ibs = integrated_brier_score(csf, X_test, t_test, e_test, t_max=12, 
                             figure_size=(12,5))

```

```{python createModelCluster3SF, results='hide', echo=FALSE, message = FALSE, warning = FALSE, fig.cap = "Conditional survival forest cluster 3"}
compare_to_actual(csf, X_test, t_test, e_test, is_at_risk = False, figure_size=(12, 6), 
                  metrics = ['rmse', 'mean', 'median'])

```

The features importance in the survival model (table \@ref(tab:summarytableCluster3))
identify the three most relevant features to predict survival $dayswfreq$, 
$tbilled$, and $nentries$.

```{r summarytableCluster3, echo=FALSE, message=FALSE, warning=FALSE}
tbl <- py$csf$variable_importance_table
kbl(tbl, booktabs = T,
    caption = "Features importance in the survival model with cluster 3")
```

### Model Comparison

Table \@ref(tab:summarytable3) shows the performance of both approaches, with or without clusters. 
The RMSE in the clusters 1, 2 and 3 is lower than not using clusters to predict dropout.
Overall the performance improved. The performance is also better using mean and median.

```{r summarytable3, echo=FALSE, message=FALSE, warning=FALSE}
tbl <- py$df_resultados
# append results without clusters
tbl <- rbind(tbl, c("w/cluster",
                   py$results$root_mean_squared,
                   py$results$median_absolute_error,
                   py$results$mean_absolute_error))
kbl(tbl, booktabs = T,
    caption = "Performance of prediction in each cluster", digits = 3)
```

The model accuracy without clusters is very high with a root mean
square error of 13, the mean absolute error mean was 7.53 customers, and the median 
absolute error was 4.04.
The model using clusters had an mean absolute error of 1.5 customers, the mean median absolute
error was 1.2 and the Root Mean Square Error of 2.08.
The performance using clusters improved significantly.

# Conclusion

This paper investigated the customer dropout in a Health Club organization, using
a dynamic perspective that the dropout risk varies along the time.
Exploring two approaches, using a survival model based on random forests 
with or without clusters.
The model using clusters allowed to combine the customers in different clusters,
an hybrid approach.
Based on this performance the proposed model using clusters allows to improve 
the accuracy on the survival model allowing to target approaches considering
the timing when the dropout occurs, considering the clusters where the customer is.
Is very important for managers use this information to improve their retention
strategies.

<!--

```{=tex}
\begin{itemize}
    \item RQ1: What is the current state of the research being developed?
    \item RQ2: What algorithms have been used to predict dropout?
    \item RQ3: What are the features used to predict dropout?
    \item RQ4: When does dropout occur?
    \item RQ5: How is the accuracy of the machine learning algorithms in predicting dropout measured?
\end{itemize}
```
From RQ1, it was possible to identify some business areas that are
under-researched, such as the energy sector, education, logistics and
hospitality. Compared to other business areas such telecom or the financial
sector, research on the energy sector or water supply is lacking, considering
the contractual settings that are assumed to provide such types of services.
Considering the business model of many software companies as software as a
services (SaS), the number of research works is also surprisingly low.

RQ2 also provided an overall perspective related to the algorithms being used to
predict customer dropout. The first could be the importance and wider adoption
of decision trees and random forests [@antipov_applying_2010;
@benoit_improving_2012; @burez_crm_2007], and logistic regression
[@coussement_improved_2010], which could be due to its higher interpretability
and flexibility [@keramati_improved_2014]. Interpretability is an important
aspect for the marketing department in the extraction of valuable information
from the model to develop effective retention strategies [@verbeke_new_2012].
The problem arises in the balancing between interpretability and the higher
performance of the algorithms inspired by nature (such as neural networks). From
a business perspective, dropout prediction should also be considered as a
business objective, which requires more than predicting if the customer will
churn or not [@devriendt_why_2019], where higher interpretability provides
better support in the development of retention strategies. The developed SLR
also raises the possibility of integrating different algorithms using ensemble
methods or integrating several models using a hybrid approach. None of the
studies integrated the survival approach to predict customer dropout, for
example, using a hybrid approach.

It is considered positive if actions are developed to retain customers, but the
problems should also be considered, such as the following: (1) customers who
have greater risk of dropout should be targeted to provide a base for a better
ROI in the retention strategies [@xie_customer_2009; @coussement_churn_2008] and
(2) the retention strategies should be developed focused on customers with
higher satisfaction, or its inclusion could be a reminder of the contractual
agreement nearing an end and could lead to churn [@devriendt_why_2019].

From RQ3, several types of features being used were able to be identified, such
as demographic, behavioral, and economic indicators, pictorial data, network
relationships or high cardinality features. The problem that arises is that some
studies used data and features that were not described, and this creates a major
issue, How can reproducibility be developed in a study without the availability
of the data or the identification of the features used? Considering that science
is driven by data, with the development of new technologies, the increasing
complexity of research and the amount of data collected, the challenge is to
ensure that research is available to all [@Hanson_Sugden_Alberts_2011]; this
requires both availability of the data and the algorithms so that they can be
explored by other researchers. The features are selected mainly to verify the
performance of the models, and are essential to performance prediction,
accuracy, and the steps for processing the data, which are fundamental to
improve the model accuracy [@azeem_churn_2017].

There are several challenges around the timing related to dropout, or
considering the dynamic behavior of the customer in the intent to drop out
[@alboukaey_dynamic_2020]. The importance of understanding when dropout will
occur and the risk when discarding the temporal perspective of the problem seems
to be an element that should be addressed. Few studies considered this
[@perianez_churn_2016; @burez_separating_2008]. This shows an opportunity to
address the importance of the timeframe and its influence on the efficiency of
the model.

According to each business model, the timeframe could be addressed considering
the survival probability according to the customer relationship age, and dropout
predictions could be developed according to these survival probabilities, as
suggested by @esteves_churn_2016, to investigate which data timeframe produces
the best result and how the efficiency of the models is influenced by this
timeframe. Exploring the duration of the relation and the understanding of the
features that increase or decrease that duration seems to be an important
approach that could complement the existing approaches to predicting dropout.

From RQ5, the literature analysis showed that different types of questions
arise. Which are the best approaches to develop the analysis of the performance
in predicting dropout? Several metrics are customer dropout is to improve the
performance of organizations in retaining customers, which is a management
problem in which data mining is adopted [@verbeke_new_2012]. The goals of the
model should be formulated considering the context of the problem that is being
addressed; in marketing retention strategies, the up-lift supports the
development of proactive actions to minimize the investment in retention
strategies [@coussement_churn_2008]. Some assumptions that underlie the
adoption of uplift metrics consider that customers with a higher risk of
churning could not be the best targets, as suggested by @ascarza_retention_2018.
Other researchers addressed the problem using the top-decile lift to develop
more proactive actions to retain the customers at risk of churning
[@coussement_churn_2008;@xie_customer_2009]. This approach considers the 10% of
customers with more risk, and investments in retention strategies should be
developed that distinguish the churners susceptible to marketing actions from
those who will leave anyway [@coussement_comparative_2017]. Although uplift
models seem to be good strategies, they should also used, such as AUC,
sensitivity, specificity, recall, precision, and F-score. However, the goal
ofconsider factors other than risk and customer satisfaction, as not taking this
into consideration could be counterproductive and the model should be removed
from the retention strategy.

The true business objective is to reduce customer churn. Customers who are about
to churn but cannot be retained should be excluded from the campaign, as
targeting them will be a waste of scarce resources [@devriendt_why_2019]. Using
these models seem to be a good strategy, as they can outperform predictive
models that consider only accuracy from a profitability busshould be considered
that customers with a higher risk of churning may not be the best targets to
develop retention strategies. Those perspectives entail the dropout.

that a business context, or the clarification of a business objective underlying
the prediction of customer dropout, should be developed, to clarify which
objectives should be achieved before employing the profitability of reducing g
machine learning algorithms. Surprisingly, the analyzed studies did not address
the customer lifetime value as an objective to optimize considerininess
perspective.

# Aspects to consider

-   Interpretability from RQ2
-   The business objective is to increase the number of members and organization
    profits
-   piping several algorithms to improve accuracy. Aka hybrid approach
-   @Alboukaey_dynamic_2020 proposes ...
-   grep the articles addressing hybrid: pdfgrep -ri "hybrid.{1,10} approach"

# Results

In this section, we present our experiments to validate the proposed models,
comparing against other approaches. The models where optimized using the
hyper-parameters Grid Search technique. The explored hyper-parameters and the
best values of these parameters for every model are listed in (Table 3).

```{=latex}
\begin{table}[ht]
    \footnotesize
    \centering
    \begin{tabular}{p{0.15\textwidth}p{0.55\textwidth}p{0.25\textwidth}}
    \hline
    \textbf{Model name} & \textbf{Explored parameters values} & \textbf{Best parameters} \\
    \hline
    Survival trees  & pysurvival random forest                  & a \\
    Survival trees
    with clusters   & pysurvival random forest with clusters    & a \\
    Scikit survival 
    trees           & scikit survival                           & a \\
    Scikit survival 
    with clusters   & scikit with clusters                      & a \\
    Scikit survival 
    gradient boost  & scikit survival gradient boost            & a \\
    Scikit survival 
    gradient boost 
    with clusters   & scikit with clusters                      & a \\
    \hline
    \end{tabular}
    \caption{Hyper-parameters best values}
    \label{hyperparametersbestvalues}
\end{table}
```
The results of the performance of the models are available in table 4. Colocar o
modelo. Resultados do modelo

```{=latex}
\begin{table}[ht]
    \footnotesize
    \centering
    \begin{tabular}{p{0.15\textwidth}p{0.35\textwidth}r p{0.10\textwidth}}
    \hline
    \textbf{Model name} & \textbf{Results} & \textbf{n}\\
    \hline
    Survival trees  & RMSE 57.815                                     & 25316\\
                    & MAE 18.966                                      & \\   
                    & MEAE 38.557                                     & \\
    Survival trees  & Cluster 0: RMSE  2.785 MAE  0.000 MEAE  0.899   &   930\\
    with clusters   & Cluster 1: RMSE 78.141 MAE 28.537 MEAE 56.989   & 17067\\
                    & Cluster 2: RMSE 15.081 MAE  1.995 MEAE  6.850   &  2423\\
                    & Cluster 3: RMSE  9.354 MAE  2.506 MEAE  6.427   &  2816\\
                    & Cluster 4: RMSE  8.805 MAE  1.349 MEAE  3.725   &  2080\\

    Scikit survival 
    trees           & scikit survival                                 & \\
    Scikit survival 
    with clusters   & scikit with clusters                            & \\
    Scikit survival 
    gradient boost  & scikit survival gradient boost                  & \\
    Scikit survival 
    gradient boost 
    with clusters   & scikit with clusters                            & \\
    \hline
    \end{tabular}
    \caption{Hyper-parameters best values}
    \label{hyperparametersbestvalues2}
    * Root Mean Square Error (RMSE), Mean Absolute Error (MAE), Median Absolute Error (MEAE)  
\end{table}
```

# Conclusion

Article Ascarza

-   Retention Futility: Targeting High-Risk Customers Might be Ineffective
    [@ascarza_retention_2018]

Ascarza, E. (2018). Retention Futility: Targeting High-Risk Customers Might be
Ineffective. Journal of Marketing Research, 55(1), 80-98. sim.
<https://doi.org/10.1509/jmr.16.0163>

Example of Developed actions place in the discussion:

``` markdown
Each month, the company identified the customers who were up for renewal and
split them (randomly and evenly) between a treatment group that received a 
"thank you" gift with the letter and a control group that received only the 
renewal latter.
```
-->

# References {.unnumbered}

::: {#refs}
:::

# Appendix: Chunk options {.unnumbered}

## Software versioning R

### R

```{r fig-versioning, echo = TRUE}
cat(paste("#", capture.output(sessionInfo()), "\n", collapse  = ""))
  # or use message() instead of cat()
```

### 2024-01-11 article/customer_fitness_artigo.tex
update

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}
\title{\hfill\break
\hfill\break
\hfill\break
\vspace{1cm}Health club customer dropout membership\footnote{Corresponding address: \href{mailto:sobreiro@esdrm.ipsantarem.pt}{\nolinkurl{sobreiro@esdrm.ipsantarem.pt}}. Quality of Life Research Centre, Polytechnic Institute of Santarém, Portugal}\vspace{0.5cm}}
\author{Pedro Sobreiro, Javier Berrocal, Domingos Martinho, José Garcia Alonso\\
\strut \\
Extremadura University\\}
\date{\hfill\break
\hfill\break
21 julho, 2022\\
\strut \\}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmainfont[]{Times New Roman}
  \setsansfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin = 1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{dcolumn}
\usepackage{color}
\usepackage{pdfpages}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\begin{document}
\maketitle
\begin{abstract}
\vspace{1cm}

Customer retention is fundamental to improve the organizations profits. Existing approaches normaly explore static models predicting when the customer will dropout. In this paper we propose a different approach considering that the customer dropout risk changes over time using clusters to improve the model performance. We explore a survival model using random forests with and with out clusters in a dataset of 5209 customers. The model using clusters improved the performance significantly. This paper shows that the use of clusters should be considered to identify dropout patterns to support the timming when the dropout occurs considering the cluster where the customer is. Improving the performance as less errors and is expected to contribute to the identification when should be developed retention strategies. \vspace{.8cm}
\end{abstract}

\setstretch{1.2}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Customer retention is a problem is a problem being addressed using the dropout
prediction as an insight to identify customers that could dropout.
The customers database is the most valuable asset that the organizations possess
(\protect\hyperlink{ref-Athanassopoulos_2000}{Athanassopoulos 2000}).
The problem of retention is not new, in a seminal paper \protect\hyperlink{ref-Copeland_1923}{Copeland} (\protect\hyperlink{ref-Copeland_1923}{1923}) address
the problem of brand loyalty and in marketing research (\protect\hyperlink{ref-Mellens_Dekimpe_Steenkamp_1996}{Mellens, Dekimpe, and Steenkamp 1996}).

The advantage of developing some strategies in retention are supported in the idea
that the costs of customer retention are lower than customer acquisition
(\protect\hyperlink{ref-Edward_Sahadev_2011}{Edward and Sahadev 2011}; \protect\hyperlink{ref-Fornell_Wernerfelt_1987}{Fornell and Wernerfelt 1987}).
The increase in the profits with the reduction of 5\% of the dropout could represent
almost a duplication of the profits (\protect\hyperlink{ref-Reichheld_1996}{Reichheld 1996}).

The development of a customer retention strategy could be supported in the
identification of the customers that will dropout (\protect\hyperlink{ref-alboukaey_dynamic_2020}{Alboukaey, Joukhadar, and Ghneim 2020a}).
However, dropout has two underlying scenarios contractual and non-contractual
settings \{\protect\hyperlink{ref-Gupta2006Modeling}{Gupta et al.} (\protect\hyperlink{ref-Gupta2006Modeling}{2006}); \protect\hyperlink{ref-ascarza_retention_2018}{Ascarza} (\protect\hyperlink{ref-ascarza_retention_2018}{2018})\},
in a contractual business the customer needs to renew their contracts to continue its usage
(\protect\hyperlink{ref-Ascarza_Hardie_2013}{Ascarza and Hardie 2013}), against non contractual where the firm has to infer if
the customer is still active
However, in contractual settings the customer dropout represents an explicit
ending of a relationship which is more penalizing than non contractual settings
(\protect\hyperlink{ref-risselada_staying_2010}{Risselada, Verhoef, and Bijmolt 2010}).
This has implications to the profitability of the organizations increasing marketing
costs and reducing sales (\protect\hyperlink{ref-amin_customer_2017}{Amin et al. 2017}).

The anticipation of the dropout allows the development of countermeasures to
reduce customer churn. Several studies address the problem related to customer
retention trying to improve the profitability (\protect\hyperlink{ref-coussement_improving_2009}{Coussement and Van den Poel 2009}; \protect\hyperlink{ref-devriendt_why_2019}{Devriendt, Berrevoets, and Verbeke 2019}; \protect\hyperlink{ref-garcia_intelligent_2017}{García, Nebot, and Vellido 2017}).
Existing organizations are addressing this problem by shifting their target from
capturing new customers to preserving existing ones (\protect\hyperlink{ref-garcia_intelligent_2017}{García et al. 2017}),
considering that investments in retention strategies are more profitable
than acquiring new customers (\protect\hyperlink{ref-coussement_improving_2009}{Coussement and Van den Poel 2009}).

The approaches normally employed use a dependent variable representing dropout
or non-dropout, without considering a dynamic perspective that the dropout risk
changes over time (\protect\hyperlink{ref-Alboukaey_dynamic_2020}{Alboukaey, Joukhadar, and Ghneim 2020b}). The survival models try to solve
this limitation (\protect\hyperlink{ref-routh_estimating_2020}{Routh, Roy, and Meyer 2020}) capturing a temporal dimension of the
customer dropout (\protect\hyperlink{ref-perianez_churn_2016}{Perianez et al. 2016}). \protect\hyperlink{ref-perianez_churn_2016}{Perianez et al.} (\protect\hyperlink{ref-perianez_churn_2016}{2016}) used survival
analysis to predict also when the dropout will occur.

Other studies proposed also the integration of several algorithms to improve the
performance in the prediction of the dropout such the usage of clusters combined
with churn prediction (\protect\hyperlink{ref-gok_case_2015}{Gök, Özyer, and Jida 2015}; \protect\hyperlink{ref-hung_applying_2006}{Hung, Yen, and Wang 2006}; \protect\hyperlink{ref-vijaya_sivasankar_2019}{Vijaya and Sivasankar 2019}). The approach relies in the assumption that combining
the customers in different clusters allows the improvement of the prediction
accuracy. \protect\hyperlink{ref-vijaya_sivasankar_2019}{Vijaya and Sivasankar} (\protect\hyperlink{ref-vijaya_sivasankar_2019}{2019}) suggested the adoption hybrid models combining
more than one classier are achieving increased performance compared to those
using single classifiers.

There are several challenges around the timing related to dropout, or
considering the dynamic behavior of the customer in the intent to drop out
(\protect\hyperlink{ref-alboukaey_dynamic_2020}{Alboukaey et al. 2020a}). The importance of understanding when dropout will
occur and the risk when discarding the temporal perspective of the problem seems
to be an element that should be addressed. Few studies considered this
(\protect\hyperlink{ref-burez_separating_2008}{Burez and Vandenpoel 2008}; \protect\hyperlink{ref-perianez_churn_2016}{Perianez et al. 2016}). This shows an opportunity to
address the importance of the timeframe and its influence on the efficiency of
the model and also evalute if the combination of clusters could improve the
performance.

Survival analysis, which origin as stands in biomedical statistics, its especially
well-suited to studying the timing of events in longitudinal data (\protect\hyperlink{ref-Singer_Willett_1993}{Singer and Willett 1993}).
Survival analysis is a class of statistical methods modelling the occurrence and timing of
an event, such as the customer dropout.
Survival analysis allow us to examine not only if an event occurred but also how long it
took to occur.
A primary value of survival analysis, however, is to compare dropout probability for
individuals classified with theoretically relevant variables.
The survival methods have enjoyed and increasing popularity in several disciplines ranging
from medicine to economics (\protect\hyperlink{ref-Singer_Willett_1993}{Singer and Willett 1993}).

Random Survival Forests does not make the proportional hazards assumption
(\protect\hyperlink{ref-Ehrlinger_2016}{Ehrlinger 2016}) and has the flexibility to model survivor curves that are of
dissimilar shapes for contrasting groups of subjects. Random Survival Forest is
an extension of Random Forest allowing efficient non-parametric analysis of time
to event data (\protect\hyperlink{ref-Breiman_2001}{Breiman 2001}). This characteristics allow us to surpass the Cox
Regression limitation of the proportional hazard assumption, requiring to
exclude variables which not fulfill the model assumption. It was shown by
\protect\hyperlink{ref-Breiman_2001}{Breiman} (\protect\hyperlink{ref-Breiman_2001}{2001}) that ensemble learning can be further improved by injecting
randomization into the base learning process - a method called Random Forests.

Some studies employed also a combination of clusters with the churn prediction
(\protect\hyperlink{ref-gok_case_2015}{Gök et al. 2015}; \protect\hyperlink{ref-hung_applying_2006}{Hung et al. 2006}; \protect\hyperlink{ref-vijaya_sivasankar_2019}{Vijaya and Sivasankar 2019}), where the customers
where grouped in clusters to improve the accuracy within each cluster.
Clusters are approaches using unsupervised algorithms to group elements
with similar characteristics.
This unsupervised method has been widely used employing approaches
such Hierarchical Clustering (\protect\hyperlink{ref-Saunders_1980}{Saunders 1980}), K-Means (\protect\hyperlink{ref-vijaya_sivasankar_2019}{Vijaya and Sivasankar 2019}), or
Random Forest Clustering (\protect\hyperlink{ref-Breiman_2001}{Breiman 2001}).
\protect\hyperlink{ref-JafariMarandi_Denton_Idris_Smith_Keramati_2020}{Jafari-Marandi et al.} (\protect\hyperlink{ref-JafariMarandi_Denton_Idris_Smith_Keramati_2020}{2020}) explored an approach combining
clustering methods in parallel to a classification.

In this study, we investigate if an hybrid approach using clusters and random survival forests
which have never been used in to predict membership in a health club using existing data
improves the prediction accuracy.
Our paper is organized as follows. In the next section, we address our research methodology,
using survival analysis to identify the survival probability along the time, how
we address the problem of determining the clusters and the performance of the model
to predict customer survival within each identified cluster.

\hypertarget{survival-analysis}{%
\section{Survival analysis}\label{survival-analysis}}

Survival analysis focus in the analysis of the time until an event of interest,
and exploring its relationship with different factors.
The main advantage is related to the concept of censoring, indicating
that observations that are not complete related to the event of interest, e.g.
customers that didn't dropout yet, which are incorporated in the analysis.
This means that there are customers still active for which we don't know if
the event of dropout has occurred, which is called censorship.
Survival models take censoring into account and incorporate this uncertainty,
instead of predicting the time of event such in regression models, the survival models
allow to predict the probability of an event happens at a particular time.

The time of dropout is represented by T, which is a non-negative random variable,
indicating the time period of the event occurring for a randomly selected individual
from the population, representing the probability of an event to occur each time period given
that has not already occurred in a previous time period, known as discrete-time hazard function
(\protect\hyperlink{ref-Singer_Willett_1993}{Singer and Willett 1993}).
The survival function represents the probability of an individual surviving after time t,
S(t) = P(T \(>\) t), t \(\geq\) 0, with the properties S(0) = 1,S(\(\infty\)) = 0.
The distribution function is represented with F, defined as F(t) = P(T \(\leq\) 0), for t \(\geq\) 0.
The function of probability density represented with f where:

\begin{equation}
f(t) = \lim_{dt{\to 0}} \frac{P[t \le T < t + dt ]}{dt}
\end{equation}

\(f(t)dt\) represents the probability of an event occurring in the moment t.
The need to represent the distribution evolution of the death probability along the time,
uses to the hazard function, represented as:

\begin{equation}
\lambda(t) = \lim_{dt \to 0} \frac{P[t \le T < t + dt | T \ge t ]}{dt}
\end{equation}

The determination of the survival curves is based in the following elements:
(1) the total value of observations removed during the time period (e.g.~days, months or years),
either by dropout or by censorship;
(2) observations that composed the sample of the study;
(3) customers who had not yet dropped out at any given time.
The survival probability until the time period ii (\(p_i\)) is calculated with:

\begin{equation}
p_i=\frac{r_i-d_i}{r_i}
\end{equation}

Where \(r_i\) is the number of individuals that survived at the beginning of the period,
\(d_i\) the number of individuals who left during the period.
The survival time estimate was also taken considering the month in which it is found (estimated).
Cox's allow test difference between survival times.
The advantage in using survival analysis was that allow us to detect if the risk of an event differs
systematically across different people, using specific predictors.
The coefficients in a Cox regression were related to the hazard, where a positive value
represents a worse prognosis and the opposite, negative value a better prognosis.
The advantage of survival analysis was that allow us to include information of covariates
that were censored up to the censoring event.

The Cox PH model assumes the covariates to be time independent, for example
gender and age when where retrieved do not change over time (\protect\hyperlink{ref-Schober2018Survival}{Schober and Vetter 2018})
Because the Cox model requires the hazards in both groups to be proportional,
researchers are often asked to ``test'' whether hazards are proportional
(\protect\hyperlink{ref-Stensrud_Hernuxe1n_2020}{Stensrud and Hernán 2020}).
Considering this we explored other approach that allow us to develop this analysis without
the proportional hazard assumptions, the survival trees.

\hypertarget{survival-trees}{%
\section{Survival Trees}\label{survival-trees}}

Survival trees are methods based in tree based models based on Random Forest (\protect\hyperlink{ref-Breiman_2001}{Breiman 2001}).
Random survival forests is an ensemble method for analysis of right-censored
data (\protect\hyperlink{ref-Ishwaran_Kogalur_Blackstone_Lauer_2008}{Ishwaran et al. 2008}), using randomization to
improve the performance.
Random survival forests follows this framework (\protect\hyperlink{ref-Ishwaran_Kogalur_Blackstone_Lauer_2008}{Ishwaran et al. 2008}):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Draw B random samples of the same size from the original dataset with replacement. The samples that are not drawn are said to be out-of-bag (OOB). Grow a survival tree on each of the b = 1 , . . . , B samples.
\item
  At each node, select a random subset of predictor variables and find the best predictor and splitting value that provide two subsets (the daughter nodes) which maximizes the difference in the objective function.
\item
  Repeat step 2 recursively on each daughter node until a stopping criterion is met.
\item
  Calculate a cumulative hazard function (CHF) for each tree and average over all CHFs for the B trees to obtain the ensemble CHF.
\item
  Compute the prediction error for the ensemble CHF using only the OOB data.
\end{enumerate}

In each node is selected a predictor \(x\) from a random selected predicted variables
and split value \(c\) (one unique value of \(x\)).
Each sample \(i\) if assigned the daughter right node if \(x_i \le c\) or left if
\(x_i \ge c\), then is calculated the logrank such as:

\begin{equation}
L(x, c) = \frac{ \sum^{N}_{i=1} \left(  d_{i, 1} - Y_{i,1} \frac{d_i}{Y_i} \right)  }
               { \sqrt{  \sum^{N}_{i=1}  \frac{Y_{i,1}}{Y_i} \left( 1 - \frac{Y_{i,1}}{Y_i} \right) \left( \frac{Y_i-d_i}{Y_i-1} \right) d_i  } }
\end{equation}

Where:

\begin{itemize}
\tightlist
\item
  \(j\): Daughter node, \(j \in \{1,2\}\)
\item
  \(d_{i,j}\): Number of events at time \(t_i\) in daughter node \(j\)
\item
  \(Y_{i,j}\): Number of elements that had the event or are in risk at time \(t_i\)
  in daughter node \(j\)
\item
  \(d_i\): Number of events at time \(t_i\), such \(d_i=\sum_j{d_{i,j}}\)
\item
  \(Y_i\): Number of elements that experienced an event or are at risk at
  \(t_i\) so \(Y_i=\sum_j Y_{i,j}\)
\end{itemize}

We loop every \(x\) and \(c\) until find \(x^*\) that satisfy \(|L(x^{*}, c^{*})| \geq |L(x, c)|\)
for every \(x\) and \(c\).
The model performance was determined with the concordance probability (C-index),
Brier Score (BS) and Mean Absolute Error (MAE) (\protect\hyperlink{ref-wangmachine2017}{Wang, Li, and Reddy 2017}). The feature
importance was determined calculating the difference between the true class
label and noised data (\protect\hyperlink{ref-Breiman_2001}{Breiman 2001}).

The BS is used to evaluate the predicted accuracy of the survival function
at a given time \(t\).
Representing the average square distance between the survival status and the
predicted survival probability, where the value 0 is the best possible outcome.

\begin{equation}
BS(t) =  \frac{1}{N} \sum_{i = 1}^{N} \left( \frac{\left( 0 - \hat{S}(t, \vec{x}_i)\right)^2 \cdot \mathbb{1}_{T_i \leq t, \delta_i = 1}}{ \hat{G}(T_i^-)} + \frac{ \left( 1 - \hat{S}(t, \vec{x}_i)\right)^2 \cdot \mathbb{1}_{T_i > t}}{ \hat{G}(t)} \right)
\end{equation}

The model should have a Brier score below \(0.25\).
Considering that if \(\forall i \in [\![1, N]\!], \hat{S}(t, \vec{x}_i) = 0.5\), then \(BS(t) = 0.25\).

\hypertarget{methodology}{%
\section{Methodology}\label{methodology}}

To simplify the analysis, the survival probabilities are presented as a survival curve.
The survival curve is a representation of the survival probabilities corresponding to a
time where the events are observed (\protect\hyperlink{ref-Bland_Altman_1998}{Bland and Altman 1998}).
The survival analysis was conducted using the package Lifelines (\protect\hyperlink{ref-Pilon_2021}{Davidson-Pilon 2021}).
Where dropout is a binary value where one represent churn and zero not churn. The
dropout happens when a member does not have a payment.

The random survival forest was developed using the package PySurvival
(\protect\hyperlink{ref-Fotso_others_2019}{Fotso et al. 2019}). PySurvival is an open source python package for
Survival Analysis modeling - the modeling concept used to analyze or predict
when an event is likely to happen.
The model was built with with 70\% of the data for training and 30\% for testing.

The model performance was determined with the concordance probability (C-index),
Brier Score (BS) and Mean Absolute Error (MAE) (\protect\hyperlink{ref-wangmachine2017}{Wang et al. 2017}). The feature
importance was determined calculating the difference between the true class
label and noised data (\protect\hyperlink{ref-Breiman_2001}{Breiman 2001}).

The BS measure the average discrepancies between the status (dropout/non-dropout)
and the estimated probabilities at a given time.
The Integrated Brier Score (IBS) was used to calculate the performance
in all available times (from \(t_1\) to \(t_{max}\)) as:

\begin{equation}
IBS = \int_{t_1}^{t_\text{max}} {BS}^c(t) d w(t)
\end{equation}

Representing the average square distance between the survival status and the
predicted survival probability, where the value 0 is the best possible outcome.

The calculation of he number of clusters used the package mclust (\protect\hyperlink{ref-scrucca2016}{Scrucca et al. 2016})
using the Bayesian Information Criterion (BIC). The model that gives the minimum
BIC score can be selected as the best model (\protect\hyperlink{ref-schwarz1978}{Schwarz 1978}) simplifying the
problem related to choosing the number of components and identifying the
structure of the covariance matrix, based on modelling with multivariate normal
distributions for each component that forms the data set (\protect\hyperlink{ref-akogul2016}{Akogul and Erisoglu 2016}).

The hybrid approach was develop as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Identify the optimal number of clusters using (\protect\hyperlink{ref-scrucca2016}{Scrucca et al. 2016})
\item
  Fit the model using the identified number of clusters
\item
  Estimate for each element the cluster
\item
  for each cluster follow the framework proposed by \protect\hyperlink{ref-Ishwaran_Kogalur_Blackstone_Lauer_2008}{Ishwaran et al.} (\protect\hyperlink{ref-Ishwaran_Kogalur_Blackstone_Lauer_2008}{2008})
  to calculate the random survival model
\end{enumerate}

\hypertarget{dataset}{%
\section{Dataset}\label{dataset}}

In this study, data from 5,209 fitness customers was analysed (mean age = 27.88, SD=11.80 years)
from a Portuguese fitness centre.
The data was collected from software e@sport (Cedis, Portugal) between 2014 and 2017.
The information retrieved was: Age of the participants in years; Sex (0-female, 1-male);
Non-attendance days before dropout; Total amount billed; Average number of visits per week;
Total number of visits; Weekly contracted accesses; Number of registration renewals;
Number of customer referrals; Registration month; Customer enrolment duration; and status
(dropout/non-dropout).
Dropout event occur when customer communicate the intention to terminate the contract or
did not pay the monthly fee during 60 days.

Table \ref{tab:summarytable} shows data's summary statistics. The average age
is 27.9 ±
11.8, the entries are
29 ±
41.2 with an inscription period of
9 ±
8.2 months.

\begin{table}

\caption{\label{tab:summarytable}Summary statistics of features used}
\centering
\begin{tabular}[t]{lc}
\toprule
\textbf{Characteristic} & \textbf{N = 5,209}\\
\midrule
Age in years, Mean (SD) & 28 (12)\\
Male or female, \% & 35\%\\
dayswfreq, Mean (SD) & 76 (102)\\
tbilled, Mean (SD) & 155 (155)\\
maccess, Mean (SD) & 0.89 (0.76)\\
\addlinespace
freeuse, \% & 4.9\%\\
nentries, Mean (SD) & 29 (41)\\
cfreq, \% & \\
\hspace{1em}2 & 1.3\%\\
\hspace{1em}4 & 2.4\%\\
\addlinespace
\hspace{1em}6 & 0.2\%\\
\hspace{1em}7 & 96\%\\
months, Mean (SD) & 9 (8)\\
dropout, \% & 88\%\\
\bottomrule
\end{tabular}
\end{table}

Figure \ref{fig:membershipmonth} shows the distribution of the dropout
considering the number of months of membership.

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/membershipmonth-1.pdf}
\caption{\label{fig:membershipmonth}Number of members by month}
\end{figure}

\hypertarget{results}{%
\section{Results}\label{results}}

The table \ref{tab:survivalprobabilities} depicts the data of the survival time of the customers
during the first months, the results showed that the customers have a survival probability of 24.44\%
at 12 months (column \(p_i\) - likelihood probability) with a median survival time of 10 months
(column estimated\_survival).
The survival probability at 6 months was 54.5\%, representing an risk of dropout of
45.5\% with a estimated survival of 6 months.

\begin{table}

\caption{\label{tab:survivalprobabilities}Determination of the survival time probabilities}
\centering
\begin{threeparttable}
\begin{tabular}[t]{rrrrrrrr}
\toprule
event\_at & removed & observed & censored & entrance & at\_risk & estimated\_survival & prob\\
\midrule
0 & 1 & 1 & 0 & 5209 & 5209 & 7 & 1.000\\
1 & 339 & 249 & 90 & 0 & 5208 & 7 & 0.952\\
2 & 543 & 449 & 94 & 0 & 4869 & 7 & 0.864\\
3 & 520 & 506 & 14 & 0 & 4326 & 7 & 0.763\\
4 & 368 & 362 & 6 & 0 & 3806 & 7 & 0.691\\
\addlinespace
5 & 361 & 350 & 11 & 0 & 3438 & 6 & 0.620\\
6 & 385 & 373 & 12 & 0 & 3077 & 6 & 0.545\\
7 & 254 & 240 & 14 & 0 & 2692 & 5 & 0.496\\
8 & 215 & 192 & 23 & 0 & 2438 & 6 & 0.457\\
9 & 192 & 171 & 21 & 0 & 2223 & 6 & 0.422\\
\addlinespace
10 & 288 & 270 & 18 & 0 & 2031 & 6 & 0.366\\
11 & 362 & 350 & 12 & 0 & 1743 & 9 & 0.293\\
12 & 240 & 229 & 11 & 0 & 1381 & 10 & 0.244\\
13 & 85 & 60 & 25 & 0 & 1141 & 9 & 0.231\\
14 & 112 & 86 & 26 & 0 & 1056 & 9 & 0.212\\
\addlinespace
15 & 78 & 73 & 5 & 0 & 944 & 9 & 0.196\\
16 & 61 & 59 & 2 & 0 & 866 & 10 & 0.183\\
17 & 68 & 64 & 4 & 0 & 805 & 10 & 0.168\\
18 & 61 & 56 & 5 & 0 & 737 & 9 & 0.155\\
19 & 50 & 39 & 11 & 0 & 676 & 9 & 0.146\\
\addlinespace
20 & 53 & 36 & 17 & 0 & 626 & 9 & 0.138\\
21 & 74 & 56 & 18 & 0 & 573 & 10 & 0.124\\
22 & 75 & 61 & 14 & 0 & 499 & 11 & 0.109\\
23 & 49 & 39 & 10 & 0 & 424 & 11 & 0.099\\
24 & 20 & 13 & 7 & 0 & 375 & 10 & 0.096\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item Removed – the sum of customers with dropout and that are censored; Censored – the event did not occur during the period of this data, collection; Risk of Dropout –  number of customers at risk of, dropout; pi – survival probability; Estimated Survival - months to survive in the sports facility.
\end{tablenotes}
\end{threeparttable}
\end{table}

Figure \ref{fig:lifelinesplot} shows the Kaplan Meier survival curve customers
considering the number of months of membership (x axis) and survival probability
(y axis).
The customer dropout is very high in the first 12 months, ranging from a survival
probability of 54\% after the first 6 months until 24\% after 12 months.

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/lifelinesplot-1.pdf}
\caption{\label{fig:lifelinesplot}Survival probabilities}
\end{figure}

Figure \ref{fig:logrank} shows the survival by gender.
The survival curves by gender are very similar, both types of
customers present a behavior that is not very different.

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/logrank-3.pdf}
\caption{\label{fig:logrank}Survival by gender}
\end{figure}

Figure \ref{fig:logrankcfreq} shows the survival by contracted frequency.
Customers with contracted frequency of 6 and 4 times a week have higher survival
probabilities, against lower of customers with contracted frequencies of 7 and
2 times a week.
Survival curves allow to explore tendencies related to survival to extract
actionable knowledge.

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/logrankcfreq-5.pdf}
\caption{\label{fig:logrankcfreq}Survival by contracted frequency}
\end{figure}

The proportional hazard assumptions failed in the following variables:
age p\textless0.01, cfreq p\textless0.01, dayswfreq p\textless0.01, tbilled p\textless0.01, freeuse
p\textless0.01, nentries p\textless0.01.

\hypertarget{survival-trees-1}{%
\subsection{Survival trees}\label{survival-trees-1}}

To evaluate the performance of the random survival forest we have calculated
the concordance probability (C-index), IBS and Mean Absolute Error (MAE).
The IBS presents an accuracy along the 12 months of 0.08 (figure \ref{fig:createModel1}).

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/createModel1-7.pdf}
\caption{\label{fig:createModel1}Model performance}
\end{figure}

The actual versus predicted model presents the actual and predicted customers which dropout
during the 40 months, which as an average absolute error of 7.5 customers (figure \ref{fig:createModel2}).

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/createModel2-9.pdf}
\caption{\label{fig:createModel2}Conditional survival forest}
\end{figure}

Table \ref{tab:summarytable2} shows features importance calculated according (\protect\hyperlink{ref-Breiman_2001}{Breiman 2001}),
where the percent increase in misclassification rate as compared to the out-of-bag rate (with all
variables intact), out-of-bag is a bootstrap aggregating (subsampling with replacement to create
training samples for the model to learn from) where two independent sets are created.
One set, the bootstrap sample, data chosen to be ``in-the-bag'' by sampling with replacement and the
out-of-bag is all data not chosen in the sampling process.
The most important variable is the \(dayswfreq\), followed by \(tbilled\) and
\(nentries\), compared with the \(cfreq\), \(age\), and \(sex\).

\begin{table}

\caption{\label{tab:summarytable2}Features importance in the survival model}
\centering
\begin{tabular}[t]{lrr}
\toprule
feature & importance & pct\_importance\\
\midrule
tbilled & 7.9434176 & 0.2845915\\
dayswfreq & 5.3046172 & 0.1900503\\
nentries & 4.7644466 & 0.1706974\\
maccess & 4.6711845 & 0.1673561\\
freeuse & 3.2174794 & 0.1152737\\
\addlinespace
cfreq & 2.0105041 & 0.0720310\\
age & -0.2880629 & 0.0000000\\
sex\_1 & -1.5439344 & 0.0000000\\
\bottomrule
\end{tabular}
\end{table}

The prediction is very similar to the actual value.
The model accuracy is very high with a root mean square error of
14.
The mean absolute error mean was 7.23 customers,
and the median absolute error was 3.76.

\hypertarget{survival-trees-based-model-with-clusters}{%
\subsection{Survival trees based model with clusters}\label{survival-trees-based-model-with-clusters}}

In this approach we have created clusters and applied the survival trees within
each cluster.
The determination of the clusters using the BIC criterion where the EEV model: 7 clusters -57159.24;
6 clusters -63937.59; and 4 clusters -77088.81 figure \ref{fig:clusters} shows the
determination of the number of clusters using BIC, also the elbow analysis available in
figure \ref{fig:elbowCalculation}.
An optimal number of clusters was considered of five.
Considering that was the value after the average distortion was flattened.

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/clusters-1.pdf}
\caption{\label{fig:clusters}Analysis number of clusters}
\end{figure}

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/elbowCalculation-1.pdf}
\caption{\label{fig:elbowCalculation}Elbow analysis}
\end{figure}

The calculation of the clusters to each member in the dataset was developed considering,
7,6 and 4 clusters.
However after the executions KMeans only performed with 3 clusters.

\includegraphics{customer_fitness_artigo_files/figure-latex/kmeans PCA-3.pdf}

The model accuracy is very high in the first years. The prediction is very
similar to the actual value. The absolute error mean of
6 customers.

\includegraphics{customer_fitness_artigo_files/figure-latex/createModelClusters-5.pdf} \includegraphics{customer_fitness_artigo_files/figure-latex/createModelClusters-6.pdf} \includegraphics{customer_fitness_artigo_files/figure-latex/createModelClusters-7.pdf}

The performance of the cluster 1 the IBS presents an accuracy of 0.06
(figure \ref{fig:createModelCluster1IBS}) along all time.
The actual versus predicted model presents the actual and predicted customers which dropout during
the 40 months, which as an mean absolute error of 1.5 customers, the mean median absolute
error was 0.61 and the Root Mean Square Error of 2.8 (figure \ref{fig:createModelCluster1SF}).

The features importance in the survival model cluster 1 (table \ref{tab:summarytableCluster1})
identify the three most relevant features to predict survival \(maccess\),
\(tbilled\), and \(dayswfreq\). The features with lower relevance were \(freeuse\), \(sex\) and \(cfreq\).

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/createModelCluster1IBS-11.pdf}
\caption{\label{fig:createModelCluster1IBS}Model performance cluster 1}
\end{figure}

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/createModelCluster1SF-13.pdf}
\caption{\label{fig:createModelCluster1SF}Conditional survival forest cluster 1}
\end{figure}

\begin{table}

\caption{\label{tab:summarytableCluster1}Features importance in the survival model with cluster 1}
\centering
\begin{tabular}[t]{lrr}
\toprule
feature & importance & pct\_importance\\
\midrule
tbilled & 6.7827606 & 0.2644082\\
freeuse & 5.7521637 & 0.2242331\\
maccess & 4.4870321 & 0.1749152\\
dayswfreq & 3.9224018 & 0.1529046\\
nentries & 3.0360807 & 0.1183537\\
\addlinespace
cfreq & 0.8262370 & 0.0322087\\
age & 0.4607052 & 0.0179594\\
sex\_1 & 0.3852319 & 0.0150173\\
\bottomrule
\end{tabular}
\end{table}

The performance of the cluster 2 the IBS presents an accuracy along time 0.09
(figure \ref{fig:createModelCluster2IBS}) along all time.
The actual versus predicted model presents the actual and predicted customers which dropout during
the 40 months, which as an mean absolute error of 6.5 customers, the mean median absolute
error was 2.3 and the Root Mean Square Error of 11.79 (figure \ref{fig:createModelCluster2SF}).
The features importance in the survival model cluster 2 (table \ref{tab:summarytableCluster2})
identify the three most relevant features to predict survival \(dayswfreq\),
\(tbilled\), and \(freeuse\). The least relevant were \(nentries\), \(cfreq\), and \(age\).

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/createModelCluster2IBS-1.pdf}
\caption{\label{fig:createModelCluster2IBS}Model performance cluster 2}
\end{figure}

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/createModelCluster2SF-3.pdf}
\caption{\label{fig:createModelCluster2SF}Conditional survival forest cluster 2}
\end{figure}

The performance of the cluster 3 the IBS presents an accuracy along time 0.01
(figure \ref{fig:createModelCluster3IBS}) along all time.
The actual versus predicted model presents the actual and predicted customers which dropout during
the 40 months, which as an mean absolute error of 1.5 customers, the mean median absolute
error was 1.2 and the Root Mean Square Error of 2.08 (figure \ref{fig:createModelCluster3SF}).
The features importance in the survival model cluster 3 (table \ref{tab:summarytableCluster3})
identify the three most relevant features to predict survival \(tbilled\),
\(dayswfreq\), and \(nentries\).
The least relevant were \(sex\), \(age\), and \(cfreq\).

\begin{table}

\caption{\label{tab:summarytableCluster2}Features importance in the survival model with cluster 2}
\centering
\begin{tabular}[t]{lrr}
\toprule
feature & importance & pct\_importance\\
\midrule
tbilled & 6.6524419 & 0.2970936\\
maccess & 4.0288064 & 0.1799238\\
dayswfreq & 3.9304155 & 0.1755297\\
nentries & 2.7923452 & 0.1247043\\
freeuse & 2.4090838 & 0.1075881\\
\addlinespace
age & 1.4997470 & 0.0669777\\
sex\_1 & 1.0789000 & 0.0481829\\
cfreq & -0.6615769 & 0.0000000\\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/createModelCluster3IBS-1.pdf}
\caption{\label{fig:createModelCluster3IBS}Model performance cluster 3}
\end{figure}

\begin{figure}
\centering
\includegraphics{customer_fitness_artigo_files/figure-latex/createModelCluster3SF-3.pdf}
\caption{\label{fig:createModelCluster3SF}Conditional survival forest cluster 3}
\end{figure}

The features importance in the survival model (table \ref{tab:summarytableCluster3})
identify the three most relevant features to predict survival \(dayswfreq\),
\(tbilled\), and \(nentries\).

\begin{table}

\caption{\label{tab:summarytableCluster3}Features importance in the survival model with cluster 3}
\centering
\begin{tabular}[t]{lrr}
\toprule
feature & importance & pct\_importance\\
\midrule
tbilled & 3.7920500 & 0.3033014\\
nentries & 2.8590522 & 0.2286769\\
dayswfreq & 2.3304981 & 0.1864014\\
maccess & 1.8131070 & 0.1450186\\
age & 1.5533976 & 0.1242461\\
\addlinespace
sex\_1 & 0.1544767 & 0.0123556\\
freeuse & 0.0000000 & 0.0000000\\
cfreq & 0.0000000 & 0.0000000\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{model-comparison}{%
\subsubsection{Model Comparison}\label{model-comparison}}

Table \ref{tab:summarytable3} shows the performance of both approaches, with or without clusters.
The RMSE in the clusters 1, 2 and 3 is lower than not using clusters to predict dropout.
Overall the performance improved. The performance is also better using mean and median.

\begin{table}

\caption{\label{tab:summarytable3}Performance of prediction in each cluster}
\centering
\begin{tabular}[t]{llll}
\toprule
cluster & rmse & mean & median\\
\midrule
0 & 10.9534010098597 & 3.24949079763695 & 6.27043826620657\\
2 & 2.86277805553651 & 0.592331070836667 & 1.69865648584295\\
1 & 1.92696670365884 & 1.39451757989365 & 1.42197114146461\\
w/cluster & 13.8883596176347 & 3.75609199994054 & 7.22583527582812\\
\bottomrule
\end{tabular}
\end{table}

The model accuracy without clusters is very high with a root mean
square error of 13, the mean absolute error mean was 7.53 customers, and the median
absolute error was 4.04.
The model using clusters had an mean absolute error of 1.5 customers, the mean median absolute
error was 1.2 and the Root Mean Square Error of 2.08.
The performance using clusters improved significantly.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

This paper investigated the customer dropout in a Health Club organization, using
a dynamic perspective that the dropout risk varies along the time.
Exploring two approaches, using a survival model based on random forests
with or without clusters.
The model using clusters allowed to combine the customers in different clusters,
an hybrid approach.
Based on this performance the proposed model using clusters allows to improve
the accuracy on the survival model allowing to target approaches considering
the timing when the dropout occurs, considering the clusters where the customer is.
Is very important for managers use this information to improve their retention
strategies.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-akogul2016}{}}%
Akogul, Serkan and Murat Erisoglu. 2016. {``A Comparison of Information Criteria in Clustering Based on Mixture of Multivariate Normal Distributions.''} \emph{Mathematical and Computational Applications} 21(3):34.

\leavevmode\vadjust pre{\hypertarget{ref-Alboukaey_dynamic_2020}{}}%
Alboukaey, Nadia, Ammar Joukhadar, and Nada Ghneim. 2020b. {``Dynamic Behavior Based Churn Prediction in Mobile Telecom.''} \emph{Expert Systems with Applications} 162:113779.

\leavevmode\vadjust pre{\hypertarget{ref-alboukaey_dynamic_2020}{}}%
Alboukaey, Nadia, Ammar Joukhadar, and Nada Ghneim. 2020a. {``Dynamic Behavior Based Churn Prediction in Mobile Telecom.''} \emph{Expert Systems with Applications} 162:113779.

\leavevmode\vadjust pre{\hypertarget{ref-amin_customer_2017}{}}%
Amin, Adnan, Sajid Anwar, Awais Adnan, Muhammad Nawaz, Khalid Alawfi, Amir Hussain, and Kaizhu Huang. 2017. {``Customer Churn Prediction in the Telecommunication Sector Using a Rough Set Approach.''} \emph{Neurocomputing} 237:242--54.

\leavevmode\vadjust pre{\hypertarget{ref-ascarza_retention_2018}{}}%
Ascarza, Eva. 2018. {``Retention Futility: Targeting High-Risk Customers Might Be Ineffective.''} \emph{Journal of Marketing Research} 55(1):80--98.

\leavevmode\vadjust pre{\hypertarget{ref-Ascarza_Hardie_2013}{}}%
Ascarza, Eva and Bruce G. S. Hardie. 2013. {``A Joint Model of Usage and Churn in Contractual Settings.''} \emph{Marketing Science} 32(4):570--90.

\leavevmode\vadjust pre{\hypertarget{ref-Athanassopoulos_2000}{}}%
Athanassopoulos, Antreas D. 2000. {``Customer Satisfaction Cues to Support Market Segmentation and Explain Switching Behavior.''} \emph{Journal of Business Research} 47(3):191--207.

\leavevmode\vadjust pre{\hypertarget{ref-Bland_Altman_1998}{}}%
Bland, J. M. and D. G. Altman. 1998. {``Survival Probabilities (the Kaplan-Meier Method).''} \emph{BMJ (Clinical Research Ed.)} 317(7172):1572.

\leavevmode\vadjust pre{\hypertarget{ref-Breiman_2001}{}}%
Breiman, Leo. 2001. {``Random Forests.''} \emph{Machine Learning} 45(1):5--32.

\leavevmode\vadjust pre{\hypertarget{ref-burez_separating_2008}{}}%
Burez, J. and D. Vandenpoel. 2008. {``Separating Financial from Commercial Customer Churn: {A} Modeling Step Towards Resolving the Conflict Between the Sales and Credit Department.''} \emph{Expert Systems with Applications} 35(1-2):497--514.

\leavevmode\vadjust pre{\hypertarget{ref-Copeland_1923}{}}%
Copeland, Melvin T. 1923. {``Relation of Consumers' Buying Habits to Marketing Methods.''} \emph{Harvard Business Review} 1(3):282--89.

\leavevmode\vadjust pre{\hypertarget{ref-coussement_improving_2009}{}}%
Coussement, Kristof and Dirk Van den Poel. 2009. {``Improving Customer Attrition Prediction by Integrating Emotions from Client/Company Interaction Emails and Evaluating Multiple Classifiers.''} \emph{Expert Systems with Applications} 36(3, Part 2):6127--34.

\leavevmode\vadjust pre{\hypertarget{ref-Pilon_2021}{}}%
Davidson-Pilon, Cameron. 2021. \emph{CamDavidsonPilon/Lifelines}.

\leavevmode\vadjust pre{\hypertarget{ref-devriendt_why_2019}{}}%
Devriendt, Floris, Jeroen Berrevoets, and Wouter Verbeke. 2019. {``Why You Should Stop Predicting Customer Churn and Start Using Uplift Models.''} \emph{Information Sciences}.

\leavevmode\vadjust pre{\hypertarget{ref-Edward_Sahadev_2011}{}}%
Edward, Manoj and Sunil Sahadev. 2011. {``Role of Switching Costs in the Service Quality, Perceived Value, Customer Satisfaction and Customer Retention Linkage.''} \emph{Asia Pacific Journal of Marketing and Logistics} 23(3):327--45.

\leavevmode\vadjust pre{\hypertarget{ref-Ehrlinger_2016}{}}%
Ehrlinger, John. 2016. {``ggRandomForests: Exploring Random Forest Survival.''} \emph{arXiv:1612.08974 {[}Stat{]}}.

\leavevmode\vadjust pre{\hypertarget{ref-Fornell_Wernerfelt_1987}{}}%
Fornell, Claes and Birger Wernerfelt. 1987. {``Defensive Marketing Strategy by Customer Complaint Management: A Theoretical Analysis.''} \emph{Journal of Marketing Research} 24(4):337--46.

\leavevmode\vadjust pre{\hypertarget{ref-Fotso_others_2019}{}}%
Fotso, Stephane and others. 2019. \emph{PySurvival: Open Source Package for Survival Analysis Modeling}.

\leavevmode\vadjust pre{\hypertarget{ref-garcia_intelligent_2017}{}}%
García, David L., Angela Nebot, and Alfredo Vellido. 2017. {``Intelligent Data Analysis Approaches to Churn as a Business Problem: A Survey.''} \emph{Knowledge and Information Systems} 51(3):719--74.

\leavevmode\vadjust pre{\hypertarget{ref-gok_case_2015}{}}%
Gök, Mehmet, Tansel Özyer, and Jamal Jida. 2015. {``A {Case} {Study} for the {Churn} {Prediction} in {Turksat} {Internet} {Service} {Subscription}.''} Pp. 1220--24 in \emph{Proceedings of the 2015 {IEEE}/{ACM} {International} {Conference} on {Advances} in {Social} {Networks} {Analysis} and {Mining} 2015 - {ASONAM} '15}. Paris, France: ACM Press.

\leavevmode\vadjust pre{\hypertarget{ref-Gupta2006Modeling}{}}%
Gupta, Sunil, Dominique Hanssens, Bruce Hardie, Wiliam Kahn, V. Kumar, Nathaniel Lin, Nalini Ravishanker, and S. Sriram. 2006. {``Modeling Customer Lifetime Value.''} \emph{Journal of Service Research} 9(2):139--55.

\leavevmode\vadjust pre{\hypertarget{ref-hung_applying_2006}{}}%
Hung, Shin-Yuan, David C. Yen, and Hsiu-Yu Wang. 2006. {``Applying Data Mining to Telecom Churn Management.''} \emph{Expert Systems with Applications} 31(3):515--24.

\leavevmode\vadjust pre{\hypertarget{ref-Ishwaran_Kogalur_Blackstone_Lauer_2008}{}}%
Ishwaran, Hemant, Udaya B. Kogalur, Eugene H. Blackstone, and Michael S. Lauer. 2008. {``Random Survival Forests.''} \emph{The Annals of Applied Statistics} 2(3):841--60.

\leavevmode\vadjust pre{\hypertarget{ref-JafariMarandi_Denton_Idris_Smith_Keramati_2020}{}}%
Jafari-Marandi, Ruholla, Joshua Denton, Adnan Idris, Brian K. Smith, and Abbas Keramati. 2020. {``Optimum Profit-Driven Churn Decision Making: Innovative Artificial Neural Networks in Telecom Industry.''} \emph{Neural Computing and Applications} 32(18):14929--62.

\leavevmode\vadjust pre{\hypertarget{ref-Mellens_Dekimpe_Steenkamp_1996}{}}%
Mellens, Martin, Marnik Dekimpe, and JBEM Steenkamp. 1996. {``A Review of Brand-Loyalty Measures in Marketing.''} \emph{Review of Business and Economic Literature} 41(4):507--33.

\leavevmode\vadjust pre{\hypertarget{ref-perianez_churn_2016}{}}%
Perianez, Africa, Alain Saas, Anna Guitart, and Colin Magne. 2016. {``Churn {Prediction} in {Mobile} {Social} {Games}: {Towards} a {Complete} {Assessment} {Using} {Survival} {Ensembles}.''} Pp. 564--73 in \emph{2016 {IEEE} {International} {Conference} on {Data} {Science} and {Advanced} {Analytics} ({DSAA})}. Montreal, QC, Canada: IEEE.

\leavevmode\vadjust pre{\hypertarget{ref-Reichheld_1996}{}}%
Reichheld, Frederick F. 1996. {``Learning from Customer Defections.''} \emph{Harvard Business Review} 74(2):56--67.

\leavevmode\vadjust pre{\hypertarget{ref-risselada_staying_2010}{}}%
Risselada, Hans, Peter C. Verhoef, and Tammo H. A. Bijmolt. 2010. {``Staying {Power} of {Churn} {Prediction} {Models}.''} \emph{Journal of Interactive Marketing} 24(3):198--208.

\leavevmode\vadjust pre{\hypertarget{ref-routh_estimating_2020}{}}%
Routh, Pallav, Arkajyoti Roy, and Jeff Meyer. 2020. {``Estimating Customer Churn Under Competing Risks.''} \emph{Journal of the Operational Research Society} 1--18.

\leavevmode\vadjust pre{\hypertarget{ref-Saunders_1980}{}}%
Saunders, J. a. 1980. {``Cluster Analysis for Market Segmentation.''} \emph{European Journal of Marketing} 14(7):422--35.

\leavevmode\vadjust pre{\hypertarget{ref-Schober2018Survival}{}}%
Schober, Patrick and Thomas R. Vetter. 2018. {``Survival Analysis and Interpretation of Time-to-Event Data: The Tortoise and the Hare.''} \emph{Anesthesia and Analgesia} 127(3):792--98.

\leavevmode\vadjust pre{\hypertarget{ref-schwarz1978}{}}%
Schwarz, Gideon. 1978. {``Estimating the Dimension of a Model.''} \emph{The Annals of Statistics} 6(2):461--64.

\leavevmode\vadjust pre{\hypertarget{ref-scrucca2016}{}}%
Scrucca, Luca, Michael Fop, T. ,Brendan Murphy, and Adrian,E. Raftery. 2016. {``Mclust 5: Clustering, Classification and Density Estimation Using Gaussian Finite Mixture Models.''} \emph{The R Journal} 8(1):289.

\leavevmode\vadjust pre{\hypertarget{ref-Singer_Willett_1993}{}}%
Singer, Judith D. and John B. Willett. 1993. {``It's about Time: Using Discrete-Time Survival Analysis to Study Duration and the Timing of Events.''} \emph{Journal of Educational Statistics} 18(2):155--95.

\leavevmode\vadjust pre{\hypertarget{ref-Stensrud_Hernuxe1n_2020}{}}%
Stensrud, Mats J. and Miguel A. Hernán. 2020. {``Why Test for Proportional Hazards?''} \emph{JAMA} 323(14):1401--2.

\leavevmode\vadjust pre{\hypertarget{ref-vijaya_sivasankar_2019}{}}%
Vijaya, J. and E. Sivasankar. 2019. {``An Efficient System for Customer Churn Prediction Through Particle Swarm Optimization Based Feature Selection Model with Simulated Annealing.''} \emph{Cluster Computing} 22(S5):10757--68.

\leavevmode\vadjust pre{\hypertarget{ref-wangmachine2017}{}}%
Wang, Ping, Yan Li, and Chandan K. Reddy. 2017. {``Machine Learning for Survival Analysis: A Survey.''} \emph{arXiv:1708.04649 {[}Cs, Stat{]}}.

\end{CSLReferences}

\hypertarget{appendix-chunk-options}{%
\section*{Appendix: Chunk options}\label{appendix-chunk-options}}
\addcontentsline{toc}{section}{Appendix: Chunk options}

\hypertarget{software-versioning-r}{%
\subsection{Software versioning R}\label{software-versioning-r}}

\hypertarget{r}{%
\subsubsection{R}\label{r}}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{cat}\NormalTok{(}\FunctionTok{paste}\NormalTok{(}\StringTok{"\#"}\NormalTok{, }\FunctionTok{capture.output}\NormalTok{(}\FunctionTok{sessionInfo}\NormalTok{()), }\StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{, }\AttributeTok{collapse  =} \StringTok{""}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # R version 4.1.3 (2022-03-10) 
## # Platform: x86_64-w64-mingw32/x64 (64-bit) 
## # Running under: Windows 10 x64 (build 22622) 
## #  
## # Matrix products: default 
## #  
## # locale: 
## # [1] LC_COLLATE=Portuguese_Portugal.1252  LC_CTYPE=Portuguese_Portugal.1252    
## # [3] LC_MONETARY=Portuguese_Portugal.1252 LC_NUMERIC=C                         
## # [5] LC_TIME=Portuguese_Portugal.1252     
## #  
## # attached base packages: 
## # [1] stats     graphics  grDevices utils     datasets  methods   base      
## #  
## # other attached packages: 
## #  [1] mclust_5.4.10    labelled_2.9.1   kableExtra_1.3.4 gtsummary_1.6.0  
## #  [5] visdat_0.5.3     readxl_1.4.0     stargazer_5.2.3  reticulate_1.25  
## #  [9] ggplot2_3.3.6    dlookr_0.5.6     dplyr_1.0.9      
## #  
## # loaded via a namespace (and not attached): 
## #  [1] reactable_0.2.3     webshot_0.5.3       httr_1.4.3          
## #  [4] tools_4.1.3         utf8_1.2.2          R6_2.5.1            
## #  [7] rpart_4.1.16        colorspace_2.0-3    withr_2.5.0         
## # [10] tidyselect_1.1.2    gridExtra_2.3       curl_4.3.2          
## # [13] compiler_4.1.3      extrafontdb_1.0     cli_3.3.0           
## # [16] rvest_1.0.2         gt_0.6.0            xml2_1.3.3          
## # [19] labeling_0.4.2      bookdown_0.27       scales_1.2.0        
## # [22] mvtnorm_1.1-3       rappdirs_0.3.3      systemfonts_1.0.4   
## # [25] stringr_1.4.0       digest_0.6.29       rmarkdown_2.14      
## # [28] svglite_2.1.0       pkgconfig_2.0.3     htmltools_0.5.2     
## # [31] showtext_0.9-5      extrafont_0.18      fastmap_1.1.0       
## # [34] highr_0.9           htmlwidgets_1.5.4   rlang_1.0.2         
## # [37] rstudioapi_0.13     sysfonts_0.8.8      shiny_1.7.1         
## # [40] generics_0.1.2      farver_2.1.0        jsonlite_1.8.0      
## # [43] magrittr_2.0.3      Formula_1.2-4       Matrix_1.4-1        
## # [46] Rcpp_1.0.8.3        munsell_0.5.0       fansi_1.0.3         
## # [49] gdtools_0.2.4       partykit_1.2-15     lifecycle_1.0.1     
## # [52] stringi_1.7.6       yaml_2.3.5          inum_1.0-4          
## # [55] grid_4.1.3          hrbrthemes_0.8.0    promises_1.2.0.1    
## # [58] forcats_0.5.1       crayon_1.5.1        lattice_0.20-45     
## # [61] haven_2.5.0         splines_4.1.3       hms_1.1.1           
## # [64] knitr_1.39          pillar_1.7.0        glue_1.6.2          
## # [67] evaluate_0.15       pagedown_0.18       broom.helpers_1.7.0 
## # [70] vctrs_0.4.1         png_0.1-7           httpuv_1.6.5        
## # [73] Rttf2pt1_1.3.10     cellranger_1.1.0    gtable_0.3.0        
## # [76] purrr_0.3.4         tidyr_1.2.0         xfun_0.31           
## # [79] mime_0.12           libcoin_1.0-9       xtable_1.8-4        
## # [82] later_1.3.0         survival_3.3-1      viridisLite_0.4.0   
## # [85] tibble_3.1.7        showtextdb_3.0      ellipsis_0.3.2
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
  \CommentTok{\# or use message() instead of cat()}
\end{Highlighting}
\end{Shaded}

\end{document}
